---
title: "Spatial Weights and Applications"
subtitle: "HANDS-ON EXERCISE 4"
author:
  - name: Jedidiah Asaf Tallulembang
    affiliation: "ISSS626 — Geospatial Analytics and Applications"
date: last-modified
format:
  html:
    code-fold: false
    toc: true
execute:
  warning: false
  message: false
---

::: {.pill-row}
[R]{.pill} [spdep]{.pill} [Contiguity]{.pill} [Spatial Lag]{.pill}
:::

This exercise follows [Spatial Weights and
Applications](https://r4gdsa.netlify.app/chap08), the first reading in the Spatial
Weights block of the course.

## 1. Overview

Everything in Exercises 2 and 3 treated space as distance between *points*. This
exercise switches to **areal** data — polygons with values attached — where the
question is no longer "how far apart are these events" but "which regions are
neighbours, and how much should each neighbour count?"

That is what a **spatial weights matrix** encodes. It is the object that every
subsequent technique in the course depends on: spatial autocorrelation, clustering,
and geographically weighted regression all take a weights matrix as input. Getting it
wrong quietly invalidates everything downstream, which is why a whole exercise is
spent on it.

By the end I should be able to compute contiguity weights, distance-based weights,
inverse-distance weights, row-standardise them, and use them to derive spatially lagged
variables.

## 2. The Study Area and Data

| Dataset | Description | Format |
|---|---|---|
| `Hunan` | County boundary layer for Hunan province, China — 88 counties | ESRI Shapefile (polygon) |
| `Hunan_2012.csv` | Selected local development indicators for 2012 | CSV |

A change of scene from Singapore. Hunan gives 88 areal units with a real spread of
development levels, which is exactly what is needed to see spatial weighting do
something visible — Singapore's 332 subzones are too small and too uniformly urban to
make a good contiguity example.

## 3. Getting Started

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

## 4. Getting the Data Into R

### 4.1 Importing the Shapefile

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan",
                 quiet = TRUE)
```

```{r}
hunan
```

### 4.2 Importing the CSV

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv", show_col_types = FALSE)
```

```{r}
glimpse(hunan2012)
```

### 4.3 Performing the Relational Join

The join key is `County`, which appears in both tables. I name it explicitly rather
than letting `dplyr` guess, so that the code fails loudly if the column ever changes
rather than silently joining on something else.

```{r}
hunan <- left_join(hunan, hunan2012, by = "County") %>%
  select(1:4, 7, 15)

names(hunan)
```

```{r}
nrow(hunan)
```

88 counties in, 88 out — no rows duplicated or dropped by the join, which is the first
thing worth checking after any `left_join` onto a spatial layer.

## 5. Visualising the Regional Development Indicator

```{r}
#| fig-width: 10
basemap <- tm_shape(hunan) +
  tm_polygons(fill = "grey95", col = "white", lwd = 0.5) +
  tm_text("NAME_3", size = 0.35)

gdppc <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(style = "quantile", values = "viridis"),
              fill.legend = tm_legend(title = "GDP per capita"),
              col = "white", lwd = 0.4)

tmap_arrange(basemap, gdppc, asp = 1, ncol = 2)
```

::: {.callout-note}
## What the choropleth already suggests
High GDP per capita is not scattered at random. The brighter counties sit together,
clustered around Changsha in the north-east, and the lower quantiles form a contiguous
band through the west. That visual impression of "like next to like" is precisely what
spatial autocorrelation statistics formalise — but none of them can be computed until I
have defined what "next to" means, which is this exercise.
:::

## 6. Computing Contiguity Spatial Weights

Contiguity weights define neighbours by shared boundaries. The two standard rules are
named after chess moves:

- **Queen**: regions are neighbours if they share *any* boundary point, edge or corner.
- **Rook**: regions are neighbours only if they share an *edge*.

### 6.1 Queen Contiguity

```{r}
wm_q <- poly2nb(hunan, queen = TRUE)
summary(wm_q)
```

The most connected county:

```{r}
hunan$County[which.max(card(wm_q))]
```

Its neighbours:

```{r}
hunan$NAME_3[wm_q[[which.max(card(wm_q))]]]
```

And their GDP per capita:

```{r}
hunan$GDPPC[wm_q[[which.max(card(wm_q))]]]
```

The full weight matrix, printed as the first few rows of neighbour lists:

```{r}
str(wm_q, list.len = 5)
```

### 6.2 Rook Contiguity

```{r}
wm_r <- poly2nb(hunan, queen = FALSE)
summary(wm_r)
```

```{r}
data.frame(
  Rule = c("Queen", "Rook"),
  `Total links` = c(sum(card(wm_q)), sum(card(wm_r))),
  `Average links` = round(c(sum(card(wm_q)), sum(card(wm_r))) / nrow(hunan), 3),
  check.names = FALSE
)
```

::: {.callout-note}
## Queen vs rook, and why the difference is small here
Queen finds a handful more links than rook. Those extra links are the corner-only
contacts — counties meeting at a single point. On a regular grid the difference is
large and systematic; on irregular administrative boundaries like these, exact
corner-touching is rare, so the two rules very nearly agree.

That does **not** make the choice unimportant. It makes it cheap to defend: with so few
cases in dispute, I can state that the substantive results are insensitive to the
choice, which is a stronger position than having to argue for one rule on theory alone.
:::

### 6.3 Visualising Contiguity Weights

The neighbour lists have to be drawn from county centroids, so those come first.
`st_centroid` on the geometry column returns points; `st_coordinates` turns them into
the plain longitude/latitude matrix that `spdep` expects.

```{r}
coords <- st_coordinates(st_centroid(st_geometry(hunan)))
head(coords, 3)
```

```{r}
#| fig-width: 10
#| fig-height: 5
par(mfrow = c(1, 2))
plot(st_geometry(hunan), border = "grey70", main = "Queen contiguity")
plot(wm_q, coords, pch = 19, cex = 0.5, add = TRUE, col = "#c0392b")
plot(st_geometry(hunan), border = "grey70", main = "Rook contiguity")
plot(wm_r, coords, pch = 19, cex = 0.5, add = TRUE, col = "#2980b9")
par(mfrow = c(1, 1))
```

## 7. Computing Distance-based Neighbours

Contiguity ignores distance entirely: a shared border counts the same whether the two
county centres are 20 km or 200 km apart. Distance-based weights take the opposite view.

### 7.1 Determining the Cut-off Distance

The cut-off must be large enough that no county is left with zero neighbours. The
standard approach is to find the distance to each county's *first* nearest neighbour and
take the maximum of those.

```{r}
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

::: {.callout-important}
## The mistake that cost me this section
My first attempt passed the output of `st_centroid()` straight into `knearneigh()` —
an `sfc` object rather than a coordinate matrix — and got a maximum first-neighbour
distance of **0.6** instead of 61.8. The `longlat = TRUE` argument was silently
overridden, so `spdep` measured in *degrees of latitude and longitude* as though they
were planar units. A 62-unit cut-off then made every county a neighbour of every other
county, and the fixed-distance matrix came out with 87 links per county instead of
about 4.

What makes this dangerous is that it does not error. It produces a perfectly
well-formed weights matrix that is completely wrong, and every statistic computed from
it afterwards would have been wrong too, with nothing to indicate a problem. The
lesson: pass `st_coordinates()` output, and sanity-check that the distances come back in
plausible kilometres before using them.
:::

The maximum first-nearest-neighbour distance is about 62 km, so every county is
guaranteed at least one neighbour at that cut-off.

### 7.2 Fixed Distance Weight Matrix

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

```{r}
str(wm_d62, list.len = 5)
```

```{r}
#| fig-width: 10
#| fig-height: 5
par(mfrow = c(1, 2))
plot(st_geometry(hunan), border = "grey70", main = "1st nearest neighbour")
plot(k1, coords, add = TRUE, col = "#c0392b", length = 0.08)
plot(st_geometry(hunan), border = "grey70", main = "Fixed distance, 62 km")
plot(wm_d62, coords, add = TRUE, pch = 19, cex = 0.5)
par(mfrow = c(1, 1))
```

### 7.3 Adaptive Distance Weight Matrix

A fixed distance band gives densely settled areas many neighbours and sparse areas few.
Fixing the *number* of neighbours instead makes the bandwidth adapt to local density.

```{r}
knn6 <- knn2nb(knearneigh(coords, k = 6))
knn6
```

```{r}
str(knn6, list.len = 5)
```

```{r}
#| fig-height: 5
plot(st_geometry(hunan), border = "grey70", main = "Adaptive: 6 nearest neighbours")
plot(knn6, coords, pch = 19, cex = 0.5, add = TRUE, col = "#27ae60")
```

::: {.callout-note}
## Fixed vs adaptive, stated as a trade
Every county now has exactly six neighbours. That guarantees no isolates and a balanced
weights matrix, but it forces a relationship where none may exist: a remote western
county gets six neighbours even if the nearest is far away and arguably irrelevant.

The fixed-distance matrix is the honest one about geography and the dishonest one about
balance; the adaptive matrix is the reverse. Neither is correct in the abstract. What
matters is stating which was used and why, because the choice propagates into every
subsequent result.
:::

## 8. Weights Based on Inverse Distance

Contiguity says neighbours count equally. Inverse distance weighting (IDW) says nearer
neighbours count more, with weight proportional to 1/distance.

```{r}
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids  <- lapply(dist, function(x) 1 / x)
head(ids, 2)
```

## 9. Row-standardised Weights Matrix

Raw binary weights mean a county with eight neighbours carries eight times the total
influence of a county with one. Row standardisation divides each row by its number of
neighbours so every county's weights sum to 1.

```{r}
rswm_q <- nb2listw(wm_q, style = "W", zero.policy = TRUE)
rswm_q
```

The weights for the first county:

```{r}
rswm_q$weights[1]
```

Five neighbours, each weighted 0.2. The equivalent using the inverse-distance list:

```{r}
rswm_ids <- nb2listw(wm_q, glist = ids, style = "B", zero.policy = TRUE)
rswm_ids$weights[1]
```

```{r}
summary(unlist(rswm_ids$weights))
```

## 10. Application: Spatially Lagged Variables

A spatial lag is the weighted average (or sum) of a variable across a region's
neighbours. It is the basic building block of every spatial autocorrelation statistic.

### 10.1 Spatial Lag with Row-standardised Weights

```{r}
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
head(GDPPC.lag, 5)
```

```{r}
lag.list <- list(hunan$NAME_3, lag.listw(rswm_q, hunan$GDPPC))
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan, lag.res, by = "NAME_3")

kable(head(hunan %>% st_drop_geometry() %>% select(NAME_3, GDPPC, `lag GDPPC`), 6))
```

```{r}
#| fig-width: 10
gdppc_map <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(style = "quantile", values = "viridis"),
              col = "white", lwd = 0.3) +
  tm_title("GDPPC")

lag_map <- tm_shape(hunan) +
  tm_polygons(fill = "lag GDPPC",
              fill.scale = tm_scale_intervals(style = "quantile", values = "viridis"),
              col = "white", lwd = 0.3) +
  tm_title("Spatial lag of GDPPC")

tmap_arrange(gdppc_map, lag_map, asp = 1, ncol = 2)
```

### 10.2 Spatial Lag as a Sum of Neighbouring Values

Using binary weights instead, the lag becomes a sum rather than an average.

```{r}
b_weights  <- lapply(wm_q, function(x) 0 * x + 1)
b_weights2 <- nb2listw(wm_q, glist = b_weights, style = "B")

lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
hunan <- left_join(hunan, lag.res, by = "NAME_3")

kable(head(hunan %>% st_drop_geometry() %>% select(NAME_3, GDPPC, `lag_sum GDPPC`), 6))
```

### 10.3 Spatial Window Average

The window average includes the region itself alongside its neighbours. `include.self()`
adds the diagonal.

```{r}
wm_qs <- include.self(wm_q)
wm_qs
```

```{r}
wm_qs <- nb2listw(wm_qs)
lag_w_avg_gpdpc <- lag.listw(wm_qs, hunan$GDPPC)

lag.list.wm_qs <- list(hunan$NAME_3, lag.listw(wm_qs, hunan$GDPPC))
lag_wm_qs.res <- as.data.frame(lag.list.wm_qs)
colnames(lag_wm_qs.res) <- c("NAME_3", "lag_window_avg GDPPC")
hunan <- left_join(hunan, lag_wm_qs.res, by = "NAME_3")

kable(head(hunan %>% st_drop_geometry() %>%
             select(NAME_3, GDPPC, `lag GDPPC`, `lag_window_avg GDPPC`), 6))
```

### 10.4 Spatial Window Sum

```{r}
wm_qs2 <- include.self(wm_q)
b_weights  <- lapply(wm_qs2, function(x) 0 * x + 1)
b_weights2 <- nb2listw(wm_qs2, glist = b_weights, style = "B")

w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
hunan <- left_join(hunan, w_sum_gdppc.res, by = "NAME_3")

kable(head(hunan %>% st_drop_geometry() %>%
             select(NAME_3, GDPPC, `lag_sum GDPPC`, `w_sum GDPPC`), 6))
```

### 10.5 Comparing the Four Lag Definitions

```{r}
#| fig-height: 8
lag_maps <- function(var, ttl) {
  tm_shape(hunan) +
    tm_polygons(fill = var,
                fill.scale = tm_scale_intervals(style = "quantile", values = "viridis"),
                fill.legend = tm_legend(show = FALSE),
                col = "white", lwd = 0.3) +
    tm_title(ttl)
}

tmap_arrange(lag_maps("lag GDPPC", "Lag (row-std average)"),
             lag_maps("lag_sum GDPPC", "Lag (binary sum)"),
             lag_maps("lag_window_avg GDPPC", "Window average"),
             lag_maps("w_sum GDPPC", "Window sum"),
             ncol = 2, nrow = 2)
```

## 11. What I Take From This Exercise

**The weights matrix is a modelling assumption, not a data property.** Nothing in the
Hunan shapefile tells me whether two counties are neighbours. Queen or rook, 62 km or
6 nearest neighbours, binary or inverse-distance, row-standardised or not — each is a
decision I make, and each produces a different matrix from identical data. Every
autocorrelation statistic and spatial regression later in the course inherits whichever
one I pick. This exercise is not really about `spdep` syntax; it is about the fact that
a large part of a spatial analysis is decided before any statistic is computed.

**The silent-failure lesson.** The `sfc`-versus-matrix mistake in section 7.1 is the one
I will actually remember. It did not throw an error, it did not produce an obviously
absurd map, and it yielded a valid weights object — it was simply measuring degrees and
calling them kilometres. The only thing that caught it was noticing that a
first-nearest-neighbour distance of 0.6 km between Chinese counties is impossible. Unit
plausibility checks are not busywork; here they were the only line of defence.

**Averages versus sums matter more than they look.** The four lag definitions in section
10 produce visibly different maps from the same variable and the same neighbour list.
The binary-sum versions are partly just a map of *how many neighbours each county has*,
which is a property of the geography rather than of GDP. Row standardisation removes
that artefact, which is why it is the sensible default for comparing across regions of
uneven connectivity.

**Connecting forward.** The row-standardised queen contiguity matrix built here is the
direct input to Moran's I and Geary's C in the next exercise. Now that I have seen how
much the matrix can change with defensible choices, the obvious question for that
exercise is how sensitive the autocorrelation result is to the weights — and I intend to
check rather than assume.
