---
title: "Spatial Weights and Applications"
subtitle: "HANDS-ON EXERCISE 4"
author:
  - name: Jedidiah Asaf Tallulembang
    affiliation: "ISSS626 Geospatial Analytics and Applications"
date: last-modified
format:
  html:
    code-fold: false
    toc: true
execute:
  warning: false
  message: false
---

::: {.pill-row}
[R]{.pill} [spdep]{.pill} [Contiguity]{.pill} [Spatial Lag]{.pill}
:::

This exercise follows [Spatial Weights and
Applications](https://r4gdsa.netlify.app/chap08), the first reading in the Spatial
Weights block of the course.

## 1. Overview

Everything in Exercises 2 and 3 treated space as distance between *points*. This
exercise switches to **areal** data, meaning polygons with values attached, where the
question is no longer "how far apart are these events" but "which regions are
neighbours, and how much should each neighbour count?"

That is what a **spatial weights matrix** encodes. It is the object that every
subsequent technique in the course depends on: spatial autocorrelation, clustering,
and geographically weighted regression all take a weights matrix as input. Getting it
wrong quietly invalidates everything downstream, which is why a whole exercise is
spent on it.

By the end I should be able to compute contiguity weights, distance-based weights,
inverse-distance weights, row-standardise them, and use them to derive spatially lagged
variables.

## 2. The Study Area and Data

| Dataset | Description | Format |
|---|---|---|
| `Hunan` | County boundary layer for Hunan province, China, 88 counties | ESRI Shapefile (polygon) |
| `Hunan_2012.csv` | Selected local development indicators for 2012 | CSV |

A change of scene from Singapore. Hunan gives 88 areal units with a real spread of
development levels, which is exactly what is needed to see spatial weighting do
something visible. Singapore's 332 subzones are too small and too uniformly urban to
make a good contiguity example.

## 3. Getting Started

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

## 4. Getting the Data Into R

### 4.1 Importing the Shapefile

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan",
                 quiet = TRUE)
```

```{r}
hunan
```

### 4.2 Importing the CSV

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv", show_col_types = FALSE)
```

```{r}
glimpse(hunan2012)
```

### 4.3 Performing the Relational Join

The join key is `County`, which appears in both tables. I name it explicitly rather
than letting `dplyr` guess, so that the code fails loudly if the column ever changes
rather than silently joining on something else.

```{r}
hunan <- left_join(hunan, hunan2012, by = "County") %>%
  select(1:4, 7, 15)

names(hunan)
```

```{r}
nrow(hunan)
```

88 counties in, 88 out, with no rows duplicated or dropped by the join, which is the first
thing worth checking after any `left_join` onto a spatial layer.

## 5. Visualising the Regional Development Indicator

```{r}
#| fig-width: 10
basemap <- tm_shape(hunan) +
  tm_polygons(fill = "grey95", col = "white", lwd = 0.5) +
  tm_text("NAME_3", size = 0.35)

gdppc <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(style = "quantile", values = "viridis"),
              fill.legend = tm_legend(title = "GDP per capita"),
              col = "white", lwd = 0.4)

tmap_arrange(basemap, gdppc, asp = 1, ncol = 2)
```

::: {.callout-note}
## What the choropleth already suggests
High GDP per capita is not scattered at random. The brighter counties sit together,
clustered around Changsha in the north-east, and the lower quantiles form a contiguous
band through the west. That visual impression of "like next to like" is precisely what
spatial autocorrelation statistics formalise, but none of them can be computed until I
have defined what "next to" means, which is this exercise.
:::

## 6. Computing Contiguity Spatial Weights

Contiguity weights define neighbours by shared boundaries. The two standard rules are
named after chess moves:

- **Queen**: regions are neighbours if they share *any* boundary point, edge or corner.
- **Rook**: regions are neighbours only if they share an *edge*.

### 6.1 Queen Contiguity

```{r}
wm_q <- poly2nb(hunan, queen = TRUE)
summary(wm_q)
```

The most connected county:

```{r}
hunan$County[which.max(card(wm_q))]
```

Its neighbours:

```{r}
hunan$NAME_3[wm_q[[which.max(card(wm_q))]]]
```

And their GDP per capita:

```{r}
hunan$GDPPC[wm_q[[which.max(card(wm_q))]]]
```

The full weight matrix, printed as the first few rows of neighbour lists:

```{r}
str(wm_q, list.len = 5)
```

### 6.2 Rook Contiguity

```{r}
wm_r <- poly2nb(hunan, queen = FALSE)
summary(wm_r)
```

```{r}
data.frame(
  Rule = c("Queen", "Rook"),
  `Total links` = c(sum(card(wm_q)), sum(card(wm_r))),
  `Average links` = round(c(sum(card(wm_q)), sum(card(wm_r))) / nrow(hunan), 3),
  check.names = FALSE
)
```

::: {.callout-note}
## Queen vs rook, and why the difference is small here
Queen finds a handful more links than rook. Those extra links are the corner-only
contacts, where counties meet at a single point. On a regular grid the difference is
large and systematic; on irregular administrative boundaries like these, exact
corner-touching is rare, so the two rules very nearly agree.

That does **not** make the choice unimportant. It makes it cheap to defend: with so few
cases in dispute, I can state that the substantive results are insensitive to the
choice, which is a stronger position than having to argue for one rule on theory alone.
:::

### 6.3 Visualising Contiguity Weights

The neighbour lists have to be drawn from county centroids, so those come first.
`st_centroid` on the geometry column returns points; `st_coordinates` turns them into
the plain longitude/latitude matrix that `spdep` expects.

```{r}
coords <- st_coordinates(st_centroid(st_geometry(hunan)))
head(coords, 3)
```

```{r}
#| fig-width: 10
#| fig-height: 5
par(mfrow = c(1, 2))
plot(st_geometry(hunan), border = "grey70", main = "Queen contiguity")
plot(wm_q, coords, pch = 19, cex = 0.5, add = TRUE, col = "#c0392b")
plot(st_geometry(hunan), border = "grey70", main = "Rook contiguity")
plot(wm_r, coords, pch = 19, cex = 0.5, add = TRUE, col = "#2980b9")
par(mfrow = c(1, 1))
```

## 7. Computing Distance-based Neighbours

Contiguity ignores distance entirely: a shared border counts the same whether the two
county centres are 20 km or 200 km apart. Distance-based weights take the opposite view.

### 7.1 Determining the Cut-off Distance

The cut-off must be large enough that no county is left with zero neighbours. The
standard approach is to find the distance to each county's *first* nearest neighbour and
take the maximum of those.

```{r}
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

::: {.callout-important}
## The mistake that cost me this section
My first attempt passed the output of `st_centroid()` straight into `knearneigh()`,
an `sfc` object rather than a coordinate matrix, and got a maximum first-neighbour
distance of **0.6** instead of 61.8. The `longlat = TRUE` argument was silently
overridden, so `spdep` measured in *degrees of latitude and longitude* as though they
were planar units. A 62-unit cut-off then made every county a neighbour of every other
county, and the fixed-distance matrix came out with 87 links per county instead of
about 4.

What makes this dangerous is that it does not error. It produces a perfectly
well-formed weights matrix that is completely wrong, and every statistic computed from
it afterwards would have been wrong too, with nothing to indicate a problem. The
lesson: pass `st_coordinates()` output, and sanity-check that the distances come back in
plausible kilometres before using them.
:::

The maximum first-nearest-neighbour distance is about 62 km, so every county is
guaranteed at least one neighbour at that cut-off.

### 7.2 Fixed Distance Weight Matrix

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

```{r}
str(wm_d62, list.len = 5)
```

```{r}
#| fig-width: 10
#| fig-height: 5
par(mfrow = c(1, 2))
plot(st_geometry(hunan), border = "grey70", main = "1st nearest neighbour")
plot(k1, coords, add = TRUE, col = "#c0392b", length = 0.08)
plot(st_geometry(hunan), border = "grey70", main = "Fixed distance, 62 km")
plot(wm_d62, coords, add = TRUE, pch = 19, cex = 0.5)
par(mfrow = c(1, 1))
```

### 7.3 Adaptive Distance Weight Matrix

A fixed distance band gives densely settled areas many neighbours and sparse areas few.
Fixing the *number* of neighbours instead makes the bandwidth adapt to local density.

```{r}
knn6 <- knn2nb(knearneigh(coords, k = 6))
knn6
```

```{r}
str(knn6, list.len = 5)
```

```{r}
#| fig-height: 5
plot(st_geometry(hunan), border = "grey70", main = "Adaptive: 6 nearest neighbours")
plot(knn6, coords, pch = 19, cex = 0.5, add = TRUE, col = "#27ae60")
```

::: {.callout-note}
## Fixed vs adaptive, stated as a trade
Every county now has exactly six neighbours. That guarantees no isolates and a balanced
weights matrix, but it forces a relationship where none may exist: a remote western
county gets six neighbours even if the nearest is far away and arguably irrelevant.

The fixed-distance matrix is the honest one about geography and the dishonest one about
balance; the adaptive matrix is the reverse. Neither is correct in the abstract. What
matters is stating which was used and why, because the choice propagates into every
subsequent result.
:::

## 8. Weights Based on Inverse Distance

Contiguity says neighbours count equally. Inverse distance weighting (IDW) says nearer
neighbours count more, with weight proportional to 1/distance.

```{r}
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids  <- lapply(dist, function(x) 1 / x)
head(ids, 2)
```

## 9. Row-standardised Weights Matrix

Raw binary weights mean a county with eight neighbours carries eight times the total
influence of a county with one. Row standardisation divides each row by its number of
neighbours so every county's weights sum to 1.

```{r}
rswm_q <- nb2listw(wm_q, style = "W", zero.policy = TRUE)
rswm_q
```

The weights for the first county:

```{r}
rswm_q$weights[1]
```

Five neighbours, each weighted 0.2. The equivalent using the inverse-distance list:

```{r}
rswm_ids <- nb2listw(wm_q, glist = ids, style = "B", zero.policy = TRUE)
rswm_ids$weights[1]
```

```{r}
summary(unlist(rswm_ids$weights))
```

## 10. Application: Spatially Lagged Variables

A spatial lag is the weighted average (or sum) of a variable across a region's
neighbours. It is the basic building block of every spatial autocorrelation statistic.

### 10.1 Spatial Lag with Row-standardised Weights

```{r}
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
head(GDPPC.lag, 5)
```

```{r}
lag.list <- list(hunan$NAME_3, lag.listw(rswm_q, hunan$GDPPC))
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan, lag.res, by = "NAME_3")

kable(head(hunan %>% st_drop_geometry() %>% select(NAME_3, GDPPC, `lag GDPPC`), 6))
```

```{r}
#| fig-width: 10
gdppc_map <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(style = "quantile", values = "viridis"),
              col = "white", lwd = 0.3) +
  tm_title("GDPPC")

lag_map <- tm_shape(hunan) +
  tm_polygons(fill = "lag GDPPC",
              fill.scale = tm_scale_intervals(style = "quantile", values = "viridis"),
              col = "white", lwd = 0.3) +
  tm_title("Spatial lag of GDPPC")

tmap_arrange(gdppc_map, lag_map, asp = 1, ncol = 2)
```

### 10.2 Spatial Lag as a Sum of Neighbouring Values

Using binary weights instead, the lag becomes a sum rather than an average.

```{r}
b_weights  <- lapply(wm_q, function(x) 0 * x + 1)
b_weights2 <- nb2listw(wm_q, glist = b_weights, style = "B")

lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
hunan <- left_join(hunan, lag.res, by = "NAME_3")

kable(head(hunan %>% st_drop_geometry() %>% select(NAME_3, GDPPC, `lag_sum GDPPC`), 6))
```

### 10.3 Spatial Window Average

The window average includes the region itself alongside its neighbours. `include.self()`
adds the diagonal.

```{r}
wm_qs <- include.self(wm_q)
wm_qs
```

```{r}
wm_qs <- nb2listw(wm_qs)
lag_w_avg_gpdpc <- lag.listw(wm_qs, hunan$GDPPC)

lag.list.wm_qs <- list(hunan$NAME_3, lag.listw(wm_qs, hunan$GDPPC))
lag_wm_qs.res <- as.data.frame(lag.list.wm_qs)
colnames(lag_wm_qs.res) <- c("NAME_3", "lag_window_avg GDPPC")
hunan <- left_join(hunan, lag_wm_qs.res, by = "NAME_3")

kable(head(hunan %>% st_drop_geometry() %>%
             select(NAME_3, GDPPC, `lag GDPPC`, `lag_window_avg GDPPC`), 6))
```

### 10.4 Spatial Window Sum

```{r}
wm_qs2 <- include.self(wm_q)
b_weights  <- lapply(wm_qs2, function(x) 0 * x + 1)
b_weights2 <- nb2listw(wm_qs2, glist = b_weights, style = "B")

w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
hunan <- left_join(hunan, w_sum_gdppc.res, by = "NAME_3")

kable(head(hunan %>% st_drop_geometry() %>%
             select(NAME_3, GDPPC, `lag_sum GDPPC`, `w_sum GDPPC`), 6))
```

### 10.5 Comparing the Four Lag Definitions

```{r}
#| fig-height: 8
lag_maps <- function(var, ttl) {
  tm_shape(hunan) +
    tm_polygons(fill = var,
                fill.scale = tm_scale_intervals(style = "quantile", values = "viridis"),
                fill.legend = tm_legend(show = FALSE),
                col = "white", lwd = 0.3) +
    tm_title(ttl)
}

tmap_arrange(lag_maps("lag GDPPC", "Lag (row-std average)"),
             lag_maps("lag_sum GDPPC", "Lag (binary sum)"),
             lag_maps("lag_window_avg GDPPC", "Window average"),
             lag_maps("w_sum GDPPC", "Window sum"),
             ncol = 2, nrow = 2)
```

## 11. My Own Exploration: Does the Choice of Weights Matrix Actually Matter?

Everything above follows the reference chapter. This section does not. Writing section
10 left me with a question the exercise never asks: I had just built five different
definitions of "neighbour" from one unchanged shapefile, and I had no idea whether that
choice changes any answer or is merely bookkeeping. Rather than assert it matters,
I decided to measure it.

The test is simple. Compute the spatially lagged GDP per capita under every weighting
scheme, then correlate the resulting vectors. If the correlations are near 1, the choice
is cosmetic. If they are not, every downstream statistic inherits the choice.

### 11.1 Building All Five Schemes

```{r}
wm_d62_e  <- dnearneigh(coords, 0, 62, longlat = TRUE)
knn6_e    <- knn2nb(knearneigh(coords, k = 6))
dist_e    <- nbdists(wm_q, coords, longlat = TRUE)
ids_e     <- lapply(dist_e, function(x) 1 / x)

lagged <- data.frame(
  Queen     = lag.listw(nb2listw(wm_q,    style = "W", zero.policy = TRUE), hunan$GDPPC),
  Rook      = lag.listw(nb2listw(wm_r,    style = "W", zero.policy = TRUE), hunan$GDPPC),
  Fixed62   = lag.listw(nb2listw(wm_d62_e,style = "W", zero.policy = TRUE), hunan$GDPPC,
                        zero.policy = TRUE),
  Adaptive6 = lag.listw(nb2listw(knn6_e,  style = "W", zero.policy = TRUE), hunan$GDPPC),
  IDW       = lag.listw(nb2listw(wm_q, glist = ids_e, style = "W", zero.policy = TRUE),
                        hunan$GDPPC)
)
```

### 11.2 How Connected Is Each Scheme?

```{r}
data.frame(
  Scheme   = c("Queen", "Rook", "Fixed 62 km", "Adaptive k=6"),
  Links    = c(sum(card(wm_q)), sum(card(wm_r)), sum(card(wm_d62_e)), sum(card(knn6_e))),
  Min      = c(min(card(wm_q)), min(card(wm_r)), min(card(wm_d62_e)), min(card(knn6_e))),
  Max      = c(max(card(wm_q)), max(card(wm_r)), max(card(wm_d62_e)), max(card(knn6_e))),
  Isolates = c(sum(card(wm_q) == 0), sum(card(wm_r) == 0),
               sum(card(wm_d62_e) == 0), sum(card(knn6_e) == 0))
)
```

```{r}
#| fig-height: 4
par(mfrow = c(1, 4), mar = c(4, 3, 3, 1))
for (nm in c("Queen", "Rook", "Fixed 62 km", "Adaptive k=6")) {
  nb <- switch(nm, "Queen" = wm_q, "Rook" = wm_r,
               "Fixed 62 km" = wm_d62_e, "Adaptive k=6" = knn6_e)
  hist(card(nb), breaks = 0:12, main = nm, xlab = "neighbours",
       col = "#c0392b", border = "white")
}
par(mfrow = c(1, 1))
```

### 11.3 Do the Lagged Values Agree?

```{r}
round(cor(lagged), 4)
```

```{r}
#| fig-height: 5
pairs(lagged, pch = 19, cex = 0.4, col = "#2980b9",
      main = "Lagged GDPPC under five weighting schemes")
```

::: {.callout-important}
## The answer, and it is not uniform
The correlations split the five schemes into two camps.

**Queen and rook are effectively the same thing here.** They correlate at **0.998**.
I had already guessed this in section 6.2 from the link counts, but guessing and
measuring are different, and now I can say it rather than hedge it.

**Contiguity and distance are not the same thing at all.** Queen against the fixed
62 km band correlates at only **0.875**. That is a real divergence. Roughly a quarter
of the variance in one lagged variable is unexplained by the other, from identical
input data, purely because of how I defined "neighbour".

So the honest summary is that the queen/rook decision is safe to make casually, and the
contiguity/distance decision is not. Those two choices are presented side by side in the
chapter as though they were the same kind of decision. They are not.
:::

### 11.4 Which Counties Disagree Most?

An average hides where the disagreement lives, so I looked at the individual counties
where queen contiguity and adaptive k=6 diverge most.

```{r}
pct_diff <- abs(lagged$Queen - lagged$Adaptive6) / lagged$Queen * 100
ord <- order(-pct_diff)[1:8]

data.frame(
  County        = hunan$NAME_3[ord],
  GDPPC         = hunan$GDPPC[ord],
  `Queen lag`   = round(lagged$Queen[ord]),
  `Adaptive lag`= round(lagged$Adaptive6[ord]),
  `Difference %`= round(pct_diff[ord], 1),
  `Queen nbrs`  = card(wm_q)[ord],
  check.names = FALSE
)
```

```{r}
#| fig-height: 5
hunan$lag_gap <- pct_diff
tm_shape(hunan) +
  tm_polygons(fill = "lag_gap",
              fill.scale = tm_scale_intervals(style = "quantile", values = "-viridis"),
              fill.legend = tm_legend(title = "Queen vs adaptive\ndisagreement (%)"),
              col = "white", lwd = 0.3) +
  tm_layout(frame = FALSE)
```

Before asserting a pattern from the top eight, I tested it across all 88 counties.

```{r}
data.frame(
  `Queen neighbours` = c("1 to 3", "4 to 6", "7 or more"),
  Counties = c(sum(card(wm_q) <= 3),
               sum(card(wm_q) >= 4 & card(wm_q) <= 6),
               sum(card(wm_q) >= 7)),
  `Mean disagreement %` = round(c(mean(pct_diff[card(wm_q) <= 3]),
                                  mean(pct_diff[card(wm_q) >= 4 & card(wm_q) <= 6]),
                                  mean(pct_diff[card(wm_q) >= 7])), 1),
  check.names = FALSE
)

cor(card(wm_q), pct_diff)
```

::: {.callout-note}
## Where the disagreement concentrates, and how strong that pattern really is
My first reading of the top-eight table was that poorly connected counties disagree
most. Longshan is the worst case at about **32%** and has only 3 queen neighbours,
which fits neatly.

Then I checked it properly, and the neat story is only partly true. Counties with 3 or
fewer neighbours do disagree more on average (**10.4%** against roughly **7.2%** for
everyone else), but the correlation between neighbour count and disagreement is only
about **-0.18**. That is a weak tendency, not a rule. Taojiang sits second in the table
with a 25% gap and has 7 neighbours, which the story does not explain at all.

The mechanism I can defend is narrower than the one I first reached for. If a county
has three contiguous neighbours and adaptive weighting forces it to six, three
non-touching counties each receive a sixth of the weight, which can move the average a
long way. That explains Longshan. It does not explain Taojiang, where something else is
going on, most likely a neighbour with an extreme GDPPC entering or leaving the set.

What I take from this is less about Hunan than about my own habit. I had the
explanation written before I had the correlation, and the correlation turned out to be
0.18 rather than the 0.6 or 0.7 my confident phrasing implied. Ranking a table by one
column and reading a cause off the top rows is a good way to generate a hypothesis and
a bad way to confirm one.
:::

## 12. What I Take From This Exercise

**The weights matrix is a modelling assumption, not a data property.** Nothing in the
Hunan shapefile tells me whether two counties are neighbours. Queen or rook, 62 km or
6 nearest neighbours, binary or inverse-distance, row-standardised or not. Each is a
decision I make, and each produces a different matrix from identical data. Every
autocorrelation statistic and spatial regression later in the course inherits whichever
one I pick. This exercise is not really about `spdep` syntax; it is about the fact that
a large part of a spatial analysis is decided before any statistic is computed.

**The silent-failure lesson.** The `sfc`-versus-matrix mistake in section 7.1 is the one
I will actually remember. It did not throw an error, it did not produce an obviously
absurd map, and it yielded a valid weights object. It was simply measuring degrees and
calling them kilometres. The only thing that caught it was noticing that a
first-nearest-neighbour distance of 0.6 km between Chinese counties is impossible. Unit
plausibility checks are not busywork; here they were the only line of defence.

**Averages versus sums matter more than they look.** The four lag definitions in section
10 produce visibly different maps from the same variable and the same neighbour list.
The binary-sum versions are partly just a map of *how many neighbours each county has*,
which is a property of the geography rather than of GDP. Row standardisation removes
that artefact, which is why it is the sensible default for comparing across regions of
uneven connectivity.

**Connecting forward.** The row-standardised queen contiguity matrix built here is the
direct input to Moran's I and Geary's C in the next exercise. Section 11 gives me a
concrete expectation to carry into that: switching between queen and rook should barely
move Moran's I, while switching from contiguity to a distance band plausibly could. I
intend to run the autocorrelation statistic under at least two weighting schemes rather
than report a single number as though the matrix behind it were a fact about Hunan
instead of a decision I made.
