---
title: "Hands-on Exercise 1: Geospatial Data Wrangling with R"
date: last-modified
format:
  html:
    code-fold: false
execute:
  warning: false
  message: false
---

## 1. Overview

This exercise works through the core geospatial data science workflow in R: importing
geospatial and aspatial data, checking and setting coordinate reference systems,
converting aspatial data to spatial data, and performing basic geoprocessing with the
`sf` package.

## 2. Getting Started

```{r}
pacman::p_load(sf, tidyverse)
```

## 3. Data Acquisition

Three datasets are used:

| Dataset | Source | Format used |
|---|---|---|
| Master Plan 2014 Subzone Boundary (No Sea) | [data.gov.sg](https://data.gov.sg/datasets/d_226cacceceff94f0c8b814962a5307c9/view) | GeoJSON |
| Pre-Schools Location | [data.gov.sg](https://data.gov.sg/datasets/d_61eefab99958fd70e6aab17320a71f1c/view) | GeoJSON |
| Singapore Airbnb listings | [Inside Airbnb](https://insideairbnb.com/singapore/) | CSV |

**Note on format:** the original exercise reference uses the Master Plan boundary and
pre-school location layers in Shapefile/KML format. Both have since been retired on the
data.gov.sg portal in favour of GeoJSON-only downloads, so this write-up imports the
GeoJSON versions directly with `st_read()` instead. The underlying `sf` workflow is
identical regardless of source format.

The Cycling Path dataset from LTA DataMall (used in the reference chapter's buffer
geoprocessing example) requires a DataMall API account to download, which I did not have
set up in time for this submission. That section is noted but not run below; I'll revisit
it once I have API access.

## 4. Importing Geospatial Data

### 4.1 Master Plan 2014 Subzone Boundary

```{r}
mpsz <- st_read(dsn = "data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
```

### 4.2 Pre-Schools Location

```{r}
preschool <- st_read("data/geospatial/PreSchoolsLocation.geojson")
```

## 5. Checking the Content of a Simple Feature Data Frame

```{r}
st_geometry(mpsz)
```

```{r}
glimpse(mpsz)
```

```{r}
head(mpsz, n = 5)
```

## 6. Plotting the Geospatial Data

```{r}
plot(st_geometry(mpsz))
```

```{r}
plot(mpsz["PLN_AREA_N"])
```

## 7. Working with Projection

Checking the coordinate reference system of each layer:

```{r}
st_crs(mpsz)
```

```{r}
st_crs(preschool)
```

Both layers come in as WGS84 (EPSG:4326 / CRS84), Singapore's projected coordinate
system is SVY21 (EPSG:3414), which is needed for any accurate area or distance
calculation. Transforming both layers:

```{r}
mpsz3414 <- st_transform(mpsz, crs = 3414)
preschool3414 <- st_transform(preschool, crs = 3414)
```

```{r}
st_crs(mpsz3414)
```

## 8. Importing and Converting an Aspatial Data Set

The Airbnb listings CSV has `latitude`/`longitude` columns but no spatial geometry —
it needs to be converted into an `sf` object before it can be used alongside the other
two layers.

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

```{r}
listings_sf <- st_as_sf(listings,
                         coords = c("longitude", "latitude"),
                         crs = 4326) %>%
  st_transform(crs = 3414)
```

```{r}
listings_sf
```

## 9. Geoprocessing with sf Package

### 9.1 Point-in-polygon count: pre-schools per subzone

```{r}
mpsz3414$`PreSch Count` <- lengths(st_intersects(mpsz3414, preschool3414))
```

```{r}
summary(mpsz3414$`PreSch Count`)
```

### 9.2 Subzone area and pre-school density

```{r}
mpsz3414$Area <- mpsz3414 %>%
  st_area()

mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = as.numeric(`PreSch Count`) / (as.numeric(Area) / 1000000))
```

```{r}
ggplot(data = mpsz3414, aes(x = `PreSch Density`)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue") +
  labs(title = "Distribution of pre-school density by subzone",
       x = "Pre-school density (per sq km)",
       y = "Number of subzones")
```

```{r}
ggplot(data = mpsz3414, aes(y = `PreSch Count`, x = `PreSch Density`)) +
  geom_point(color = "black", fill = "light blue") +
  labs(title = "Pre-school count vs. density by subzone",
       x = "Pre-school density (per sq km)",
       y = "Pre-school count")
```

## 10. Reflections

The main friction point in this exercise wasn't the `sf` workflow itself, it was that
the reference chapter's data sources have moved on: data.gov.sg no longer serves the
Master Plan boundary or pre-school layers as Shapefile/KML, only GeoJSON, and the LTA
DataMall cycling path layer sits behind an API key I hadn't set up yet. `st_read()`
handles GeoJSON exactly the same way it handles a Shapefile or KML, so the substitution
didn't change anything downstream — the point-in-polygon and density workflow ran the
same way once both layers were reprojected to EPSG:3414.
