---
title: "Hands-on Exercise 1"
subtitle: "Geospatial Data Wrangling with R, and Choropleth Mapping with R"
date: last-modified
format:
  html:
    code-fold: false
    toc: true
execute:
  warning: false
  message: false
---

This exercise is in two parts, following the two Hands-on Exercise 1 readings listed
under Lesson 1 of the course outline: [Geospatial Data Wrangling with
R](https://r4gdsa.netlify.app/chap01.html) and [Choropleth Mapping with
R](https://r4gdsa.netlify.app/chap02.html).

# Part 1: Geospatial Data Wrangling with R {#part1}

## 1. Overview

This part works through the core geospatial data science workflow in R: importing
geospatial and aspatial data, checking and setting coordinate reference systems,
converting aspatial data to spatial data, and performing basic geoprocessing with the
`sf` package.

## 2. Getting Started

```{r}
pacman::p_load(sf, tidyverse)
```

## 3. Data Acquisition

Three datasets are used:

| Dataset | Source | Format used |
|---|---|---|
| Master Plan 2014 Subzone Boundary (No Sea) | [data.gov.sg](https://data.gov.sg/datasets/d_226cacceceff94f0c8b814962a5307c9/view) | GeoJSON |
| Pre-Schools Location | [data.gov.sg](https://data.gov.sg/datasets/d_61eefab99958fd70e6aab17320a71f1c/view) | GeoJSON |
| Cycling Path Network | [data.gov.sg](https://data.gov.sg/datasets/d_8f468b25193f64be8a16fa7d8f60f553/view) | GeoJSON |
| Singapore Airbnb listings | [Inside Airbnb](https://insideairbnb.com/singapore/) | CSV |

**Note on format:** the original exercise reference uses the Master Plan boundary,
pre-school location, and cycling path layers in Shapefile/KML format. All three have
since been retired on the data.gov.sg portal in favour of GeoJSON-only downloads, so
this write-up imports the GeoJSON versions directly with `st_read()` instead. The
underlying `sf` workflow is identical regardless of source format — only the CRS each
layer arrives in differs (all three come in as WGS84 here, rather than the mix of
WGS84/SVY21 the reference chapter describes).

## 4. Importing Geospatial Data

### 4.1 Master Plan 2014 Subzone Boundary

```{r}
mpsz <- st_read(dsn = "data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
```

### 4.2 Pre-Schools Location

```{r}
preschool <- st_read("data/geospatial/PreSchoolsLocation.geojson")
```

### 4.3 Cycling Path Network

```{r}
cyclingpath <- st_read("data/geospatial/CyclingPathNetwork.geojson")
```

## 5. Checking the Content of a Simple Feature Data Frame

```{r}
st_geometry(mpsz)
```

```{r}
glimpse(mpsz)
```

```{r}
head(mpsz, n = 5)
```

## 6. Plotting the Geospatial Data

```{r}
plot(st_geometry(mpsz))
```

```{r}
plot(mpsz["PLN_AREA_N"])
```

## 7. Working with Projection

Checking the coordinate reference system of each layer:

```{r}
st_crs(mpsz)
```

```{r}
st_crs(preschool)
```

Both layers come in as WGS84 (EPSG:4326 / CRS84), Singapore's projected coordinate
system is SVY21 (EPSG:3414), which is needed for any accurate area or distance
calculation. Transforming both layers:

```{r}
mpsz3414 <- st_transform(mpsz, crs = 3414)
preschool3414 <- st_transform(preschool, crs = 3414)
cyclingpath3414 <- st_transform(cyclingpath, crs = 3414)
```

```{r}
st_crs(mpsz3414)
```

## 8. Importing and Converting an Aspatial Data Set

The Airbnb listings CSV has `latitude`/`longitude` columns but no spatial geometry —
it needs to be converted into an `sf` object before it can be used alongside the other
two layers.

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

```{r}
listings_sf <- st_as_sf(listings,
                         coords = c("longitude", "latitude"),
                         crs = 4326) %>%
  st_transform(crs = 3414)
```

```{r}
listings_sf
```

## 9. Geoprocessing with sf Package

### 9.1 Buffer analysis: land needed to widen cycling paths

Scenario: the land authority wants to widen every cycling path by 5m on each side, and
needs to know how much total land will need to be acquired.

```{r}
buffer_cycling <- st_buffer(cyclingpath3414, dist = 5, nQuadSegs = 30)
buffer_cycling$AREA <- st_area(buffer_cycling)
sum(buffer_cycling$AREA)
```

### 9.2 Point-in-polygon count: pre-schools per subzone

```{r}
mpsz3414$`PreSch Count` <- lengths(st_intersects(mpsz3414, preschool3414))
```

```{r}
summary(mpsz3414$`PreSch Count`)
```

### 9.3 Subzone area and pre-school density

```{r}
mpsz3414$Area <- mpsz3414 %>%
  st_area()

mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = as.numeric(`PreSch Count`) / (as.numeric(Area) / 1000000))
```

```{r}
ggplot(data = mpsz3414, aes(x = `PreSch Density`)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue") +
  labs(title = "Distribution of pre-school density by subzone",
       x = "Pre-school density (per sq km)",
       y = "Number of subzones")
```

```{r}
ggplot(data = mpsz3414, aes(y = `PreSch Count`, x = `PreSch Density`)) +
  geom_point(color = "black", fill = "light blue") +
  labs(title = "Pre-school count vs. density by subzone",
       x = "Pre-school density (per sq km)",
       y = "Pre-school count")
```

## 10. Reflections

The main friction point in this exercise wasn't the `sf` workflow itself, it was that
the reference chapter's data sources have moved on: data.gov.sg no longer serves the
Master Plan boundary, pre-school, or cycling path layers as Shapefile/KML, only
GeoJSON. `st_read()` handles GeoJSON exactly the same way it handles a Shapefile or
KML, so the substitution didn't change anything downstream — the buffer, point-in-
polygon, and density workflows all ran the same way once every layer was reprojected
to EPSG:3414. The one thing worth double-checking whenever a dataset gets swapped like
this is what coordinate system it actually arrives in — all three geospatial layers
here came in as WGS84, whereas the reference chapter has some of them (the cycling
path, in particular) already in SVY21.

# Part 2: Choropleth Mapping with R {#part2}

## 1. Overview

This part covers thematic and choropleth mapping using `tmap`: importing and joining
geospatial and aspatial data, plotting choropleth maps with `tmap`'s grammar-of-graphics
style API, comparing classification methods, styling with colour schemes and
cartographic furniture, small multiples, and one interactive map.

## 2. Getting Started

```{r}
pacman::p_load(sf, tmap, tidyverse)
packageVersion("tmap")
```

## 3. Data Used

| Dataset | Source | Format used |
|---|---|---|
| Master Plan 2014 Subzone Boundary (No Sea) | [data.gov.sg](https://data.gov.sg/datasets/d_226cacceceff94f0c8b814962a5307c9/view) | GeoJSON |
| Resident Population by Planning Area/Subzone, Age Group and Sex (Census of Population 2020) | [data.gov.sg](https://data.gov.sg/datasets/d_d95ae740c0f8961a0b10435836660ce0/view) | CSV |

**Note on substitution:** the reference chapter uses an annual "Singapore Residents by
Planning Area/Subzone, Age Group, Sex and Type of Dwelling" release, which has since
been retired from data.gov.sg's catalogue (it no longer turns up under its old file
name `respopagesextod`, or under a search for its own title). The nearest dataset still
published at subzone level is the Census of Population 2020 table used here. It has a
different shape to the retired series: it comes as one wide table with a single
`Number` column that mixes planning-area subtotal rows (suffixed `- Total`) with
subzone rows, rather than separate `PA` and `SZ` columns, so the wrangling below spells
out the extra reshaping step that the original exercise didn't need. It also reuses the
Master Plan **2014** boundary from Part 1 rather than 2019, so both parts of this
exercise share the same base map.

## 4. Importing and Preparing the Data

### 4.1 Geospatial data

Reusing the subzone boundary imported in Part 1 rather than re-reading it:

```{r}
mpsz <- mpsz3414
```

### 4.2 Aspatial data

```{r}
popdata <- read_csv("data/aspatial/respop2020.csv", na = "-")
```

The `na = "-"` argument matters here: SingStat suppresses small subzone counts with a
literal `-`, and without telling `read_csv()` to treat that as a missing value, every
numeric column in the table gets silently parsed as text.

### 4.3 Data wrangling

The `Number` column mixes two kinds of row: planning-area subtotals, named like
`"Ang Mo Kio - Total"`, and the individual subzone rows that follow each one. A
forward-fill turns that into a proper `PA` column before the subtotal rows are dropped:

```{r}
pop_sz <- popdata %>%
  filter(Number != "Total") %>%
  mutate(PA = if_else(str_detect(Number, "-\\s*Total$"),
                       str_remove(Number, "-\\s*Total$") %>% str_trim(),
                       NA_character_)) %>%
  fill(PA, .direction = "down") %>%
  filter(!str_detect(Number, "-\\s*Total$")) %>%
  rename(SZ = Number)
```

::: {.callout-note}
## Data quality quirk
One planning-area subtotal row breaks the `" - Total"` pattern: `Changi` is written as
`"Changi- Total"`, with no space before the dash, while every other area uses
`"Changi Bay - Total"` with a space. The regex above (`-\s*Total$`) allows for zero or
more spaces so it still catches it — a plain `" - Total$"` match would have let
`"Changi- Total"` slip through as if it were a subzone named `"Changi- Total"`, similar
in spirit to the `YISHUN`/`Yishun` case-mismatch flagged in the KNIME Data Analytics Lab
exercise.
:::

Computing the young / economically active / aged breakdown and the dependency ratio:

```{r}
pop_sz <- pop_sz %>%
  mutate(
    YOUNG = Total_0_4 + Total_5_9 + Total_10_14 + Total_15_19 + Total_20_24,
    `ECONOMY ACTIVE` = Total_25_29 + Total_30_34 + Total_35_39 + Total_40_44 +
      Total_45_49 + Total_50_54 + Total_55_59 + Total_60_64,
    AGED = Total_65_69 + Total_70_74 + Total_75_79 + Total_80_84 + Total_85_89 +
      Total_90andOver,
    TOTAL = Total_Total,
    DEPENDENCY = (YOUNG + AGED) / `ECONOMY ACTIVE`
  ) %>%
  select(PA, SZ, YOUNG, `ECONOMY ACTIVE`, AGED, TOTAL, DEPENDENCY) %>%
  mutate(SZ = toupper(SZ), PA = toupper(PA)) %>%
  filter(!is.na(`ECONOMY ACTIVE`), `ECONOMY ACTIVE` > 0)

nrow(pop_sz)
```

### 4.4 Joining geospatial and aspatial data

```{r}
mpsz_pop2020 <- left_join(mpsz, pop_sz, by = c("SUBZONE_N" = "SZ"))
sum(!is.na(mpsz_pop2020$DEPENDENCY))
```

201 of 323 subzones get a dependency ratio. The rest (industrial estates, parks, water
catchment areas, and a handful of 2014-vs-2020 subzone boundary renames) have no
resident population in the 2020 Census and are left as `NA` rather than dropped, so the
map below shows them un-shaded instead of silently disappearing.

## 5. Choropleth Mapping with tmap

### 5.1 A quick map with `qtm()`

```{r}
tmap_mode("plot")
qtm(shp = mpsz_pop2020, fill = "DEPENDENCY")
```

### 5.2 A fuller map built from tmap's elements

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_title("Dependency ratio by planning subzone, Singapore") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2) +
  tm_credits(
    "Source: Master Plan 2014 Subzone Boundary (URA) and Census of Population 2020 (SingStat)",
    position = c("left", "bottom"))
```

### 5.3 Comparing classification methods

*DIY: prepare choropleth maps using different classification methods and compare.*

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile")
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "equal", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval")
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "jenks", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Jenks natural breaks")
```

Quantile forces the same number of subzones into every bucket, which spreads colour
evenly across the map regardless of how the underlying values are actually
distributed — it's the most visually "balanced" of the three but can group very
different dependency ratios into one class if the data clusters tightly. Equal
interval slices the value range into five equal-width bands; because most subzones sit
in a fairly narrow band, that concentrates almost every subzone into one or two classes
and leaves the rest of the legend almost empty. Jenks looks for natural gaps in the
data itself, so the class breaks land where the dependency ratio actually jumps, which
here sits closer to quantile than to equal interval.

### 5.4 Comparing number of classes

*DIY: prepare choropleth maps using the same classification method but with 2, 6, 10,
and 20 classes, and compare.*

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 2)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 2")
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 6)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 6")
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 10)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 10")
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 20)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 20")
```

At `n = 2` the map only distinguishes "below median" from "above median" dependency,
which hides all the variation within each half. Pushing up to `n = 10` or `n = 20`
starts creating classes so narrow that neighbouring subzones with almost identical
ratios get visibly different colours, which reads as more precision than the data
actually supports — with only 201 subzones, 20 quantile bins means roughly 10 subzones
per class, each one a razor-thin slice of the value range. Something in the 5–6 range
is where the map still shows a clear spatial pattern without either flattening the
variation or manufacturing false precision.

## 6. Small Multiples

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = c("YOUNG", "AGED"),
    fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
    fill.scale = tm_scale_intervals(style = "equal", n = 5, values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5)
```

## 7. Interactive Map

```{r}
tmap_mode("view")
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c("Name" = "SUBZONE_N", "Dependency" = "DEPENDENCY")) +
  tm_borders()
```

```{r}
tmap_mode("plot")
```

## 8. Reflections

The biggest difference from the reference chapter was upstream of any mapping code:
the exact population dataset it names has been retired, so getting from "a CSV with an
age/sex breakdown" to "one row per subzone with a dependency ratio" took real wrangling
rather than a straight read-and-join. The `Number` column mixing planning-area
subtotals with subzone rows, `-` used for suppressed counts, and the one inconsistent
`"Changi- Total"` label were all things I only found by actually inspecting the output
at each step rather than assuming the code from the reference chapter would apply
unchanged. Once the join was right, the tmap classification and class-count comparisons
were the part that most changed how I'd read a choropleth map going forward — the
classification method and the number of classes are both editorial choices, not neutral
defaults, and they can make the same underlying data look like a strong spatial pattern
or a fairly flat one.
