---
title: "Choropleth Mapping with R"
subtitle: "HANDS-ON EXERCISE 1 · PART 02"
author:
  - name: Jedidiah Asaf Tallulembang
    affiliation: "ISSS626 — Geospatial Analytics and Applications"
date: last-modified
format:
  html:
    code-fold: false
    toc: true
execute:
  warning: false
  message: false
---

::: {.pill-row}
[R]{.pill} [sf]{.pill} [tmap]{.pill} [tidyverse]{.pill} [Choropleth Mapping]{.pill}
:::

This exercise follows the second of two Hands-on Exercise 1 readings listed under
Lesson 1 of the course outline: [Choropleth Mapping with
R](https://r4gdsa.netlify.app/chap02.html). It builds on the subzone boundary
imported in [Part 01](Part01.qmd), re-imported below so this page can run on its own.

## 1. Overview

This part covers thematic and choropleth mapping using `tmap`: importing and joining
geospatial and aspatial data, plotting choropleth maps with `tmap`'s grammar-of-graphics
style API, comparing classification methods, styling with colour schemes and
cartographic furniture, small multiples, and one interactive map.

## 2. Getting Started

```{r}
pacman::p_load(sf, tmap, tidyverse)
packageVersion("tmap")
```

## 3. Data Used

| Dataset | Source | Format used |
|---|---|---|
| Master Plan 2014 Subzone Boundary (No Sea) | [data.gov.sg](https://data.gov.sg/datasets/d_226cacceceff94f0c8b814962a5307c9/view) | GeoJSON |
| Resident Population by Planning Area/Subzone, Age Group and Sex (Census of Population 2020) | [data.gov.sg](https://data.gov.sg/datasets/d_d95ae740c0f8961a0b10435836660ce0/view) | CSV |

**Note on substitution:** the reference chapter uses an annual "Singapore Residents by
Planning Area/Subzone, Age Group, Sex and Type of Dwelling" release, which has since
been retired from data.gov.sg's catalogue (it no longer turns up under its old file
name `respopagesextod`, or under a search for its own title). The nearest dataset still
published at subzone level is the Census of Population 2020 table used here. It has a
different shape to the retired series: it comes as one wide table with a single
`Number` column that mixes planning-area subtotal rows (suffixed `- Total`) with
subzone rows, rather than separate `PA` and `SZ` columns, so the wrangling below spells
out the extra reshaping step that the original exercise didn't need. It also reuses the
Master Plan **2014** boundary from Part 01 rather than 2019, so both parts of this
exercise share the same base map.

## 4. Importing and Preparing the Data

### 4.1 Geospatial data

Re-importing the same Master Plan 2014 subzone boundary used in
[Part 01](Part01.qmd) — each exercise page runs as its own self-contained R session,
so the object isn't carried over automatically:

```{r}
mpsz <- st_read(dsn = "data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson") %>%
  st_transform(crs = 3414)
```

### 4.2 Aspatial data

```{r}
popdata <- read_csv("data/aspatial/respop2020.csv", na = "-")
```

The `na = "-"` argument matters here: SingStat suppresses small subzone counts with a
literal `-`, and without telling `read_csv()` to treat that as a missing value, every
numeric column in the table gets silently parsed as text.

### 4.3 Data wrangling

The `Number` column mixes two kinds of row: planning-area subtotals, named like
`"Ang Mo Kio - Total"`, and the individual subzone rows that follow each one. A
forward-fill turns that into a proper `PA` column before the subtotal rows are dropped:

```{r}
pop_sz <- popdata %>%
  filter(Number != "Total") %>%
  mutate(PA = if_else(str_detect(Number, "-\\s*Total$"),
                       str_remove(Number, "-\\s*Total$") %>% str_trim(),
                       NA_character_)) %>%
  fill(PA, .direction = "down") %>%
  filter(!str_detect(Number, "-\\s*Total$")) %>%
  rename(SZ = Number)
```

::: {.callout-note}
## Data quality quirk
One planning-area subtotal row breaks the `" - Total"` pattern: `Changi` is written as
`"Changi- Total"`, with no space before the dash, while every other area uses
`"Changi Bay - Total"` with a space. The regex above (`-\s*Total$`) allows for zero or
more spaces so it still catches it — a plain `" - Total$"` match would have let
`"Changi- Total"` slip through as if it were a subzone named `"Changi- Total"`, similar
in spirit to the `YISHUN`/`Yishun` case-mismatch flagged in the KNIME Data Analytics Lab
exercise.
:::

Computing the young / economically active / aged breakdown and the dependency ratio:

```{r}
pop_sz <- pop_sz %>%
  mutate(
    YOUNG = Total_0_4 + Total_5_9 + Total_10_14 + Total_15_19 + Total_20_24,
    `ECONOMY ACTIVE` = Total_25_29 + Total_30_34 + Total_35_39 + Total_40_44 +
      Total_45_49 + Total_50_54 + Total_55_59 + Total_60_64,
    AGED = Total_65_69 + Total_70_74 + Total_75_79 + Total_80_84 + Total_85_89 +
      Total_90andOver,
    TOTAL = Total_Total,
    DEPENDENCY = (YOUNG + AGED) / `ECONOMY ACTIVE`
  ) %>%
  select(PA, SZ, YOUNG, `ECONOMY ACTIVE`, AGED, TOTAL, DEPENDENCY) %>%
  mutate(SZ = toupper(SZ), PA = toupper(PA)) %>%
  filter(!is.na(`ECONOMY ACTIVE`), `ECONOMY ACTIVE` > 0)

nrow(pop_sz)
```

### 4.4 Joining geospatial and aspatial data

```{r}
mpsz_pop2020 <- left_join(mpsz, pop_sz, by = c("SUBZONE_N" = "SZ"))
sum(!is.na(mpsz_pop2020$DEPENDENCY))
```

201 of 323 subzones get a dependency ratio. The rest (industrial estates, parks, water
catchment areas, and a handful of 2014-vs-2020 subzone boundary renames) have no
resident population in the 2020 Census and are left as `NA` rather than dropped, so the
map below shows them un-shaded instead of silently disappearing.

## 5. Choropleth Mapping with tmap

### 5.1 A quick map with `qtm()`

```{r}
tmap_mode("plot")
qtm(shp = mpsz_pop2020, fill = "DEPENDENCY")
```

### 5.2 A fuller map built from tmap's elements

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_title("Dependency ratio by planning subzone, Singapore") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2) +
  tm_credits(
    "Source: Master Plan 2014 Subzone Boundary (URA) and Census of Population 2020 (SingStat)",
    position = c("left", "bottom"))
```

### 5.3 Comparing classification methods

*DIY: prepare choropleth maps using different classification methods and compare.*

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile")
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "equal", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval")
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "jenks", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Jenks natural breaks")
```

Quantile forces the same number of subzones into every bucket, which spreads colour
evenly across the map regardless of how the underlying values are actually
distributed — it's the most visually "balanced" of the three but can group very
different dependency ratios into one class if the data clusters tightly. Equal
interval slices the value range into five equal-width bands; because most subzones sit
in a fairly narrow band, that concentrates almost every subzone into one or two classes
and leaves the rest of the legend almost empty. Jenks looks for natural gaps in the
data itself, so the class breaks land where the dependency ratio actually jumps, which
here sits closer to quantile than to equal interval.

### 5.4 Comparing number of classes

*DIY: prepare choropleth maps using the same classification method but with 2, 6, 10,
and 20 classes, and compare.*

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 2)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 2")
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 6)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 6")
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 10)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 10")
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "quantile", n = 20)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 20")
```

At `n = 2` the map only distinguishes "below median" from "above median" dependency,
which hides all the variation within each half. Pushing up to `n = 10` or `n = 20`
starts creating classes so narrow that neighbouring subzones with almost identical
ratios get visibly different colours, which reads as more precision than the data
actually supports — with only 201 subzones, 20 quantile bins means roughly 10 subzones
per class, each one a razor-thin slice of the value range. Something in the 5–6 range
is where the map still shows a clear spatial pattern without either flattening the
variation or manufacturing false precision.

## 6. Small Multiples

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = c("YOUNG", "AGED"),
    fill.legend = tm_legend(position = tm_pos_in("right", "bottom")),
    fill.scale = tm_scale_intervals(style = "equal", n = 5, values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5)
```

## 7. Interactive Map

```{r}
tmap_mode("view")
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c("Name" = "SUBZONE_N", "Dependency" = "DEPENDENCY")) +
  tm_borders()
```

```{r}
tmap_mode("plot")
```

## 8. Reflections

The biggest difference from the reference chapter was upstream of any mapping code:
the exact population dataset it names has been retired, so getting from "a CSV with an
age/sex breakdown" to "one row per subzone with a dependency ratio" took real wrangling
rather than a straight read-and-join. The `Number` column mixing planning-area
subtotals with subzone rows, `-` used for suppressed counts, and the one inconsistent
`"Changi- Total"` label were all things I only found by actually inspecting the output
at each step rather than assuming the code from the reference chapter would apply
unchanged. Once the join was right, the tmap classification and class-count comparisons
were the part that most changed how I'd read a choropleth map going forward — the
classification method and the number of classes are both editorial choices, not neutral
defaults, and they can make the same underlying data look like a strong spatial pattern
or a fairly flat one.

::: {.page-nav-prev}
[← Part 01: Geospatial Data Wrangling with R](Part01.qmd)
:::
