---
title: "Geographical Segmentation with Spatially Constrained Clustering"
subtitle: "HANDS-ON EXERCISE 6"
author:
  - name: Jedidiah Asaf Tallulembang
    affiliation: "ISSS626 Geospatial Analytics and Applications"
date: last-modified
format:
  html:
    code-fold: false
    toc: true
execute:
  warning: false
  message: false
---

::: {.pill-row}
[R]{.pill} [ClustGeo]{.pill} [SKATER]{.pill} [Ward]{.pill} [Minimum spanning tree]{.pill}
:::

This exercise follows [Geographical Segmentation with Spatially Constrained Clustering
Techniques](https://r4gdsa.netlify.app/chap12.html), the reading for the Geographic
Segmentation block of the course.

## 1. Overview

Exercises 4 and 5 asked whether values were spatially clustered. This one asks a
different question: given a set of multivariate measurements attached to areal units,
**how do I carve the study area into a small number of internally homogeneous regions?**

That is segmentation rather than detection. The practical use is planning: a telecoms
regulator, a market-entry team or a ministry allocating infrastructure does not want 55
townships, it wants five or six coherent zones it can write a policy for.

The complication is that ordinary cluster analysis knows nothing about geography. It
will happily place two townships at opposite ends of Shan State in the same cluster
because their radio ownership happens to match. A region you cannot draw on a map is
not much use as a planning unit. Two families of method address this:

- **SKATER**, which prunes a minimum spanning tree built on the contiguity graph, so
  contiguity is enforced by construction;
- **ClustGeo**, which blends an attribute dissimilarity matrix with a geographic
  distance matrix through a mixing parameter `alpha`, trading homogeneity for
  compactness continuously.

By the end I should be able to run all three (plain hierarchical, SKATER, ClustGeo) and
say something defensible about what each one costs.

## 2. The Study Area and Data

| Dataset | Description | Format |
|---|---|---|
| `myanmar_township_boundaries` | Township boundaries for Myanmar, filtered to the three Shan states | ESRI Shapefile (polygon) |
| `Shan-ICT.csv` | Household ICT ownership from the 2014 Myanmar Population and Housing Census, township level | CSV |

Both come from the [Myanmar Information Management Unit
(MIMU)](http://themimu.info/). The filter keeps Shan (East), Shan (North) and Shan
(South), which gives 55 townships.

Shan State is a reasonable choice for this problem. It is large, internally very uneven,
and the ICT variables are genuinely multivariate: radio is a legacy technology with a
different geography from mobile phones, so the clustering has something to work with
rather than six proxies for one underlying wealth gradient.

## 3. Getting Started

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo,
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally)
```

## 4. Data Import and Preparation

### 4.1 Importing the township boundaries

```{r}
shan_sf <- st_read(dsn = "data/geospatial",
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))
```

```{r}
glimpse(shan_sf)
```

55 townships, six attribute fields and a geometry column, in WGS 84.

### 4.2 Importing the ICT data

```{r}
ict <- read_csv("data/aspatial/Shan-ICT.csv")
summary(ict)
```

### 4.3 Deriving penetration rates

The raw counts are households owning each device. Using them directly would cluster
townships by **size**, not by technology profile: a township with 80,000 households will
have more of everything. Converting to a rate per 1,000 households removes that.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`, `DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`,
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`)

summary(ict_derived)
```

## 5. Exploratory Data Analysis

### 5.1 Counts versus rates

```{r}
#| fig-height: 3
ggarrange(
  ggplot(ict_derived, aes(x = RADIO)) +
    geom_histogram(bins = 20, color = "black", fill = "light blue") +
    ggtitle("Households owning a radio (count)"),
  ggplot(ict_derived, aes(x = RADIO_PR)) +
    geom_histogram(bins = 20, color = "black", fill = "light blue") +
    ggtitle("Radio per 1,000 households"),
  ncol = 2)
```

The count distribution is strongly right-skewed, which is a size effect: a handful of
large townships sit far out in the tail. The rate is much closer to symmetric. This is
the whole justification for section 4.3 in one picture.

### 5.2 All six penetration rates

```{r}
#| fig-height: 5
radio <- ggplot(ict_derived, aes(x = RADIO_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")
tv <- ggplot(ict_derived, aes(x = TV_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")
llphone <- ggplot(ict_derived, aes(x = LLPHONE_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")
mphone <- ggplot(ict_derived, aes(x = MPHONE_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")
computer <- ggplot(ict_derived, aes(x = COMPUTER_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")
internet <- ggplot(ict_derived, aes(x = INTERNET_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue")

ggarrange(radio, tv, llphone, mphone, computer, internet,
          ncol = 3, nrow = 2)
```

Radio and TV are broad and roughly symmetric. Landline, computer and internet are all
heavily right-skewed with a floor near zero, which is what low-penetration technology
looks like: most townships have almost none, a few urban ones have some.

### 5.3 Joining geospatial and aspatial data

```{r}
shan_sf <- left_join(shan_sf, ict_derived, by = c("TS_PCODE" = "TS_PCODE"))
```

```{r}
sum(!is.na(shan_sf$TT_HOUSEHOLDS))
```

All 55 townships matched on `TS_PCODE`. Worth checking rather than assuming, because a
left join silently produces `NA` rather than failing.

### 5.4 Choropleths: why the rate matters

```{r}
#| fig-height: 4
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) +
  tm_polygons(fill = "TT_HOUSEHOLDS",
              fill.scale = tm_scale_intervals(n = 5, style = "jenks"),
              fill.legend = tm_legend(title = "Total households")) +
  tm_borders(col_alpha = 0.5)

RADIO.map <- tm_shape(shan_sf) +
  tm_polygons(fill = "RADIO",
              fill.scale = tm_scale_intervals(n = 5, style = "jenks"),
              fill.legend = tm_legend(title = "Radio (count)")) +
  tm_borders(col_alpha = 0.5)

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map, asp = NA, ncol = 2)
```

The two maps are near-identical. The count map is essentially a population map.

```{r}
#| fig-height: 4
RADIO_PR.map <- tm_shape(shan_sf) +
  tm_polygons(fill = "RADIO_PR",
              fill.scale = tm_scale_intervals(n = 5, style = "jenks"),
              fill.legend = tm_legend(title = "Radio per 1,000")) +
  tm_borders(col_alpha = 0.5)

tmap_arrange(TT_HOUSEHOLDS.map, RADIO_PR.map, asp = NA, ncol = 2)
```

Once normalised the pattern inverts in places: some of the *least* populous townships
have the *highest* radio penetration. That is a real signal, and it was completely
hidden in the count map.

## 6. Correlation Analysis

Cluster analysis on strongly collinear variables effectively double-counts whatever they
share.

```{r}
#| fig-height: 5
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
               lower = "ellipse",
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

::: {.callout-note}
## Which pair actually has to go

`COMPUTER_PR` and `INTERNET_PR` correlate at **0.87**, far above any other pair
(`MPHONE_PR`–`COMPUTER_PR` is 0.74, `TV_PR`–`MPHONE_PR` is 0.72). In 2014 Myanmar a
home internet connection almost always implied a computer, so the two are close to the
same measurement. I drop `INTERNET_PR` and keep `COMPUTER_PR`, following the handout.

Worth noting what is *not* collinear: `RADIO_PR` is essentially uncorrelated with
everything else (its largest correlation with any other variable is 0.15). Radio is not
a proxy for wealth here, it is its own axis, and it will do real work in the clustering.
:::

## 7. Hierarchical Cluster Analysis

### 7.1 Extracting the clustering variables

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
row.names(cluster_vars) <- cluster_vars$"TS.x"
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

### 7.2 Standardisation

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

```{r}
#| fig-height: 3
r <- ggplot(ict_derived, aes(x = RADIO_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue") +
  ggtitle("Raw")
s <- ggplot(as.data.frame(shan_ict.std), aes(x = RADIO_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue") +
  ggtitle("Min-Max")
z <- ggplot(as.data.frame(shan_ict.z), aes(x = RADIO_PR)) +
  geom_histogram(bins = 20, color = "black", fill = "light blue") +
  ggtitle("Z-score")
ggarrange(r, s, z, ncol = 3, nrow = 1)
```

The *shape* is identical in all three. Standardisation rescales an axis, it does not
reshape a distribution. What it changes is how much each variable counts when distances
between townships are computed, and I come back to that in section 11.

### 7.3 Proximity matrix

```{r}
proxmat <- dist(shan_ict, method = 'euclidean')
```

### 7.4 Choosing the linkage

`agnes()` reports an agglomerative coefficient, where values near 1 indicate a stronger
clustering structure.

```{r}
m <- c("average", "single", "complete", "ward")
names(m) <- c("average", "single", "complete", "ward")
ac <- function(x) { agnes(shan_ict, method = x)$ac }
map_dbl(m, ac)
```

Ward wins clearly at **0.943**, against 0.895 for complete and 0.663 for single linkage.
Single linkage performing worst is expected: it chains, producing one large straggly
cluster and several singletons. Ward is used from here.

### 7.5 How many clusters?

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict, FUN = hcut, nstart = 25, K.max = 10, B = 50)
print(gap_stat, method = "firstmax")
```

```{r}
#| fig-height: 3.5
fviz_gap_stat(gap_stat)
```

::: {.callout-important}
## The gap statistic says one cluster, and I should not pretend otherwise

`firstmax` returns **k = 1**. The honest reading is that the gap statistic finds no
strong evidence of cluster structure in these five variables at all: the gap never rises
above its k=1 value of 0.274, and from k=2 to k=10 it sits in a narrow band between
0.203 and 0.222.

That is not a number I can report as "the answer is 6". What I can say is that *given* a
decision to partition, k = 6 is the local maximum within the usable range and is what
the handout adopts, so I use 6 for comparability. But the segmentation here rests on a
planning decision to have six regions, not on a statistical finding that six exist.
:::

### 7.6 The dendrogram

```{r}
#| fig-height: 5
plot(hclust_ward <- hclust(proxmat, method = 'ward.D'), cex = 0.6)
rect.hclust(hclust_ward, k = 6, border = 2:5)
```

### 7.7 Visually-driven analysis with a cluster heatmap

```{r}
shan_ict_mat <- data.matrix(shan_ict)
heatmaply(normalize(shan_ict_mat),
          Colv = NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA, 200, 60, NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main = "Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State")
```

### 7.8 Mapping the clusters

```{r}
groups <- as.factor(cutree(hclust_ward, k = 6))
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
table(shan_sf_cluster$CLUSTER)
```

```{r}
#| fig-height: 4.5
qtm(shan_sf_cluster, "CLUSTER")
```

The clusters are badly fragmented. Cluster 1 alone appears in pieces scattered across
the whole state. As a *typology* this is fine, but as a set of planning regions it is
unusable, which is exactly the motivation for the next two sections.

## 8. Spatially Constrained Clustering: SKATER

### 8.1 Converting to a spatial object

```{r}
shan_sp <- as_Spatial(shan_sf)
```

### 8.2 Neighbour list

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

264 non-zero links, 4.8 neighbours on average, nothing isolated. A township with no
neighbours would break the minimum spanning tree, so this check matters.

```{r}
#| fig-height: 5
coords <- st_coordinates(st_centroid(st_geometry(shan_sf)))
plot(st_geometry(shan_sf), border = grey(.5))
plot(shan.nb, coords, col = "blue", add = TRUE)
```

### 8.3 Edge costs and the minimum spanning tree

The cost of each edge is the attribute dissimilarity between the two townships it joins.
Note the weights style is `B`, not row-standardised: these are distances, and scaling
them per row would destroy their meaning.

```{r}
lcosts <- nbcosts(shan.nb, shan_ict)
shan.w <- nb2listw(shan.nb, lcosts, style = "B")
summary(shan.w)
```

```{r}
shan.mst <- mstree(shan.w)
dim(shan.mst)
```

54 edges for 55 nodes, as a spanning tree must have.

```{r}
#| fig-height: 5
plot(st_geometry(shan_sf), border = gray(.5))
plot.mst(shan.mst, coords, col = "blue",
         cex.lab = 0.7, cex.circles = 0.005, add = TRUE)
```

### 8.4 Pruning the tree

```{r}
clust6 <- spdep::skater(edges = shan.mst[,1:2],
                        data = shan_ict,
                        method = "euclidean",
                        ncuts = 5)
table(clust6$groups)
```

::: {.callout-important}
## SKATER buys contiguity with badly unbalanced clusters

The six groups are **22, 18, 11, 2, 1, 1**. Two of the six "regions" are single
townships and a third has two. Effectively this is a three-region solution with three
outliers shaved off.

That is a structural property of the method, not bad luck. SKATER can only cut edges of
the spanning tree, so an extreme township that sits on a leaf gets isolated by the
cheapest available cut long before the algorithm will split a large interior block. If
the goal is six comparably sized planning regions, SKATER at `ncuts = 5` does not
deliver them.
:::

```{r}
clust6$ssto
clust6$ssw
```

Total sum of squares is 12,613 and each of the five cuts reduces the within-cluster sum
of squares, ending at 8,735. The cuts are doing real work, just unevenly.

```{r}
#| fig-height: 5
plot(st_geometry(shan_sf), border = gray(.5))
plot(clust6, coords, cex.lab = .7,
     groups.colors = c("red", "green", "blue", "brown", "pink"),
     cex.circles = 0.005, add = TRUE)
```

### 8.5 Comparing unconstrained and constrained

```{r}
#| fig-height: 4.5
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER` = `as.factor.groups_mat.`)

tmap_arrange(qtm(shan_sf_cluster, "CLUSTER") + tm_borders(col_alpha = 0.5),
             qtm(shan_sf_spatialcluster, "SP_CLUSTER") + tm_borders(col_alpha = 0.5),
             asp = NA, ncol = 2)
```

The right-hand map is made of solid contiguous blocks. That is the entire point of
SKATER, and it is visually obvious.

## 9. Spatially Constrained Clustering: ClustGeo

### 9.1 Ward-like clustering without a spatial constraint

```{r}
#| fig-height: 5
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, k = 6, border = 2:5)
```

::: {.callout-important}
## `hclustgeo()` with no constraint is not the same as the `hclust()` above

The handout presents `hclustgeo(proxmat)` as doing "a typical Ward-like hierarchical
clustering just like `hclust()`". It is Ward-like, but it is not the same Ward.

My `hclust()` call in section 7.6 used `method = "ward.D"` and gave cluster sizes
20/4/13/6/7/5. `hclustgeo(proxmat)` gives **16/8/13/6/7/5**. I checked where the
difference comes from, and `hclustgeo(proxmat)` reproduces `hclust(proxmat, "ward.D2")`
*exactly*, township for township, not `ward.D`.

The distinction is that `ward.D2` squares the dissimilarities before applying the Ward
update and `ward.D` does not, so only `ward.D2` implements Ward's actual criterion on a
non-squared input distance matrix. Twelve townships move between clusters 1 and 2 as a
result. It is a small difference here but it means the section 7 dendrogram and the
section 9 dendrogram are not two views of one clustering.
:::

```{r}
groups <- as.factor(cutree(nongeo_cluster, k = 6))
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
#| fig-height: 4.5
qtm(shan_sf_ngeo_cluster, "CLUSTER")
```

### 9.2 Choosing the mixing parameter

`D0` holds attribute dissimilarity, `D1` holds geographic distance, and `alpha` sets how
much weight goes to each.

```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
```

```{r}
#| fig-height: 4
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K = 6, graph = TRUE)
```

```{r}
round(cr$Qnorm, 3)
```

::: {.callout-note}
## Reading the alpha curves

`Q0norm` is the proportion of attribute homogeneity retained, `Q1norm` the geographic
homogeneity. Moving from alpha = 0 to alpha = 0.2 costs about **14%** of attribute
homogeneity (1.000 → 0.855) and buys a jump in geographic homogeneity from 0.139 to
**0.665**. That is a very favourable exchange and it is why 0.2 is the conventional
choice here.

Pushing on to alpha = 0.3 costs a further 3 points of attribute homogeneity for 7 points
of geographic, still positive but clearly past the knee. I follow the handout and use
0.2.
:::

### 9.3 The spatially constrained solution

```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.2)
groups <- as.factor(cutree(clustG, k = 6))
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
table(shan_sf_Gcluster$CLUSTER)
```

```{r}
#| fig-height: 4.5
qtm(shan_sf_Gcluster, "CLUSTER")
```

Much more balanced than SKATER (12/10/9/9/8/7, no singletons) and visibly more spatially
coherent than the unconstrained solution, though not perfectly contiguous. Section 11
measures that.

## 10. Visual Interpretation of Clusters

### 10.1 One variable at a time

```{r}
#| fig-height: 3.5
ggplot(data = shan_sf_ngeo_cluster, aes(x = CLUSTER, y = RADIO_PR)) +
  geom_boxplot()
```

Cluster 3 has the highest median radio penetration and cluster 6 the lowest, with
cluster 5 also low. A single variable only ever tells part of the story though.

### 10.2 Parallel coordinate plot

```{r}
#| fig-height: 4
pc_cols <- match(c("RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR"),
                 names(shan_sf_ngeo_cluster))

ggparcoord(data = shan_sf_ngeo_cluster,
           columns = pc_cols,
           scale = "globalminmax",
           alphaLines = 0.2,
           boxplot = TRUE,
           title = "Parallel coordinates of ICT variables by cluster") +
  facet_grid(~ CLUSTER) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### 10.3 Cluster means

```{r}
shan_sf_ngeo_cluster %>%
  st_set_geometry(NULL) %>%
  group_by(CLUSTER) %>%
  summarise(n = n(),
            mean_RADIO_PR = round(mean(RADIO_PR), 1),
            mean_TV_PR = round(mean(TV_PR), 1),
            mean_LLPHONE_PR = round(mean(LLPHONE_PR), 1),
            mean_MPHONE_PR = round(mean(MPHONE_PR), 1),
            mean_COMPUTER_PR = round(mean(COMPUTER_PR), 1))
```

The six profiles are interpretable:

- **Cluster 4** (n = 6) is the connected one: highest on TV (744), landline (99), mobile
  (651) and computer (65.5) by a wide margin. These are the urban centres.
- **Cluster 5** (n = 7) is the opposite on everything modern (TV 224, mobile 132,
  computer 6.7) but mid-range on radio. Low-penetration remote townships.
- **Cluster 6** (n = 5) is the interesting one: lowest radio of all (98.6) yet high
  mobile (468). These townships appear to have skipped radio rather than lagged on it.
- **Cluster 3** (n = 13) is the high-radio group (300) with solid mobile (392).
- **Clusters 1 and 2** are the middle, separated mainly by mobile and TV.

::: {.callout-note}
## Radio is doing the discriminating

Clusters 5 and 6 have almost the same low-ish profile on landline and computer, yet they
are separated decisively by radio: 124 versus 98.6 on radio but 132 versus 468 on
mobile. Without `RADIO_PR` in the variable set these two groups would likely collapse
together. That is consistent with the correlation matrix in section 6, where radio was
the one variable independent of the rest.
:::

## 11. My Own Exploration: What Does the Spatial Constraint Actually Cost?

Every method here claims a different balance between two things I can measure directly:

- **attribute homogeneity**, as the total within-cluster sum of squares on standardised
  variables (lower is better);
- **spatial coherence**, as the number of contiguous pieces the six clusters break into
  (6 is perfect, higher is worse).

The handout asserts the trade-off exists. I wanted the numbers.

### 11.1 Counting contiguous pieces

```{r}
pieces <- function(lab, nb) {
  total <- 0
  for (k in unique(lab)) {
    idx <- which(lab == k); seen <- integer(0); comp <- 0
    for (i in idx) {
      if (i %in% seen) next
      comp <- comp + 1; stack <- i
      while (length(stack)) {
        v <- stack[1]; stack <- stack[-1]
        if (v %in% seen) next
        seen <- c(seen, v)
        nbrs <- nb[[v]]; nbrs <- nbrs[nbrs != 0]
        stack <- c(stack, intersect(nbrs[!(nbrs %in% seen)], idx))
      }
    }
    total <- total + comp
  }
  total
}

wss <- function(lab) {
  x <- scale(shan_ict)
  sum(sapply(unique(lab), function(k) {
    s <- x[lab == k, , drop = FALSE]
    if (nrow(s) < 2) return(0)
    sum(scale(s, scale = FALSE)^2)
  }))
}
```

### 11.2 The trade-off, measured

```{r}
cg3 <- cutree(hclustgeo(proxmat, distmat, alpha = 0.3), 6)

solutions <- list(
  `hclust (ward.D)`   = cutree(hclust_ward, 6),
  `hclustgeo (a = 0)` = cutree(nongeo_cluster, 6),
  `ClustGeo a = 0.2`  = cutree(clustG, 6),
  `ClustGeo a = 0.3`  = cg3,
  `SKATER`            = clust6$groups
)

tibble(
  method = names(solutions),
  pieces = sapply(solutions, pieces, nb = shan.nb),
  within_SS = round(sapply(solutions, wss), 1),
  sizes = sapply(solutions, function(l)
    paste(sort(as.vector(table(l)), decreasing = TRUE), collapse = "/"))
)
```

::: {.callout-important}
## The constraint is expensive, and ClustGeo does not actually enforce it

Three things came out of this that I did not get from the maps.

**The unconstrained solutions are extremely fragmented.** Six clusters break into 24 and
28 separate patches. That is roughly half of the 55 townships sitting in a patch of
their own or nearly so. "Fragmented" undersells it.

**SKATER's contiguity is absolute and the price is 42%.** It returns exactly 6 pieces
for 6 clusters, as guaranteed by construction. Within-cluster sum of squares rises from
106.6 to 151.1, which is **42% worse attribute homogeneity**. That is the real cost of
mappable regions on this data, and it is not small.

**ClustGeo at alpha = 0.2 does not deliver contiguity.** This is the finding that
surprised me. It still breaks into **18 pieces**, only a third of the way from 24 to 6,
while already giving up 22% of homogeneity. Raising alpha to 0.3 barely helps (17
pieces). ClustGeo *pulls* clusters toward geographic compactness through a distance
penalty, but nothing in the algorithm forbids a cluster from being in two places at
once. SKATER's graph-based formulation is categorically different: contiguity is a
constraint, not a preference.

So the two methods are not interchangeable alternatives. If the deliverable must be
drawable as contiguous regions, only SKATER guarantees it. ClustGeo buys balance (no
singletons, sizes 12/10/9/9/8/7 against SKATER's 22/18/11/2/1/1) but not contiguity.
:::

### 11.3 A second question: the handout standardises, then clusters on raw rates

Sections 7.2 computed both a min-max and a z-score standardisation, and then the
proximity matrix in 7.3 was built from `shan_ict`, the **unstandardised** rates. All five
variables are per-1,000 rates so they share a unit, which is presumably the reasoning.
But sharing a unit is not the same as sharing a scale.

```{r}
v <- apply(shan_ict, 2, var)
tibble(variable = names(v),
       sd = round(sqrt(v), 1),
       pct_of_total_variance = round(100 * v / sum(v), 1))
```

`MPHONE_PR` and `TV_PR` together carry **81%** of the total variance. `COMPUTER_PR`
contributes **0.6%** and `LLPHONE_PR` 2.4%. Euclidean distance on these raw rates is
therefore, to a good approximation, a distance on mobile phones and television only.

```{r}
raw_k6 <- cutree(hclust(dist(shan_ict), "ward.D"), 6)
std_k6 <- cutree(hclust(dist(scale(shan_ict)), "ward.D"), 6)

adj_rand <- function(a, b) {
  t <- table(a, b); n <- sum(t)
  s <- sum(choose(t, 2))
  sa <- sum(choose(rowSums(t), 2)); sb <- sum(choose(colSums(t), 2))
  e <- sa * sb / choose(n, 2)
  (s - e) / ((sa + sb) / 2 - e)
}

tibble(raw_sizes = paste(table(raw_k6), collapse = "/"),
       z_sizes = paste(table(std_k6), collapse = "/"),
       adjusted_rand = round(adj_rand(raw_k6, std_k6), 3))
```

::: {.callout-important}
## Standardising changes the answer substantially

The adjusted Rand index between the raw-rate and z-scored partitions is **0.30**. For
reference, 1 would be identical and 0 would be the agreement expected by chance. 0.30 is
weak agreement: these are largely different segmentations of Shan State, from the same
variables and the same algorithm.

The z-scored version also produces much more even clusters (11/12/9/8/9/6 against
20/4/13/6/7/5), which is what I would expect once `COMPUTER_PR` is no longer
contributing 0.6% of the distance.

I am not claiming the handout is wrong to use raw rates: if the analyst genuinely
believes a 100-per-1,000 difference in mobile ownership matters more than a 100-per-1,000
difference in computer ownership, raw rates encode that belief correctly. The point is
that this is a *decision*, it is unstated, and it moves the result a long way. Reporting
the six clusters without mentioning it would make a choice look like a finding.
:::

## 12. What I Take From This Exercise

**Segmentation is not detection.** Exercise 5 tested a hypothesis and returned a p-value.
This exercise returns a partition, and there is no p-value for a partition. The gap
statistic in section 7.5 pointed at k = 1, which in plain terms means these five
variables do not contain strong natural groupings. Six regions is a planning decision
made on top of a weak statistical signal, and that should be stated out loud rather than
buried.

**Contiguity costs 42% here, and I can now quantify it rather than assert it.** The
measurement in section 11.2 is the thing I will carry forward. Before this exercise I
would have described the spatial constraint as a mild regularisation. On Shan State it
raises within-cluster variance by two-fifths. Whether that is acceptable depends entirely
on whether the output has to be drawn on a map and administered, which is a question
about the client, not about the data.

**ClustGeo and SKATER solve different problems.** This is the correction I most needed.
They are presented side by side as two spatially constrained methods, which made me
assume they were two routes to the same place. They are not: SKATER *enforces*
contiguity and distorts cluster sizes to do it; ClustGeo *encourages* compactness and
preserves balance but left 18 disconnected pieces at the recommended alpha. Picking
between them is picking which guarantee matters.

**Two silent decisions moved the result more than any modelling choice.** Dropping
`INTERNET_PR` for collinearity was explicit and defensible. Clustering on unstandardised
rates was neither, and it changed the partition to an adjusted Rand of 0.30 against the
standardised alternative. The `ward.D` versus `ward.D2` discrepancy in section 9.1 is
smaller but the same species of problem: a default that quietly differs between two
sections of the same analysis.

**Connecting forward.** The next exercise moves to geographically weighted regression,
where a bandwidth plays a role structurally similar to alpha here, deciding how much
geography constrains the estimate. The habit worth keeping from this exercise is to
report what the constraint cost, not just that it was applied.
