---
title: "Network Constrained Spatial Point Patterns Analysis"
subtitle: "HANDS-ON EXERCISE 3"
author:
  - name: Jedidiah Asaf Tallulembang
    affiliation: "ISSS626 — Geospatial Analytics and Applications"
date: last-modified
format:
  html:
    code-fold: false
    toc: true
execute:
  warning: false
  message: false
---

::: {.pill-row}
[R]{.pill} [spNetwork]{.pill} [NKDE]{.pill} [Network K-function]{.pill}
:::

This exercise follows [Network Constrained Spatial Point Patterns
Analysis](https://r4gdsa.netlify.app/chap07), the reading listed under Lesson 3,
Advanced Spatial Point Patterns Analysis.

## 1. Overview

Everything in [Hands-on Exercise 2](../Hands-on_Ex02/Part01.qmd) assumed events could
fall anywhere on a two-dimensional plane, and measured distance as the straight line
between them. For childcare centres that assumption is wrong in a specific way: a centre
sits on a street, a parent walks along streets, and two centres on opposite banks of
Punggol Waterway are 80 metres apart in Euclidean terms and roughly a kilometre apart in
walking terms.

Network constrained SPPA fixes that by confining both the events and the distance metric
to the network itself. This exercise uses `spNetwork` to:

- derive a **network kernel density estimate** (NKDE), and
- perform **network-constrained G- and K-function** analysis.

The study area is Punggol.

## 2. The Data

| Dataset | Description | Format |
|---|---|---|
| `Punggol_St` | Punggol road network | ESRI Shapefile (line) |
| `Punggol_CC` | Childcare centres in Punggol | ESRI Shapefile (point) |

Both were supplied with the course under Lesson 3 and are already in SVY21
(EPSG:3414), so no reprojection is needed — a first for these exercises.

## 3. Getting Started

```{r}
pacman::p_load(sf, spNetwork, tmap, tidyverse)
```

## 4. Data Import and Preparation

```{r}
network <- st_read(dsn = "data/geospatial",
                   layer = "Punggol_St",
                   quiet = TRUE)

childcare <- st_read(dsn = "data/geospatial",
                     layer = "Punggol_CC",
                     quiet = TRUE) %>%
  st_zm(drop = TRUE, what = "ZM")
```

```{r}
network
```

```{r}
childcare
```

The `st_zm()` call on the childcare layer drops the Z (and M) dimension. The shapefile
stores points as XYZ with a zero elevation, and `spNetwork` will refuse to snap 3D points
onto a 2D network, so this is not cosmetic — leaving it out is an error later.

```{r}
st_crs(network)$epsg
st_geometry_type(network) %>% unique()
nrow(network)
nrow(childcare)
```

2,642 road segments and 61 childcare centres.

## 5. Visualising the Geospatial Data

```{r}
plot(st_geometry(network), col = "grey60", lwd = 0.6)
plot(childcare, add = TRUE, col = "#c0392b", pch = 19, cex = 0.8)
```

The reference chapter renders this interactively. I have kept the static version: with
2,645 network segments the self-contained interactive page ran to several megabytes,
which is too heavy to publish for the amount it adds here.

```{r}
#| fig-height: 6
tm_shape(network) +
  tm_lines(col = "grey55", lwd = 0.8) +
  tm_shape(childcare) +
  tm_dots(fill = "#c0392b", size = 0.35) +
  tm_layout(frame = FALSE)
```

## 6. Network KDE Analysis

### 6.1 Preparing the Lixels

A kernel density estimate on a network is evaluated along the network, not over a grid.
The first step is to cut the network into **lixels** — line pixels — of roughly equal
length.

```{r}
lixels <- lixelize_lines(network,
                         700,
                         mindist = 375)
nrow(lixels)
```

The 700m target length with a 375m minimum means any leftover segment shorter than 375m
is merged into the previous lixel rather than left as a stub. The count barely changes
from the original 2,642 segments, which tells me most Punggol road segments were already
shorter than 700m — this is a dense residential network, not a highway corridor.

### 6.2 Generating Line Centre Points

```{r}
samples <- lines_center(lixels)
nrow(samples)
```

These centre points are where the density will actually be evaluated. Note they are the
centre of each lixel by line length, not the geometric centroid, so they always fall
*on* the line even where it bends.

### 6.3 Performing NKDE

```{r}
densities <- nkde(network,
                  events = childcare,
                  w = rep(1, nrow(childcare)),
                  samples = samples,
                  kernel_name = "quartic",
                  bw = 300,
                  div = "bw",
                  method = "simple",
                  digits = 1,
                  tol = 1,
                  grid_shape = c(1, 1),
                  max_depth = 8,
                  agg = 5,
                  sparse = TRUE,
                  verbose = FALSE)
```

A note on the arguments that actually change the answer:

- `kernel_name = "quartic"` — the kernel shape. `spNetwork` also offers triangle,
  gaussian, tricube, cosine, epanechnikov and uniform.
- `bw = 300` — the bandwidth, in metres **along the network**. This is the single most
  consequential choice here.
- `method = "simple"` — the simplest of the three NKDE variants. It treats network
  distance as a straight replacement for Euclidean distance in the kernel. It is fast
  and intuitive, but it is **not** a true density: the values do not integrate to the
  number of events, because mass is duplicated at every intersection. `"discontinuous"`
  and `"continuous"` correct that at increasing computational cost.

Attaching the computed densities and rescaling from events-per-metre to
events-per-kilometre so the legend is readable:

```{r}
samples$density <- densities
lixels$density  <- densities

samples$density <- samples$density * 1000
lixels$density  <- lixels$density * 1000
```

```{r}
summary(lixels$density)
```

### 6.4 Visualising the NKDE

```{r}
#| fig-height: 6
tm_shape(lixels) +
  tm_lines(col = "density",
           col.scale = tm_scale_intervals(style = "quantile", values = "viridis"),
           col.legend = tm_legend(title = "Density (events/km)"),
           lwd = 1.5) +
  tm_layout(frame = FALSE)
```

::: {.callout-note}
## What the NKDE map shows that a planar KDE would not
The high-density road segments are not a blob in the middle of Punggol. They are
specific stretches — the roads threading the HDB precincts either side of Punggol
Field and Punggol Central — while the roads along the waterway and the LRT viaduct
run cold despite being geometrically close to plenty of centres. A planar KDE would
smear density straight across the waterway and report a hotspot in the middle of the
water, which is exactly the artefact this method exists to avoid.
:::

## 7. Network Constrained G- and K-Function Analysis

The same complete spatial randomness question as Hands-on Exercise 2 Part 02, but with
randomness defined **on the network**: the simulated points are scattered along the road
network rather than across the plane.

$H_0$: the observed spatial point events are uniformly distributed over the street
network in Punggol.
$H_1$: the observed spatial point events are **not** uniformly distributed over the
street network.

The test is at the 95% confidence interval, with the null rejected if the observed curve
leaves the simulated envelope.

```{r}
set.seed(1234)
kfun_childcare <- kfunctions(network,
                             childcare,
                             start = 0,
                             end = 1000,
                             step = 50,
                             width = 50,
                             nsim = 50,
                             resolution = 50,
                             verbose = FALSE,
                             conf_int = 0.05)
```

### 7.1 The K-Function Result

```{r}
kfun_childcare$plotk
```

### 7.2 The G-Function Result

```{r}
kfun_childcare$plotg
```

Reading the direction off the plots by eye is unreliable, so I compute where the observed
curve actually leaves the simulated envelope:

```{r}
v <- kfun_childcare$values

rng <- function(d) {
  if (length(d) == 0) {
    "never leaves envelope"
  } else {
    paste0(min(d), "-", max(d), " m")
  }
}

data.frame(
  Function = c("Network K", "Network K", "Network G", "Network G"),
  Side = c("above upper (clustered)", "below lower (regular)",
           "above upper (clustered)", "below lower (regular)"),
  Range = c(rng(v$distances[v$obs_k > v$upper_k & v$distances > 0]),
            rng(v$distances[v$obs_k < v$lower_k]),
            rng(v$distances[v$obs_g > v$upper_g & v$distances > 0]),
            rng(v$distances[v$obs_g < v$lower_g]))
)
```

::: {.callout-important}
## Conclusion of the test — and it is not what I expected
The observed curves leave the envelope on the **lower** side, not the upper one. Fewer
centres fall within a few hundred network metres of each other than complete spatial
randomness along the network would produce.

I reject the null hypothesis — the centres are not uniformly distributed along the
network — but in the **opposite direction to clustering**. This is *regularity*
(dispersion, inhibition). At longer distances the observed curve returns inside the
envelope, so beyond that range the pattern is indistinguishable from random.

The seed is fixed above so these ranges are reproducible; without it the exact crossing
distances shift on every rebuild.
:::

## 8. What I Take From This Exercise

**The result contradicts Exercise 2, and the contradiction is the whole lesson.**
Exercise 2 Part 02 concluded that childcare centres cluster at short range. Measured
along the road network, the same kind of test on Punggol says the opposite: over the
mid-range distances in the table above, centres are *further apart* than chance.

I do not think either result is wrong. They are answering different questions:

- The planar K-function asks "are centres closer together than random points scattered
  over this **polygon**?" A large share of that polygon is water, park and reserve where
  nothing can be built, so the random comparison points are spread over space that real
  centres can never occupy. Against that null, anything confined to buildable land looks
  clustered.
- The network K-function asks "are centres closer together than random points scattered
  over this **road network**?" That null already excludes the unbuildable space. What is
  left is the actual siting decision — and the actual siting decision turns out to space
  centres out, roughly one per residential precinct, which is exactly what a planner
  allocating childcare places by catchment would do.

So the planar "clustering" in Exercise 2 was substantially an artefact of the null, not
a finding about childcare policy. That is an uncomfortable thing to discover about work
I had already written up, and it is the strongest argument I have seen for why the
network-constrained method exists.

**Where the method is fragile.** Three things would change the answer and none of them
are statistical:

1. **Bandwidth.** 300m is a defensible walking distance, but it is a choice, and the
   hotspots move if I halve or double it. There is no `bw.diggle` equivalent here that
   picks it for me.
2. **Network completeness.** `Punggol_St` is a road network. Parents in Punggol walk
   through void decks, park connectors and the waterway bridges, none of which are in
   this layer. The network I am measuring distance on is not the network people use.
3. **`method = "simple"`.** It duplicates kernel mass at intersections, and Punggol is
   full of intersections. The continuous method exists precisely to fix this and I would
   want to re-run with it before quoting any of these density values as real.

**On `nsim = 50`.** The chapter uses 50 simulations and I have kept that, but 50 is thin
for a 95% envelope — it puts the pointwise significance level at 2/51, close enough to
0.04 that a marginal result would not survive scrutiny. Here the observed curve sits
clearly outside the envelope across a contiguous band of distances rather than grazing it
at a single point, so the conclusion holds. A single-band excursion would have been a
reason to raise `nsim` rather than report it.

**Connecting back.** Reading Exercise 2 and this one together is the actual lesson.
First-order KDE says where events are dense. Second-order G/F/K/L say whether events
interact. Network-constrained versions say whether either conclusion survives once you
stop pretending people travel in straight lines and stop pretending reservoirs are
candidate sites. Each layer removed an assumption the previous one was quietly making —
and unlike the first two steps, this one did not confirm the earlier answer. It reversed
it. The thing I will carry forward is that in spatial statistics the null hypothesis is
a modelling choice as consequential as the estimator, and a result is only as meaningful
as the randomness it was compared against.
