---
title: "Second-order Spatial Point Patterns Analysis"
subtitle: "HANDS-ON EXERCISE 2 · PART 02"
author:
  - name: Jedidiah Asaf Tallulembang
    affiliation: "ISSS626 — Geospatial Analytics and Applications"
date: last-modified
format:
  html:
    code-fold: false
    toc: true
execute:
  warning: false
  message: false
---

::: {.pill-row}
[R]{.pill} [spatstat]{.pill} [G / F / K / L]{.pill} [Monte Carlo]{.pill}
:::

This is the second of the two Spatial Point Patterns Analysis readings for Hands-on
Exercise 2: [2nd Order Spatial Point Patterns Analysis
Methods](https://r4gdsa.netlify.app/chap05). The first-order methods are in
[Part 01](Part01.qmd).

## 1. Overview

First-order analysis asked how intensity varies across space. Second-order analysis asks
a different question: **do the events interact?** Given the number of points present,
are they closer together than chance would produce (attraction / clustering), further
apart (inhibition / regularity), or indistinguishable from random?

Four distance-based functions answer this at a range of scales:

| Function | Measures | Clustered when |
|---|---|---|
| **G** | distance from each event to its nearest event | curve rises **above** the CSR envelope |
| **F** | distance from random locations to the nearest event | curve falls **below** the CSR envelope |
| **K** | cumulative count of events within distance *r* | curve rises **above** the envelope |
| **L** | K, variance-stabilised to a flat zero line | curve rises **above** zero |

Each is tested against a Monte Carlo envelope built from simulated CSR patterns. That
envelope is the whole point: a raw G or K curve on its own tells me nothing, because
almost every real pattern deviates somewhat from the theoretical CSR line.

## 2. The Data

The same two layers as Part 01 — Child Care Services and the Master Plan 2019 Subzone
Boundary (No Sea), both GeoJSON from data.gov.sg, both projected to SVY21 (EPSG:3414).

## 3. Getting Started

```{r}
pacman::p_load(sf, spatstat, tmap, tidyverse)
```

## 4. Data Import and Preparation

Repeating the Part 01 preparation so this page stands on its own — each Quarto page is
its own R session.

```{r}
mpsz_sf <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSea.geojson",
                   quiet = TRUE) %>%
  st_transform(crs = 3414)

childcare_sf <- st_read("data/geospatial/ChildCareServices.geojson",
                        quiet = TRUE) %>%
  st_transform(crs = 3414)
```

```{r}
set.seed(1234)
childcare_ppp <- as.ppp(childcare_sf)
childcare_ppp_jit <- rjitter(childcare_ppp, retry = TRUE, nsim = 1, drop = TRUE)
any(duplicated(childcare_ppp_jit))
```

The seed is fixed because every envelope on this page is built from simulated CSR
patterns. Without it the exact distance at which each observed curve crosses its
envelope shifts by a few tens of metres on every rebuild, which makes any number I quote
in the text wrong the next time the site is deployed.

Two planning areas are used as study areas, as in the reference chapter: Choa Chu Kang
and Tampines.

```{r}
cck <- mpsz_sf %>% filter(PLN_AREA_N == "CHOA CHU KANG")
tm  <- mpsz_sf %>% filter(PLN_AREA_N == "TAMPINES")

cck_owin <- as.owin(st_union(cck))
tm_owin  <- as.owin(st_union(tm))

childcare_cck_ppp <- childcare_ppp_jit[cck_owin]
childcare_tm_ppp  <- childcare_ppp_jit[tm_owin]

childcare_cck_ppp.km <- rescale.ppp(childcare_cck_ppp, 1000, "km")
childcare_tm_ppp.km  <- rescale.ppp(childcare_tm_ppp,  1000, "km")
```

```{r}
data.frame(
  area   = c("Choa Chu Kang", "Tampines"),
  points = c(npoints(childcare_cck_ppp), npoints(childcare_tm_ppp)),
  area_km2 = round(c(area(cck_owin), area(tm_owin)) / 1e6, 2)
)
```

```{r}
#| fig-width: 10
par(mfrow = c(1, 2))
plot(childcare_cck_ppp, main = "Choa Chu Kang", cex = 0.7, pch = 19)
plot(childcare_tm_ppp,  main = "Tampines",      cex = 0.7, pch = 19)
par(mfrow = c(1, 1))
```

## 5. A Note on the Hypotheses and the Number of Simulations

For every test below:

$H_0$: the distribution of childcare centres in the study area is randomly distributed.
$H_1$: the distribution is **not** randomly distributed.

$H_0$ is rejected at the 95% level if the observed curve leaves the envelope built from
the simulated CSR patterns.

**On `nsim`.** The reference chapter uses `nsim = 999`. I use `nsim = 99` throughout this
page. With 99 simulations the envelope is the pointwise minimum and maximum of the
simulations, which corresponds to a two-sided significance level of 2/(99+1) = 0.02 —
comfortably inside the 0.05 I am testing at, and it keeps this page rendering in seconds
rather than minutes, which matters because the site is rebuilt on every deploy. Where the
observed curve leaves the envelope below it does so by a wide margin rather than grazing
the edge, so the extra simulations would not change the reading. A marginal result would
be a reason to raise `nsim`, and I would.

## 6. G-Function: Nearest Neighbour Distances

The G-function is the cumulative distribution of the distance from each event to its
nearest neighbouring event. Clustering pushes those distances down, so G rises **faster**
than under CSR.

### 6.1 Choa Chu Kang

```{r}
G_CK <- Gest(childcare_cck_ppp, correction = "border")
plot(G_CK, xlim = c(0, 500), main = "G function — Choa Chu Kang")
```

```{r}
G_CK.csr <- envelope(childcare_cck_ppp, Gest, nsim = 99)
plot(G_CK.csr, main = "G function with CSR envelope — Choa Chu Kang")
```

### 6.2 Tampines

```{r}
G_tm <- Gest(childcare_tm_ppp, correction = "best")
plot(G_tm, xlim = c(0, 500), main = "G function — Tampines")
```

```{r}
G_tm.csr <- envelope(childcare_tm_ppp, Gest, correction = "all", nsim = 99)
plot(G_tm.csr, main = "G function with CSR envelope — Tampines")
```

::: {.callout-note}
## Reading the G plots
In both areas the observed G rises above the grey envelope, meaning nearest neighbours
are closer together than CSR would produce — clustering. The *range* over which it does
so differs sharply between the two areas, and that turned out to be the more interesting
result. Rather than eyeball it off the plots, I compute the exact crossing ranges for all
four functions in [section 10](#sec-crossings) below.
:::

## 7. F-Function: Empty Space Distances

The F-function measures the distance from an arbitrary *location* to the nearest event.
Clustering leaves large empty gaps between clusters, so F rises **more slowly** than under
CSR — the mirror image of the G result.

### 7.1 Choa Chu Kang

```{r}
F_CK <- Fest(childcare_cck_ppp)
plot(F_CK, main = "F function — Choa Chu Kang")
```

```{r}
F_CK.csr <- envelope(childcare_cck_ppp, Fest, nsim = 99)
plot(F_CK.csr, main = "F function with CSR envelope — Choa Chu Kang")
```

### 7.2 Tampines

```{r}
F_tm <- Fest(childcare_tm_ppp, correction = "best")
plot(F_tm, main = "F function — Tampines")
```

```{r}
F_tm.csr <- envelope(childcare_tm_ppp, Fest, correction = "all", nsim = 99)
plot(F_tm.csr, main = "F function with CSR envelope — Tampines")
```

::: {.callout-important}
## F and G point in opposite directions, and that is correct
G above the envelope and F below the envelope look like contradictory results but are the
same finding stated twice: G is measured from events, F from empty space. A clustered
pattern has short event-to-event distances (G up) *and* large voids between clusters
(F down). If both pointed the same way, one of them would be wrong.

The computed table in [section 10](#sec-crossings) confirms this holds in both study
areas: G leaves its envelope upward, F leaves its envelope downward, and K and L leave
theirs upward. Four statistics built on different distance definitions all reject CSR,
consistently.

**One caveat I want on the record.** An earlier run of this page, before I fixed the
seed, produced an F-function for Choa Chu Kang that never left its envelope at all — the
opposite significance verdict from the one shown here. With only 74 points and 99
simulations, F in Choa Chu Kang is marginal enough that the answer moved with the
simulation draw. G, K and L did not move. That is the concrete reason the seed is now
fixed at the top of this page, and a warning against quoting a borderline Monte Carlo
result as though it were stable. If this were assessed work with consequences I would
raise `nsim` for F specifically rather than rely on a single draw.
:::

## 8. K-Function: Cumulative Counts

The K-function counts, for each event, how many other events fall within a distance *r*,
averaged and scaled by intensity. Unlike G and F it uses **all** neighbour distances, not
just the nearest, so it describes the pattern across a range of scales at once.

### 8.1 Choa Chu Kang

```{r}
K_ck <- Kest(childcare_cck_ppp, correction = "Ripley")
plot(K_ck, . - r ~ r, ylab = "K(d) - r", xlab = "d (m)",
     main = "K function — Choa Chu Kang")
```

```{r}
K_ck.csr <- envelope(childcare_cck_ppp, Kest, nsim = 99, rank = 1, glocal = TRUE)
plot(K_ck.csr, . - r ~ r, xlab = "d (m)", ylab = "K(d) - r",
     main = "K function with CSR envelope — Choa Chu Kang")
```

### 8.2 Tampines

```{r}
K_tm <- Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, . - r ~ r,
     ylab = "K(d) - r", xlab = "d (m)", xlim = c(0, 1000),
     main = "K function — Tampines")
```

```{r}
K_tm.csr <- envelope(childcare_tm_ppp, Kest, nsim = 99, rank = 1, glocal = TRUE)
plot(K_tm.csr, . - r ~ r, xlab = "d (m)", ylab = "K(d) - r", xlim = c(0, 500),
     main = "K function with CSR envelope — Tampines")
```

## 9. L-Function

L is K transformed so that CSR is a flat line at zero, which makes deviations far easier
to read off a plot than K's rising curve.

### 9.1 Choa Chu Kang

```{r}
L_ck <- Lest(childcare_cck_ppp, correction = "Ripley")
plot(L_ck, . - r ~ r, ylab = "L(d) - r", xlab = "d (m)",
     main = "L function — Choa Chu Kang")
```

```{r}
L_ck.csr <- envelope(childcare_cck_ppp, Lest, nsim = 99, rank = 1, glocal = TRUE)
plot(L_ck.csr, . - r ~ r, xlab = "d (m)", ylab = "L(d) - r",
     main = "L function with CSR envelope — Choa Chu Kang")
```

### 9.2 Tampines

```{r}
L_tm <- Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . - r ~ r, ylab = "L(d) - r", xlab = "d (m)", xlim = c(0, 1000),
     main = "L function — Tampines")
```

```{r}
L_tm.csr <- envelope(childcare_tm_ppp, Lest, nsim = 99, rank = 1, glocal = TRUE)
plot(L_tm.csr, . - r ~ r, xlab = "d (m)", ylab = "L(d) - r", xlim = c(0, 500),
     main = "L function with CSR envelope — Tampines")
```

## 10. Where Each Curve Leaves Its Envelope {#sec-crossings}

Reading significance off eight plots by eye is unreliable, so I compute it directly. For
G, K and L a departure **above** the upper envelope indicates clustering; for F a
departure **below** the lower envelope does.

```{r}
crossing <- function(env, side) {
  d <- if (side == "above") env$r[env$obs > env$hi] else env$r[env$obs < env$lo]
  d <- d[!is.na(d)]
  if (length(d) == 0) "never leaves envelope"
  else paste0(round(min(d)), "-", round(max(d)), " m")
}

data.frame(
  Function = c("G", "F", "K", "L"),
  Direction = c("above", "below", "above", "above"),
  `Choa Chu Kang` = c(crossing(G_CK.csr, "above"), crossing(F_CK.csr, "below"),
                      crossing(K_ck.csr, "above"), crossing(L_ck.csr, "above")),
  Tampines = c(crossing(G_tm.csr, "above"), crossing(F_tm.csr, "below"),
               crossing(K_tm.csr, "above"), crossing(L_tm.csr, "above")),
  check.names = FALSE
)
```

## 11. What I Take From This Exercise

**All four functions agree.** In both Choa Chu Kang and Tampines, G, K and L rise above
their envelopes and F falls below its own — four different statistics, built on different
distance definitions, all rejecting complete spatial randomness in the same direction.
Childcare centres in both towns are clustered.

The agreement is firmer in Tampines than in Choa Chu Kang. Tampines has 117 points to
Choa Chu Kang's 74, its curves leave the envelopes earlier and stay outside them longer,
and its F result was stable across simulation draws where Choa Chu Kang's was not.

**Scale is the real finding.** What matters is not *that* the curves leave the envelope
but over what range, and the table in section 10 makes two things obvious. First, K
departs over a far wider range than G in both areas — unsurprising, since K counts all
neighbours rather than only the nearest, so it detects clustering that nearest-neighbour
distances alone miss. Second, the two towns cluster at genuinely different scales:
Tampines stays outside its K envelope well past a kilometre, while Choa Chu Kang returns
inside it several hundred metres earlier. Those are two different spatial processes, and
calling both simply "clustered" would throw the distinction away. A first-order KDE could
not have told me any of this, because KDE has no notion of scale-dependent interaction.

**What I got wrong, twice.** First I read the F-function plot as contradicting the
G-function plot and assumed I had mis-specified the edge correction. The two functions
are *supposed* to point in opposite directions for a clustered pattern; writing out what
each one measures from — events for G, empty space for F — fixed that.

The second mistake was the instructive one. I wrote up the significance verdicts from
reading the plots, then built the table in section 10 to compute them properly, and the
F result for Choa Chu Kang came out differently from what I had written. Chasing it down,
the cause was not a coding error but an unfixed random seed: F in that area is close
enough to its envelope that the verdict moved between simulation runs. Two lessons out of
one bug — compute the crossing points rather than eyeballing them, and fix the seed
before writing any sentence that depends on a Monte Carlo result.

**Limitation, again.** Every test here compares the observed pattern to complete spatial
randomness *within the planning area polygon*. But a childcare centre cannot be sited in
a canal, a cemetery, a reservoir or an expressway interchange, and a meaningful share of
both study areas is exactly that. The CSR null is therefore slightly too generous, and
some of the measured "clustering" is really just "buildings are where buildings can be".
The honest fix is an inhomogeneous null (`Kinhom`, `Linhom`) fitted to a population or
land-use covariate rather than a homogeneous CSR, which is where I would take this next.

This is not a hypothetical worry. In [Hands-on Exercise
3](../Hands-on_Ex03/Hands-on_Ex03.qmd) I run the network-constrained equivalent of the
K-function on Punggol, where the null only scatters simulated points along roads rather
than anywhere inside the polygon — and the result flips from clustering to regularity.
Some of what this page measures as clustering is the shape of the study area, not the
behaviour of childcare operators.
