---
title: "First-order Spatial Point Patterns Analysis"
subtitle: "HANDS-ON EXERCISE 2 · PART 01"
author:
  - name: Jedidiah Asaf Tallulembang
    affiliation: "ISSS626 — Geospatial Analytics and Applications"
date: last-modified
format:
  html:
    code-fold: false
    toc: true
execute:
  warning: false
  message: false
---

::: {.pill-row}
[R]{.pill} [sf]{.pill} [spatstat]{.pill} [KDE]{.pill} [Point Patterns]{.pill}
:::

This is the first of the two Spatial Point Patterns Analysis readings for Hands-on
Exercise 2: [First-order Spatial Point Patterns Analysis
Methods](https://r4gdsa.netlify.app/chap04). The second-order methods are written up
separately in [Part 02](Part02.qmd).

## 1. Overview

First-order SPPA is about **intensity** — how many events fall per unit area, and how
that density varies across space. It answers "where are childcare centres concentrated?"
without yet asking "do centres attract or repel each other?" That second question is
what Part 02 handles.

The two questions I am trying to answer here:

- Are childcare centres in Singapore randomly distributed across the country?
- If not, where are the areas with a higher concentration of them?

## 2. The Data

| Dataset | Source | Format used |
|---|---|---|
| Child Care Services | [data.gov.sg](https://data.gov.sg/datasets/d_5d668e3f544335f8028f546827b773b4/view) | GeoJSON |
| Master Plan 2019 Subzone Boundary (No Sea) | [data.gov.sg](https://data.gov.sg/datasets/d_8594ae9ff96d0c708bc2af633048edfb/view) | GeoJSON |

**Note on the data.** The reference chapter offers both KML and GeoJSON and uses
`rvest` to scrape the attribute table out of the KML `Description` field. I downloaded
the GeoJSON versions, which already expose `SUBZONE_N`, `PLN_AREA_N` and the childcare
attributes as proper columns, so no scraping step is needed and `rvest` is not loaded.

Note also that this exercise uses the **2019** Master Plan boundary, not the 2014
boundary I used in Hands-on Exercise 1. The subzone names and codes differ slightly
between the two vintages, so the two exercises are not directly comparable at subzone
level.

## 3. Getting Started

```{r}
pacman::p_load(sf, spatstat, terra, tmap, tidyverse)
```

## 4. Importing and Wrangling the Data

Both layers arrive in WGS84 and are projected to SVY21 (EPSG:3414) on import, because
every density and distance calculation that follows is measured in metres.

```{r}
mpsz_sf <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSea.geojson",
                   quiet = TRUE) %>%
  st_transform(crs = 3414)

childcare_sf <- st_read("data/geospatial/ChildCareServices.geojson",
                        quiet = TRUE) %>%
  st_transform(crs = 3414)
```

```{r}
glimpse(childcare_sf)
```

```{r}
st_crs(mpsz_sf)$epsg
nrow(mpsz_sf)
nrow(childcare_sf)
```

332 subzones and 1,925 childcare centres.

### 4.1 Mapping the Data

```{r}
tmap_mode("plot")
tm_shape(mpsz_sf) +
  tm_polygons(fill = "grey95", col = "white", lwd = 0.4) +
  tm_shape(childcare_sf) +
  tm_dots(fill = "#c0392b", size = 0.02) +
  tm_title("Childcare centres in Singapore (2026)")
```

The reference chapter also renders this as an interactive leaflet map. I built that
version and then removed it: a self-contained HTML page with 1,925 interactive markers
came out at over 7 MB and took pandoc the better part of an hour to assemble, which is
not a reasonable thing to publish on a coursework site. A faceted static map carries the
same information at a fraction of the weight.

```{r}
#| fig-height: 7
tm_shape(mpsz_sf) +
  tm_polygons(fill = "grey95", col = "white", lwd = 0.3) +
  tm_shape(childcare_sf) +
  tm_dots(fill = "#c0392b", size = 0.04) +
  tm_layout(frame = FALSE)
```

::: {.callout-note}
## What I notice first
Even before any statistics, the map is obviously not uniform. The dots trace the HDB
new towns — Jurong West, Woodlands, Sengkang, Punggol, Tampines — and thin out sharply
over the central catchment, the western industrial estates and the military land. That
already suggests the answer to question one is "no", but eyeballing a map is not a test,
which is exactly why the Clark-Evans statistic below matters.
:::

## 5. Geospatial Data Wrangling

`spatstat` does not work on `sf` objects. It needs a `ppp` (point pattern) object for
the events and an `owin` (observation window) object for the study area boundary.

### 5.1 Converting to `ppp` Format

```{r}
childcare_ppp <- as.ppp(childcare_sf)
childcare_ppp
```

```{r}
summary(childcare_ppp)
```

The summary warns about duplicated points. Checking directly:

```{r}
any(duplicated(childcare_ppp))
sum(multiplicity(childcare_ppp) > 1)
```

Duplicates matter because spatial point pattern theory assumes points are distinct —
two centres at identical coordinates break the nearest-neighbour distance calculations
(a distance of zero). The standard fix is jittering, which shifts coincident points by a
tiny random amount:

```{r}
childcare_ppp_jit <- rjitter(childcare_ppp,
                             retry = TRUE,
                             nsim = 1,
                             drop = TRUE)
any(duplicated(childcare_ppp_jit))
```

::: {.callout-note}
## Why duplicates appear here
These are almost certainly real: several childcare centres share one building — a
void deck with two operators, or a mall with several centres on different floors. The
coordinates are building centroids, so the operators collapse onto one point. Jittering
by a few metres is honest in this case because the true locations really are metres
apart, not kilometres.
:::

### 5.2 Creating the `owin` Object

```{r}
sg_sf <- mpsz_sf %>%
  st_union()

sg_owin <- as.owin(sg_sf)
plot(sg_owin, main = "Singapore boundary as an owin object")
```

### 5.3 Combining Points and Window

```{r}
childcareSG_ppp <- childcare_ppp_jit[sg_owin]
childcareSG_ppp
```

```{r}
plot(childcareSG_ppp, main = "Childcare centres within the Singapore window",
     cex = 0.3, pch = 19)
```

## 6. Clark-Evans Test for Nearest Neighbour Analysis

The Clark-Evans R statistic is the ratio of observed mean nearest-neighbour distance to
the distance expected under complete spatial randomness (CSR).

- R = 1 → random
- R < 1 → clustered
- R > 1 → regular / dispersed

$H_0$: the distribution of childcare centres is randomly distributed.
$H_1$: the distribution is **not** randomly distributed.
Confidence level 95%, so I reject $H_0$ if p < 0.05.

### 6.1 Without CSR Simulation

```{r}
clarkevans.test(childcareSG_ppp,
                correction = "none",
                clipregion = "sg_owin",
                alternative = c("clustered"),
                nsim = 99)
```

### 6.2 With CSR Simulation

**On the number of simulations.** The reference chapter uses `nsim = 999`. Each
simulation has to scatter 1,925 random points inside Singapore's full multipolygon
boundary, which is slow enough at 999 that it dominates the build time for this whole
site. I use `nsim = 99`, which gives a pointwise two-sided level of 2/100 = 0.02, still
well inside the 0.05 I am testing at. The observed R sits far outside the simulated
range rather than near its edge, so the conclusion does not depend on the extra
simulations.

```{r}
clarkevans.test(childcareSG_ppp,
                correction = "none",
                clipregion = NULL,
                alternative = c("clustered"),
                method = "MonteCarlo",
                nsim = 99)
```

::: {.callout-important}
## Conclusion of the test
R is well below 1 and p is far below 0.05 under both the analytical Z-test and the
Monte Carlo version, so I reject the null hypothesis. Childcare centres
in Singapore are **clustered**, not randomly distributed. The Monte Carlo version is the
one I would quote, because it does not depend on the asymptotic normality assumption
that the plain Z-test makes.
:::

## 7. Kernel Density Estimation

Clark-Evans tells me *that* the pattern is clustered. KDE tells me *where*.

### 7.1 Automatic Bandwidth Selection

```{r}
kde_childcareSG_bw <- density(childcareSG_ppp,
                              sigma = bw.diggle,
                              edge = TRUE,
                              kernel = "gaussian")
plot(kde_childcareSG_bw, main = "KDE, bw.diggle bandwidth")
```

The bandwidth actually chosen:

```{r}
bw <- bw.diggle(childcareSG_ppp)
bw
```

### 7.2 Rescaling to Kilometres

The density values above are unreadable because the unit is metres — density comes out
as "events per square metre", which is a very small number. Rescaling the pattern to
kilometres fixes the legend without changing the analysis.

```{r}
childcareSG_ppp.km <- rescale.ppp(childcareSG_ppp, 1000, "km")

kde_childcareSG.bw <- density(childcareSG_ppp.km,
                              sigma = bw.diggle,
                              edge = TRUE,
                              kernel = "gaussian")
plot(kde_childcareSG.bw, main = "KDE rescaled to km")
```

### 7.3 Comparing Bandwidth Methods

```{r}
bw.CvL(childcareSG_ppp.km)
bw.scott(childcareSG_ppp.km)
bw.ppl(childcareSG_ppp.km)
bw.diggle(childcareSG_ppp.km)
```

```{r}
#| fig-height: 8
par(mfrow = c(2, 2))
plot(density(childcareSG_ppp.km, sigma = bw.CvL,    edge = TRUE, kernel = "gaussian"), main = "bw.CvL")
plot(density(childcareSG_ppp.km, sigma = bw.scott,  edge = TRUE, kernel = "gaussian"), main = "bw.scott")
plot(density(childcareSG_ppp.km, sigma = bw.ppl,    edge = TRUE, kernel = "gaussian"), main = "bw.ppl")
plot(density(childcareSG_ppp.km, sigma = bw.diggle, edge = TRUE, kernel = "gaussian"), main = "bw.diggle")
par(mfrow = c(1, 1))
```

::: {.callout-note}
## Which bandwidth I would use
`bw.CvL` returns a much larger bandwidth and produces a smooth blob that says little
beyond "the middle and east are busier". `bw.diggle` and `bw.ppl` are far tighter and
resolve individual towns. Baddeley's advice is that `bw.ppl` is preferable when the
pattern is a collection of tight clusters, and `bw.diggle` when the goal is detecting a
single dominant cluster. Childcare centres are clearly the former — many town-sized
clusters — so `bw.ppl` is the more defensible choice here, even though the chapter
demonstrates `bw.diggle` first.
:::

### 7.4 Comparing Kernel Functions

```{r}
#| fig-height: 8
par(mfrow = c(2, 2))
plot(density(childcareSG_ppp.km, sigma = bw.ppl, edge = TRUE, kernel = "gaussian"),     main = "Gaussian")
plot(density(childcareSG_ppp.km, sigma = bw.ppl, edge = TRUE, kernel = "epanechnikov"), main = "Epanechnikov")
plot(density(childcareSG_ppp.km, sigma = bw.ppl, edge = TRUE, kernel = "quartic"),      main = "Quartic")
plot(density(childcareSG_ppp.km, sigma = bw.ppl, edge = TRUE, kernel = "disc"),         main = "Disc")
par(mfrow = c(1, 1))
```

The four kernels are visually near-identical. That is the expected result — the choice
of bandwidth dominates the choice of kernel, which is a useful thing to have seen rather
than just read.

## 8. Fixed and Adaptive Bandwidth KDE

### 8.1 Fixed Bandwidth

A fixed 600m bandwidth (0.6 in km units):

```{r}
kde_childcareSG_600 <- density(childcareSG_ppp.km,
                               sigma = 0.6,
                               edge = TRUE,
                               kernel = "gaussian")
plot(kde_childcareSG_600, main = "Fixed bandwidth, sigma = 600m")
```

### 8.2 Adaptive Bandwidth

Fixed bandwidth is a problem when the underlying point density varies a lot across the
study area — which is exactly the case in Singapore, where the centre is dense and the
west is sparse. Adaptive bandwidth varies the smoothing locally.

```{r}
kde_childcareSG_adaptive <- adaptive.density(childcareSG_ppp.km, method = "kernel")
plot(kde_childcareSG_adaptive, main = "Adaptive bandwidth KDE")
```

```{r}
#| fig-width: 10
par(mfrow = c(1, 2))
plot(kde_childcareSG.bw, main = "Fixed bandwidth")
plot(kde_childcareSG_adaptive, main = "Adaptive bandwidth")
par(mfrow = c(1, 1))
```

::: {.callout-note}
## Fixed vs adaptive
The adaptive surface keeps detail in the sparse western and northern areas that the
fixed-bandwidth version smooths into near-nothing, while holding the dense central
corridor together instead of fragmenting it. The trade-off is that the adaptive map is
harder to explain to a non-technical reader, because the amount of smoothing is no
longer constant across the map — two equally bright pixels do not necessarily mean the
same thing.
:::

## 9. Cartographic Quality KDE Map

To put the KDE surface on a proper map, the `spatstat` image is converted to a raster,
given the SVY21 projection, and plotted with `tmap`.

```{r}
kde_childcareSG_bw_raster <- rast(kde_childcareSG_bw)
kde_childcareSG_bw_raster
```

```{r}
crs(kde_childcareSG_bw_raster) <- "EPSG:3414"
kde_childcareSG_bw_raster
```

```{r}
tmap_mode("plot")
tm_shape(kde_childcareSG_bw_raster) +
  tm_raster(col.scale = tm_scale_continuous(values = "viridis"),
            col.legend = tm_legend(title = "Density")) +
  tm_layout(frame = FALSE)
```

**Note on `tmap`.** This site renders with `tmap` 4.x, where `tm_raster()` takes
`col.scale`/`col.legend` arguments rather than the `palette =` argument used in the
version-3 reference code. The output is the same; only the API has moved on.

## 10. First-order SPPA at the Planning Area Level

A national KDE is dominated by the biggest clusters. Restricting the window to a single
planning area gives a much fairer read of the pattern within it.

### 10.1 Extracting the Study Areas

```{r}
cck <- mpsz_sf %>% filter(PLN_AREA_N == "CHOA CHU KANG")
tm  <- mpsz_sf %>% filter(PLN_AREA_N == "TAMPINES")
jw  <- mpsz_sf %>% filter(PLN_AREA_N == "JURONG WEST")
pg  <- mpsz_sf %>% filter(PLN_AREA_N == "PUNGGOL")
```

```{r}
#| fig-height: 7
par(mfrow = c(2, 2))
plot(st_geometry(cck), main = "Choa Chu Kang")
plot(st_geometry(tm),  main = "Tampines")
plot(st_geometry(jw),  main = "Jurong West")
plot(st_geometry(pg),  main = "Punggol")
par(mfrow = c(1, 1))
```

### 10.2 Creating `owin` Objects and Point Patterns

```{r}
cck_owin <- as.owin(st_union(cck))
tm_owin  <- as.owin(st_union(tm))
jw_owin  <- as.owin(st_union(jw))
pg_owin  <- as.owin(st_union(pg))

childcare_cck_ppp <- childcare_ppp_jit[cck_owin]
childcare_tm_ppp  <- childcare_ppp_jit[tm_owin]
childcare_jw_ppp  <- childcare_ppp_jit[jw_owin]
childcare_pg_ppp  <- childcare_ppp_jit[pg_owin]
```

Rescaling all four to kilometres:

```{r}
childcare_cck_ppp.km <- rescale.ppp(childcare_cck_ppp, 1000, "km")
childcare_tm_ppp.km  <- rescale.ppp(childcare_tm_ppp,  1000, "km")
childcare_jw_ppp.km  <- rescale.ppp(childcare_jw_ppp,  1000, "km")
childcare_pg_ppp.km  <- rescale.ppp(childcare_pg_ppp,  1000, "km")
```

```{r}
data.frame(
  area   = c("Choa Chu Kang", "Tampines", "Jurong West", "Punggol"),
  points = c(npoints(childcare_cck_ppp), npoints(childcare_tm_ppp),
             npoints(childcare_jw_ppp),  npoints(childcare_pg_ppp))
)
```

```{r}
#| fig-height: 7
par(mfrow = c(2, 2))
plot(childcare_cck_ppp, main = "Choa Chu Kang", cex = 0.6, pch = 19)
plot(childcare_tm_ppp,  main = "Tampines",      cex = 0.6, pch = 19)
plot(childcare_jw_ppp,  main = "Jurong West",   cex = 0.6, pch = 19)
plot(childcare_pg_ppp,  main = "Punggol",       cex = 0.6, pch = 19)
par(mfrow = c(1, 1))
```

### 10.3 Clark-Evans Test by Planning Area

```{r}
clarkevans.test(childcare_cck_ppp,
                correction = "none",
                clipregion = NULL,
                alternative = c("two.sided"),
                method = "MonteCarlo",
                nsim = 99)
```

```{r}
clarkevans.test(childcare_tm_ppp,
                correction = "none",
                clipregion = NULL,
                alternative = c("two.sided"),
                method = "MonteCarlo",
                nsim = 99)
```

### 10.4 KDE Surfaces by Planning Area

```{r}
#| fig-height: 7
par(mfrow = c(2, 2))
plot(density(childcare_cck_ppp.km, sigma = bw.diggle, edge = TRUE, kernel = "gaussian"), main = "Choa Chu Kang")
plot(density(childcare_tm_ppp.km,  sigma = bw.diggle, edge = TRUE, kernel = "gaussian"), main = "Tampines")
plot(density(childcare_jw_ppp.km,  sigma = bw.diggle, edge = TRUE, kernel = "gaussian"), main = "Jurong West")
plot(density(childcare_pg_ppp.km,  sigma = bw.diggle, edge = TRUE, kernel = "gaussian"), main = "Punggol")
par(mfrow = c(1, 1))
```

And with a fixed 250m bandwidth for a like-for-like comparison across the four areas:

```{r}
#| fig-height: 7
par(mfrow = c(2, 2))
plot(density(childcare_cck_ppp.km, sigma = 0.25, edge = TRUE, kernel = "gaussian"), main = "Choa Chu Kang")
plot(density(childcare_tm_ppp.km,  sigma = 0.25, edge = TRUE, kernel = "gaussian"), main = "Tampines")
plot(density(childcare_jw_ppp.km,  sigma = 0.25, edge = TRUE, kernel = "gaussian"), main = "Jurong West")
plot(density(childcare_pg_ppp.km,  sigma = 0.25, edge = TRUE, kernel = "gaussian"), main = "Punggol")
par(mfrow = c(1, 1))
```

## 11. What I Take From This Exercise

**The statistical answer.** Childcare centres in Singapore are clustered at the national
scale, decisively so — R = 0.51 with p effectively zero under Monte Carlo simulation. The high-density areas are the mature and newer HDB towns.

**The methodological point that cost me the most time.** The window matters as much as
the points. Running Clark-Evans on the national pattern and running it on Choa Chu Kang
are answering different questions, and the second is not a "zoomed-in" version of the
first — changing the window changes the expected nearest-neighbour distance under CSR,
which is the denominator of R. A pattern can be clustered nationally and close to random
within a single town, and that is not a contradiction.

**What I would flag as a limitation.** The whole analysis treats Singapore's land area as
uniformly available for childcare centres, which is plainly false. Catchment reserves,
airbases, port land and industrial estates cannot host a childcare centre at all. A
clustering result measured against a CSR null that allows points anywhere on land will
therefore always look clustered, partly for reasons that have nothing to do with
childcare policy. A more honest null would restrict the window to residential-zoned land.
That is beyond this exercise, but it is the first thing I would change.

**Connecting to Part 02.** First-order analysis stops at "where is it dense". It cannot
distinguish a pattern that is dense because the population is dense from a pattern that
is dense because centres genuinely attract each other. The G, F, K and L functions in
[Part 02](Part02.qmd) are what separate those two.
