---
title: "Local Measures of Spatial Autocorrelation"
subtitle: "HANDS-ON EXERCISE 5 / PART 02"
author:
  - name: Jedidiah Asaf Tallulembang
    affiliation: "ISSS626 Geospatial Analytics and Applications"
date: last-modified
format:
  html:
    code-fold: false
    toc: true
execute:
  warning: false
  message: false
---

::: {.pill-row}
[R]{.pill} [spdep]{.pill} [LISA]{.pill} [Getis-Ord Gi*]{.pill} [Cluster maps]{.pill}
:::

This exercise follows [Local Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap10), in the Spatial
Weights block of the course.

## 1. Overview

[Part 01](Part01.qmd) established that Hunan's GDP per capita is spatially clustered:
Moran's I of 0.3007 with p near 1e-06. That is one number for 88 counties, and it cannot
say *where* the clustering is.

This part decomposes it. **Local Moran's I (LISA)** returns a statistic per county and
classifies each into high-high, low-low, high-low or low-high. **Getis-Ord Gi\*** asks a
different question again: is this neighbourhood's total unusually high or low?

I expected the local maps to be a more detailed version of the global result. They are
not, and the gap between them is the most useful thing here. I also found an error in the
reference code along the way, and it inverted the finding.

## 2. Getting Started

Same packages and the same Hunan data as Part 01, re-imported so this page stands alone.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan")
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
hunan <- left_join(hunan, hunan2012, by = c("County" = "County")) %>%
  select(1:4, 7, 15)
```

```{r}
wm_q <- poly2nb(hunan, queen = TRUE)
rswm_q <- nb2listw(wm_q, style = "W", zero.policy = TRUE)
```

Queen contiguity, row-standardised, averaging 5.09 neighbours per county.

## 3. Local Indicators of Spatial Association

The global tests say clustering exists. They cannot say where. LISA decomposes Moran's I
into one value per county.

### 3.1 Computing Local Moran's I

```{r}
fips <- order(hunan$County)
localMI <- localmoran(hunan$GDPPC, rswm_q)
head(localMI)
```

```{r}
hunan.localMI <- cbind(hunan, localMI) %>%
  rename(Pr.Ii = Pr.z....E.Ii..)
```

### 3.2 Mapping Local Moran's I and its p-values

```{r}
#| fig-height: 5
localMI.map <- tm_shape(hunan.localMI) +
  tm_polygons("Ii", fill.scale = tm_scale_intervals(style = "pretty"),
              fill.legend = tm_legend(title = "Local Moran's I")) +
  tm_borders(col_alpha = 0.5)

pvalue.map <- tm_shape(hunan.localMI) +
  tm_polygons("Pr.Ii",
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
                values = "brewer.-rd_pu"),
              fill.legend = tm_legend(title = "p-value")) +
  tm_borders(col_alpha = 0.5)

tmap_arrange(localMI.map, pvalue.map, asp = 1, ncol = 2)
```

Only **11 of 88** counties reach significance at the 0.05 level.

### 3.3 LISA Cluster Map

The cluster map classifies each significant county into one of four quadrants. This is
the code the chapter gives.

```{r}
quadrant <- vector(mode = "numeric", length = nrow(localMI))
hunan$lag_GDPPC <- lag.listw(rswm_q, hunan$GDPPC)
DV <- hunan$lag_GDPPC - mean(hunan$lag_GDPPC)
LM_I <- localMI[, 1] - mean(localMI[, 1])
signif <- 0.05

quadrant[DV < 0 & LM_I > 0] <- 1   # low-high
quadrant[DV > 0 & LM_I < 0] <- 2   # high-low
quadrant[DV < 0 & LM_I < 0] <- 3   # low-low
quadrant[DV > 0 & LM_I > 0] <- 4   # high-high
quadrant[localMI[, 5] > signif] <- 0
```

Before mapping it I checked the output against the underlying values, and it does not
survive that check.

```{r}
lab <- c("insignificant", "low-high", "high-low", "low-low", "high-high")
tibble(County = hunan$County,
       GDPPC = hunan$GDPPC,
       lag = round(hunan$lag_GDPPC),
       label = lab[quadrant + 1]) %>%
  filter(quadrant != 0) %>%
  arrange(label, -GDPPC) %>%
  kable()
```

Wugang has a GDPPC of 12,112 against a provincial mean of 24,405, and its neighbours
average 11,566. A poor county surrounded by poor counties is the definition of low-low.
The code labels it **low-high**.

::: {.callout-important}
## The chapter's quadrant code is wrong, and it inverts the finding
The second axis in that code is `LM_I`, the local Moran statistic itself. That cannot
work, and the reason is in the formula. Local Moran's I for county *i* is

$$I_i = z_i \sum_j w_{ij} z_j$$

a product of the county's own standardised deviation and its neighbours'. When both are
positive, high-high, the product is positive. When both are negative, low-low, the
product is **also positive**. Local Moran's I is positive for high-high and low-low
alike, and negative for high-low and low-high alike. Using it as an axis therefore
conflates exactly the pairs the quadrant map exists to separate.

The correct second axis is the county's own deviation from the mean, `GDPPC - mean(GDPPC)`,
which is what a Moran scatterplot actually plots.

```{r}
quadrant2 <- vector(mode = "numeric", length = nrow(localMI))
X <- hunan$GDPPC - mean(hunan$GDPPC)
Y <- hunan$lag_GDPPC - mean(hunan$lag_GDPPC)

quadrant2[X < 0 & Y > 0] <- 1   # low-high
quadrant2[X > 0 & Y < 0] <- 2   # high-low
quadrant2[X < 0 & Y < 0] <- 3   # low-low
quadrant2[X > 0 & Y > 0] <- 4   # high-high
quadrant2[localMI[, 5] > signif] <- 0
```

```{r}
tibble(Quadrant = lab,
       `Chapter code` = map_int(0:4, ~sum(quadrant == .x)),
       `Corrected` = map_int(0:4, ~sum(quadrant2 == .x))) %>%
  kable()
```

**Five of the 11 significant counties change label.** Low-low goes from 0 to 2, high-low
from 3 to 0, and high-high from 6 to 7.

I had written a whole paragraph about how Hunan has no low-low clusters at all before I
ran this check. It was an artefact of the code, not a property of Hunan. That is the
uncomfortable part: the wrong version produced a *more* interesting story, and I would
have submitted it.
:::

```{r}
#| fig-height: 5
hunan.localMI$quadrant <- quadrant2
colours <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")

tm_shape(hunan.localMI) +
  tm_fill("quadrant",
          fill.scale = tm_scale_categorical(
            values = colours[sort(unique(quadrant2)) + 1],
            labels = lab[sort(unique(quadrant2)) + 1]),
          fill.legend = tm_legend(title = "LISA cluster")) +
  tm_borders(col_alpha = 0.5)
```

The corrected classification of the 11 significant counties:

| Quadrant | Counties | Count |
|---|---|---|
| High-high | Changsha, Wangcheng, Liuyang, Miluo, Liling, Zhuzhou, Xiangtan | 7 |
| Low-low | Wugang, Longhui | 2 |
| Low-high | Taojiang, Pingjiang | 2 |
| High-low | none | 0 |
| Insignificant | | 77 |

::: {.callout-note}
## What the corrected map actually says
The asymmetry I thought I had found is still there, just smaller than the buggy version
claimed. Seven counties form high-high clusters against two low-low. Wealth clusters
about three times as often as poverty does.

There are also **zero high-low** counties, meaning no rich county sits isolated in a poor
neighbourhood. Wealth in Hunan does not appear alone. Every significantly rich county has
rich neighbours, which is a stronger statement about spillover than Moran's I of 0.30
conveys on its own.

And 77 of 88 counties are locally insignificant. The global test returned p near 1e-06,
yet seven eighths of the province is indistinguishable from a random arrangement. A
significant global statistic does not mean most places participate in the pattern.
:::


## 4. Hot and Cold Spot Analysis with Getis-Ord Gi*

LISA asks whether a county resembles its neighbours. Gi* asks whether a neighbourhood's
*total* is unusually high or low, and answers in z-score units.

### 4.1 Deriving the Weights

```{r}
coords <- st_coordinates(st_centroid(st_geometry(hunan)))
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

::: {.callout-warning}
## I walked into the same trap I documented last exercise
My first run passed `st_centroid()` output straight into `knearneigh()` and got a maximum
first-neighbour distance of **0.60** instead of **61.81**. That is the exact `sfc` versus
coordinate-matrix mistake I wrote a callout about in Exercise 4, and I made it again
within a fortnight.

Writing the lesson down was evidently not the same as learning it. What saved me was the
habit the last one left behind: I looked at the number, knew 0.6 km was impossible for
Chinese counties, and checked. The value of documenting a mistake was not that I stopped
making it, but that I now recognise the symptom on sight.
:::

Maximum first-nearest-neighbour distance is **61.81 km**, the cut-off that guarantees
every county at least one neighbour.

### 4.2 Gi* with Fixed Distance Weights

Gi* includes the county itself, which is what `include.self()` is for and what the star
in the name denotes.

```{r}
d62 <- dnearneigh(coords, 0, 61.81, longlat = TRUE)
gi.fixed <- localG(hunan$GDPPC, nb2listw(include.self(d62), style = "B"))
hunan.gi <- cbind(hunan, as.matrix(gi.fixed)) %>%
  rename(gstat_fixed = as.matrix.gi.fixed.)
```

```{r}
#| fig-height: 5
tm_shape(hunan.gi) +
  tm_polygons("gstat_fixed",
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, -2, -1, 1, 2, Inf),
                values = "brewer.-rd_bu"),
              fill.legend = tm_legend(title = "Gi* (fixed 62km)")) +
  tm_borders(col_alpha = 0.5)
```

```{r}
gi_f <- as.numeric(gi.fixed)
tibble(County = hunan$County, GDPPC = hunan$GDPPC, `Gi*` = round(gi_f, 2)) %>%
  arrange(-`Gi*`) %>% head(7) %>% kable()
```

Gi* runs from **-1.90 to +5.81**. Seven counties exceed +2. **Not one falls below -2.**

::: {.callout-note}
## Xiangyin shows what Gi* is actually measuring
Xiangyin has a GDPPC of 33,983, which ranks **13th** of 88. Its Gi* ranks **3rd**, behind
only Wangcheng and Changsha.

Its neighbours explain it: Wangcheng at 70,666 and Miluo at 42,497, alongside Yuanjiang,
Yueyang and Taojiang. Gi* scores the neighbourhood sum including the county, so Xiangyin
inherits a large score from company it keeps rather than from its own output. LISA, by
contrast, leaves Xiangyin insignificant at p = 0.051.

The two statistics are answering different questions rather than giving two views of one
answer. If the task is siting infrastructure to serve concentrated economic mass, Gi* is
the right tool. If the task is finding counties that have fallen behind their own
neighbours, LISA is.
:::

### 4.3 Gi* with Adaptive Distance Weights

```{r}
knn8 <- knn2nb(knearneigh(coords, k = 8))
gi.adaptive <- localG(hunan$GDPPC, nb2listw(include.self(knn8), style = "B"))
hunan.gi$gstat_adaptive <- as.numeric(gi.adaptive)
```

```{r}
#| fig-height: 5
tm_shape(hunan.gi) +
  tm_polygons("gstat_adaptive",
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, -2, -1, 1, 2, Inf),
                values = "brewer.-rd_bu"),
              fill.legend = tm_legend(title = "Gi* (adaptive k=8)")) +
  tm_borders(col_alpha = 0.5)
```

```{r}
gi_a <- as.numeric(gi.adaptive)
tibble(Scheme = c("Fixed 62 km", "Adaptive k = 8"),
       `Avg neighbours` = c(round(mean(card(d62)), 2), 8),
       Minimum = round(c(min(gi_f), min(gi_a)), 2),
       Maximum = round(c(max(gi_f), max(gi_a)), 2),
       `Hot (> +2)` = c(sum(gi_f > 2), sum(gi_a > 2)),
       `Cold (< -2)` = c(sum(gi_f < -2), sum(gi_a < -2))) %>%
  kable()
```

::: {.callout-important}
## Whether Hunan has any cold spots at all depends on the weights
The fixed-distance scheme finds **7 hot spots and 0 cold spots**. The adaptive scheme
finds **15 hot spots and 5 cold spots**. Same variable, same province, same threshold.

The mechanism is the neighbour count. The fixed 62 km band averages 3.64 neighbours, and
in the sparse western counties it gives very few. A Gi* z-score built from three units
cannot reach far from zero, so the poor west never crosses -2. Forcing k = 8 gives those
same counties a wider window, their deficit accumulates over more neighbours, and cold
spots appear.

Pingjiang makes the point sharply. Its GDPPC is 17,252, below the provincial median of
20,433, so it is a genuinely poor county. Under the fixed band its Gi* is +1.95; under
adaptive k = 8 it is **+4.61**, which would be reported as a strong hot spot. The
corrected LISA classifies it as low-high, a poor county surrounded by wealth, which is
the honest description.

"Hunan has no economic cold spots" is a sentence I could have written from the fixed-band
map alone, and it would have been an artefact of the neighbour definition. Report the
weights with the result, every time.
:::


## 5. What I Take Away

**I caught a real error in the reference code, and only because I checked the output
against the data.** The chapter's quadrant assignment uses local Moran's I as an axis, and
because that statistic is positive for high-high and low-low alike, it cannot separate
them. Five of 11 significant counties were mislabelled, low-low was reported as empty when
there are two, and three spurious high-low counties appeared. Running the code and mapping
the result would have shown nothing wrong. Reading one row against its own values did.

**The wrong answer was the more interesting one.** "Hunan has no low-low clusters at all"
is a better line than "seven high-high against two low-low", and I had already written it.
Findings that flatter the analysis deserve more scepticism than findings that complicate
it, not less.

**Global and local measures answer different questions.** Part 01's Moran's I of 0.30 at
p near 1e-06 sounds province-wide. Locally, 77 of 88 counties are indistinguishable from
random and the pattern is carried by a minority around Changsha. A significant global
statistic does not mean most places participate.

**Two statistics can both be right and disagree.** Xiangyin is 13th by GDPPC and 3rd by
Gi\*, and insignificant under LISA. Neither is broken; Gi\* scores the neighbourhood and
LISA scores the county against its neighbours. The choice belongs to the question being
asked.

**Weights are not a preliminary step.** Whether Hunan has any cold spots at all flips
between 0 and 5 on the neighbour definition alone. Exercise 4 presented weights as setup
for the real analysis. They are part of the result.

**And I repeated a documented mistake.** The `sfc` versus coordinate-matrix trap cost me
time in Exercise 4, I wrote it up carefully, and I made it again here. What had stuck was
the sanity check rather than the rule, which is probably the realistic version of learning
from a mistake.
