---
title: "Global Measures of Spatial Autocorrelation"
subtitle: "HANDS-ON EXERCISE 5 / PART 01"
author:
  - name: Jedidiah Asaf Tallulembang
    affiliation: "ISSS626 Geospatial Analytics and Applications"
date: last-modified
format:
  html:
    code-fold: false
    toc: true
execute:
  warning: false
  message: false
---

::: {.pill-row}
[R]{.pill} [spdep]{.pill} [Moran's I]{.pill} [Geary's C]{.pill} [Correlogram]{.pill}
:::

This exercise follows [Global Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap09), in the Spatial
Weights block of the course.

## 1. Overview

Exercise 4 built the weights matrix. This exercise finally uses it to answer a question
worth asking: **is Hunan's economic development spatially clustered, or is it scattered
at random across the province?**

That matters because the two cases call for different policy. If wealthy counties are
scattered, each one succeeded on its own terms and there is little to generalise. If they
are clustered, something is spilling across county borders, and the unit of intervention
is the region rather than the county.

The techniques split into two levels:

- **Global** measures return one number for the whole map. Moran's I and Geary's C
  answer "is there clustering anywhere".
- **Local** measures return one number per county. LISA and Getis-Ord Gi* answer
  "clustering *where*", which is the question a planner actually has.

I expected the local maps to be a more detailed version of the global result. They are
not, and the gap between them turned out to be the most useful thing in this exercise.

## 2. The Study Area and Data

| Dataset | Description | Format |
|---|---|---|
| `Hunan` | County boundary layer for Hunan province, China, 88 counties | ESRI Shapefile (polygon) |
| `Hunan_2012.csv` | Selected local development indicators for 2012 | CSV |

Same data as Exercise 4, which is deliberate. Having already built and inspected the
weights matrix for these 88 counties, I can attribute anything surprising here to the
statistics rather than to the geography being unfamiliar.

The variable of interest is **GDPPC**, GDP per capita, which ranges from 8,497 to 88,656
with a mean of 24,405. That is better than a tenfold spread across a single province.

## 3. Getting Started

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

## 4. Getting the Data Into R

### 4.1 Importing the Shapefile

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan")
```

### 4.2 Importing the CSV

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### 4.3 Performing the Relational Join

```{r}
hunan <- left_join(hunan, hunan2012, by = c("County" = "County")) %>%
  select(1:4, 7, 15)
```

## 5. Visualising the Distribution

Before any test, look at the map. A statistic that disagrees with the picture usually
means the statistic was computed wrong.

```{r}
#| fig-height: 5
equal <- tm_shape(hunan) +
  tm_polygons("GDPPC", fill.scale = tm_scale_intervals(n = 5, style = "equal"),
              fill.legend = tm_legend(title = "Equal interval")) +
  tm_borders(col_alpha = 0.5)

quantile <- tm_shape(hunan) +
  tm_polygons("GDPPC", fill.scale = tm_scale_intervals(n = 5, style = "quantile"),
              fill.legend = tm_legend(title = "Equal quantile")) +
  tm_borders(col_alpha = 0.5)

tmap_arrange(equal, quantile, asp = 1, ncol = 2)
```

::: {.callout-note}
## The two classifications disagree about how many rich counties there are
Equal interval puts almost every county in the bottom class, because Changsha at 88,656
stretches the range so far that the other 87 counties are squeezed together. Equal
quantile forces 20% into each class and shows a north-east cluster clearly.

Neither is wrong. Equal interval is honest about the magnitude of the gap; quantile is
honest about the ranking. The reason this matters for the rest of the exercise is that
Moran's I works on deviations from the mean, and that mean of 24,405 is dragged upward
by a handful of counties. Most of Hunan sits below its own provincial average.
:::

## 6. Global Measures of Spatial Autocorrelation

### 6.1 Computing Row-standardised Weights

```{r}
wm_q <- poly2nb(hunan, queen = TRUE)
rswm_q <- nb2listw(wm_q, style = "W", zero.policy = TRUE)
rswm_q
```

Queen contiguity, row-standardised, as built in Exercise 4. Average of 5.09 neighbours
per county.

### 6.2 Moran's I Test

```{r}
moran.test(hunan$GDPPC, listw = rswm_q, zero.policy = TRUE, na.action = na.omit)
```

Moran's I is **0.3007** against an expected value of **-0.0115** under spatial
randomness. The z-score is **4.735** and p is about **1.1e-06**.

So: positive spatial autocorrelation. Counties resemble their neighbours more than
chance would produce. On the standard reading of Moran's I, which runs roughly -1 to +1,
0.30 is moderate rather than overwhelming.

### 6.3 Monte Carlo Moran's I

The analytical p-value assumes normality, and GDPPC is clearly skewed. A permutation
test avoids that assumption entirely by shuffling the values across counties 999 times
and asking where the observed value falls.

```{r}
set.seed(1234)
bperm <- moran.mc(hunan$GDPPC, listw = rswm_q, nsim = 999, zero.policy = TRUE, na.action = na.omit)
bperm
```

The observed statistic ranks **1000 out of 1000**, giving p = 0.001. No permutation of
the data produced clustering as strong as the real arrangement. The conclusion survives
dropping the normality assumption, which is the point of running it.

### 6.4 Visualising the Simulation

```{r}
#| fig-height: 4
hist(bperm$res, freq = TRUE, breaks = 20, xlab = "Simulated Moran's I",
     main = "999 permutations against the observed value")
abline(v = 0, col = "grey40", lty = 2)
abline(v = bperm$statistic, col = "red", lwd = 2)
```

The simulated distribution sits around zero, as it should when values are scattered at
random. The observed value sits outside the whole distribution.

### 6.5 Geary's C Test

```{r}
geary.test(hunan$GDPPC, listw = rswm_q)
```

Geary's C is **0.6907** against an expected 1. Values below 1 indicate positive
autocorrelation, so this agrees with Moran's I.

### 6.6 Monte Carlo Geary's C

```{r}
set.seed(1234)
bperm_geary <- geary.mc(hunan$GDPPC, listw = rswm_q, nsim = 999)
bperm_geary
```

p = 0.001 again.

::: {.callout-note}
## Why run both when they agree
Moran's I is built on cross-products of deviations from the global mean, so it responds
to the overall pattern. Geary's C is built on squared differences between neighbouring
pairs, so it is more sensitive to local differences between adjacent units.

When they agree, as here, the reading is that the pattern is not an artefact of one
formulation. If Moran's I had come back positive while Geary's C indicated no structure,
that would have pointed to a broad regional gradient without sharp neighbour-to-neighbour
similarity, and would have been worth chasing. Agreement is the boring outcome, but it is
only boring because it was checked.
:::

## 7. Spatial Correlogram

Contiguity only looks at immediate neighbours. A correlogram extends the question: how
far does the similarity reach before it dies out?

### 7.1 Moran's I Correlogram

```{r}
#| fig-height: 5
MI_corr <- sp.correlogram(wm_q, hunan$GDPPC, order = 6, method = "I", style = "W")
plot(MI_corr)
```

```{r}
print(MI_corr)
```

| Lag | Moran's I |
|---|---|
| 1 | +0.3007 |
| 2 | +0.2060 |
| 3 | +0.0668 |
| 4 | +0.0299 |
| 5 | -0.1530 |
| 6 | -0.1187 |

::: {.callout-important}
## The sign flip is the finding, not the decay
I expected the correlation to decay toward zero as lag increases, and it does through
lag 4. What I did not expect is that it goes **negative** at lags 5 and 6.

Decay to zero would mean influence simply fades with distance. Going negative means
something stronger: counties five or six steps apart are *systematically dissimilar*.
That is the signature of a single dominant cluster rather than a smooth gradient. Once
you walk five counties away from the rich north-east, you are reliably in the poor
periphery, so the correlation inverts.

A gradient across the province would have produced a slow decay that stayed positive.
The negative tail says Hunan has a core and a periphery, and the boundary between them is
crossed at about four steps.
:::

### 7.2 Geary's C Correlogram

```{r}
#| fig-height: 5
GC_corr <- sp.correlogram(wm_q, hunan$GDPPC, order = 6, method = "C", style = "W")
plot(GC_corr)
```

Geary's C mirrors it, rising above 1 at the longer lags, which is the same statement in
the opposite direction.


## 8. Additional Exploration: Does the Choice of Weights Change Moran's I?

Everything above used queen contiguity. Exercise 4 built four different weighting schemes,
and I was left wondering how much of Moran's I is the data and how much is the matrix.

First the distance-based neighbours, which need a cut-off large enough that no county is
left isolated.

```{r}
coords <- st_coordinates(st_centroid(st_geometry(hunan)))
k1 <- knn2nb(knearneigh(coords))
dmax <- max(unlist(nbdists(k1, coords, longlat = TRUE)))
round(dmax, 2)
```

::: {.callout-warning}
## Pass coordinates, not geometry
My first attempt passed `st_centroid()` output straight into `knearneigh()`, an `sfc`
object rather than a coordinate matrix, and the maximum first-neighbour distance came back
as **0.60** instead of **61.81**. The `longlat = TRUE` argument is silently ignored, so
`spdep` measures in degrees as though they were planar units.

It does not error. It produces a perfectly well-formed weights matrix that is completely
wrong. I made this exact mistake in Exercise 4, wrote it up, and made it again here. What
saved me was the habit the first one left behind: look at the number and ask whether 0.6 km
is plausible for Chinese counties.
:::

Now the four schemes.

```{r}
d62 <- dnearneigh(coords, 0, dmax, longlat = TRUE)
wm_r <- poly2nb(hunan, queen = FALSE)
k6 <- knn2nb(knearneigh(coords, k = 6))
schemes <- list(Queen = wm_q, Rook = wm_r, `knn (k=6)` = k6, `Fixed 62km` = d62)

map_dfr(names(schemes), function(nm) {
  lw <- nb2listw(schemes[[nm]], style = "W", zero.policy = TRUE)
  m <- moran.test(hunan$GDPPC, lw, zero.policy = TRUE)
  tibble(Scheme = nm,
         `Avg neighbours` = round(mean(card(schemes[[nm]])), 2),
         `Moran's I` = round(unname(m$estimate[1]), 4),
         z = round(unname(m$statistic), 3),
         p = format.pval(m$p.value, digits = 3))
}) %>% kable()
```

| Scheme | Avg neighbours | Moran's I | z | p |
|---|---|---|---|---|
| Queen | 5.09 | 0.3007 | 4.735 | 1.09e-06 |
| Rook | 5.00 | 0.3084 | 4.810 | 7.53e-07 |
| knn (k=6) | 6.00 | 0.3429 | 6.429 | 6.41e-11 |
| Fixed 62 km | 3.64 | 0.4210 | 5.388 | 3.56e-08 |

::: {.callout-note}
## The direction is robust. The magnitude is not.
Every scheme agrees that autocorrelation is positive, and the weakest p-value is still
about 1e-06. That conclusion is safe.

Moran's I itself ranges from **0.30 to 0.42**, a swing of 40% of the lower value. Reporting
"Moran's I for Hunan GDPPC is 0.42" without naming the matrix would not be reproducible by
someone who reasonably chose queen contiguity.

The mechanism is the neighbour count. The fixed-distance scheme has the fewest neighbours
at 3.64, so each county is compared against a tighter and nearer set, and near neighbours
are more alike. Fewer and closer neighbours means a higher measured index. Part 02 finds
the same sensitivity in the local statistics, where it decides whether Hunan has any cold
spots at all.

The rule I am taking from this: treat the ranking of places as more trustworthy than the
absolute value of any index, and state the weights specification alongside the statistic.
:::


## 9. What I Take Away

**Hunan's development is spatially clustered, and the evidence is not marginal.** Moran's
I of 0.3007 against an expected -0.0115 gives z = 4.735 and p near 1e-06, and the
permutation test puts the observed value outside all 999 simulations. Geary's C agrees
from a different formulation. The conclusion survives dropping the normality assumption,
which is the reason to run the Monte Carlo version at all.

**The correlogram's negative tail was worth plotting.** Decay to zero would mean influence
simply fades with distance. Turning negative at lags 5 and 6 means counties that far apart
are systematically dissimilar, which is the signature of a core and a periphery rather
than a smooth gradient. The boundary is crossed at about four steps.

**The direction is robust, the magnitude is not.** Moran's I moves between 0.30 and 0.42
across four reasonable weighting schemes. Reporting the number without naming the matrix
that produced it would not be reproducible. Fewer and nearer neighbours means a higher
index, every time.

**And a global statistic is an average over the whole map.** It says clustering exists
somewhere. It cannot say where, it cannot say whether the clustering is in the rich
counties or the poor ones, and it cannot say how much of the province participates.
Part 02 takes the same data apart county by county, and the answer turns out to be
narrower than this number suggests.
