---
title: "Emerging Hot Spot Analysis"
subtitle: "IN-CLASS EXERCISE 5"
author:
  - name: Jedidiah Asaf Tallulembang
    affiliation: "ISSS626 Geospatial Analytics and Applications"
date: last-modified
format:
  html:
    code-fold: false
    toc: true
execute:
  warning: false
  message: false
---

::: {.pill-row}
[R]{.pill} [sfdep]{.pill} [Space-time cube]{.pill} [Gi*]{.pill} [Mann-Kendall]{.pill} [EHSA]{.pill}
:::

This exercise follows [Emerging Hot Spot
Analysis](https://r4gdsa.netlify.app/chap11), the first reading in the Geospatial
Multivariate Data Analysis block.

## 1. Overview

Exercise 5 found a hot spot around Changsha in 2012. It could not say whether that hot
spot was new, fading, or permanent, because it only ever saw one year.

Emerging Hot Spot Analysis adds the time axis. It computes Gi* for every location in
every time period, then runs a Mann-Kendall trend test on each location's sequence of
Gi* values, and classifies the result into categories such as persistent, intensifying,
diminishing, sporadic and oscillating.

The question this answers is the one a planner actually cares about: not "where is it
hot" but "where is it *becoming* hot".

I will say at the outset that the headline result of this exercise is a bug I found in
the `sfdep` package rather than a fact about Hunan. The fact about Hunan only became
visible after fixing it.

## 2. The Study Area and Data

| Dataset | Description | Format |
|---|---|---|
| `Hunan` | County boundary layer, 88 counties | ESRI Shapefile (polygon) |
| `Hunan_GDPPC.csv` | GDP per capita by county by year, 2005 to 2021 | CSV (long format) |

The boundary layer is the same one used in Exercises 4 and 5. The attribute file is new
and is a **balanced panel**: 88 counties by 17 years gives 1,496 rows with no gaps and no
missing values. Its 2012 column reconciles with the `Hunan_2012.csv` used in Exercise 5
to within rounding, which is the check that confirms the two files describe the same
place.

## 3. Getting Started

```{r}
pacman::p_load(sf, sfdep, tmap, plotly, tidyverse, Kendall, knitr)
set.seed(1234)
```

## 4. Getting the Data Into R

```{r}
hunan <- st_read(dsn = "data/geospatial", layer = "Hunan")
GDPPC <- read_csv("data/aspatial/Hunan_GDPPC.csv")
```

```{r}
tibble(Rows = nrow(GDPPC),
       Counties = n_distinct(GDPPC$County),
       Years = n_distinct(GDPPC$Year),
       From = min(GDPPC$Year), To = max(GDPPC$Year),
       Balanced = nrow(GDPPC) == n_distinct(GDPPC$County) * n_distinct(GDPPC$Year),
       `Missing GDPPC` = sum(is.na(GDPPC$GDPPC))) %>% kable()
```

### 4.1 What Happened to Hunan Between 2005 and 2021

Before any spatial statistic, it is worth knowing what the variable did.

```{r}
w <- GDPPC %>% pivot_wider(names_from = Year, values_from = GDPPC)
tibble(Measure = c("Mean GDPPC", "Coefficient of variation", "Max / min ratio"),
       `2005` = c(round(mean(w$`2005`)), round(sd(w$`2005`)/mean(w$`2005`), 3),
                  round(max(w$`2005`)/min(w$`2005`), 1)),
       `2021` = c(round(mean(w$`2021`)), round(sd(w$`2021`)/mean(w$`2021`), 3),
                  round(max(w$`2021`)/min(w$`2021`), 1))) %>% kable()
```

::: {.callout-note}
## Everything grew, and the gaps stayed the same shape
Mean GDP per capita rose from about 7,500 to about 50,400, a factor of **6.7** in sixteen
years. That is the headline a growth statistic would report.

The dispersion measures did not move. The coefficient of variation went from 0.467 to
0.471, and the ratio of richest to poorest county actually narrowed slightly, from 7.4 to
6.4. Hunan got six times richer without meaningfully changing how unequally that wealth
was distributed across counties.

This matters for what follows. Because every county grew, and grew at broadly similar
proportional rates, a trend test on raw GDPPC is useless: Kendall's tau on the raw series
averages **0.97** across counties, which just says "GDP went up everywhere". The whole
point of running the trend test on **Gi\*** rather than on GDPPC is that Gi\* is a
*relative* measure. It asks whether a county's standing among its neighbours changed, and
that is a question the raw series cannot answer.
:::

## 5. Creating a Space-Time Cube

A spacetime cube pairs each location with each time period. `sfdep` validates that the
pairing is complete, which is why the balanced-panel check above mattered.

```{r}
GDPPC_st <- spacetime(GDPPC, hunan, .loc_col = "County", .time_col = "Year")
is_spacetime_cube(GDPPC_st)
```

`TRUE` confirms every county appears in every year.

## 6. Computing Gi* Year by Year

First derive the neighbours and weights, then compute Gi* within each year.

```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(st_contiguity(geometry)),
         wt = st_inverse_distance(nb, geometry, scale = 1, alpha = 1),
         .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")
```

```{r}
gi_stars <- GDPPC_nb %>%
  group_by(Year) %>%
  mutate(gi = local_gstar_perm(GDPPC, nb, wt, nsim = 99)) %>%
  tidyr::unnest(gi) %>%
  ungroup()
```

```{r}
#| fig-height: 4
gi_stars %>%
  group_by(Year) %>%
  summarise(Hot = sum(gi_star > 2), Cold = sum(gi_star < -2)) %>%
  pivot_longer(-Year) %>%
  ggplot(aes(Year, value, colour = name)) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.8) +
  scale_colour_manual(values = c(Hot = "#C0392B", Cold = "#2c7bb6")) +
  labs(y = "Counties beyond |Gi*| = 2", colour = NULL,
       title = "Hot and cold counties per year") +
  theme_minimal(base_size = 11)
```

The hot count is remarkably stable at around 11 counties throughout. The cold count drifts
upward in the later years.

## 7. Mann-Kendall on a Single County

Before running the full analysis, it helps to see the test on one location.

```{r}
#| fig-height: 3.6
cbg <- gi_stars %>% filter(County == "Changsha")
ggplot(cbg, aes(Year, gi_star)) +
  geom_line(colour = "#C0392B", linewidth = 0.9) +
  geom_point(colour = "#C0392B", size = 1.8) +
  labs(y = "Gi*", title = "Changsha: Gi* over time") +
  theme_minimal(base_size = 11)
```

```{r}
mk <- Kendall::MannKendall(cbg$gi_star)
tibble(tau = round(unname(mk$tau), 3),
       `2-sided p` = signif(as.numeric(mk$sl), 3),
       S = unname(mk$S)) %>% kable()
```

Changsha's Gi\* is high in every year and trends gently upward. Nothing surprising.

## 8. Emerging Hot Spot Analysis

```{r}
ehsa_raw <- emerging_hotspot_analysis(x = GDPPC_st, .var = "GDPPC",
                                      k = 1, nsim = 99)
ehsa_raw %>% count(classification, sort = TRUE) %>% kable()
```

### 8.1 Checking the Output Before Mapping It

The classification is only meaningful if it agrees with the Gi\* series it was computed
from. A county labelled a hot spot should have high Gi\*. So I joined the two together.

```{r}
county_gi <- gi_stars %>%
  group_by(County) %>%
  summarise(mean_gi = mean(gi_star),
            years_hot = sum(gi_star > 2),
            mean_GDPPC = mean(GDPPC))

ehsa_raw %>%
  left_join(county_gi, by = c("location" = "County")) %>%
  group_by(classification) %>%
  summarise(n = n(),
            `mean Gi*` = round(mean(mean_gi), 2),
            `mean GDPPC` = round(mean(mean_GDPPC))) %>%
  arrange(-n) %>% kable()
```

The counties labelled **hotspot** have a *lower* mean Gi\* than the counties labelled
**coldspot**. That is not a subtle problem.

```{r}
ehsa_raw %>%
  left_join(county_gi, by = c("location" = "County")) %>%
  filter(years_hot == 17) %>%
  select(location, mean_gi, years_hot, classification) %>%
  mutate(mean_gi = round(mean_gi, 2)) %>%
  arrange(-mean_gi) %>% kable()
```

Nine counties have Gi\* above 2 in **all seventeen years**. Wangcheng has a mean Gi\* of
about 5, the second highest in the province, and sits next to Changsha in the richest
corner of Hunan. It is labelled a **coldspot**.

### 8.2 Locating the Bug

`emerging_hotspot_analysis()` ends with this line (sfdep 0.2.5):

```r
res_ehs <- cbind(
  location = attr(x, "geometry")[[.loc_col]],
  do.call(rbind, lapply(split(all_gis, x[[.loc_col]]),
                        function(.x) classify_hotspot(.x, threshold)))
)
```

::: {.callout-important}
## The results are computed in alphabetical order and labelled in shapefile order
`split()` groups by the location column and returns those groups **sorted alphabetically**.
That is the order the classification rows come out in.

`location` is taken from `attr(x, "geometry")[[.loc_col]]`, which is the county column of
the **shapefile**, in whatever order the file happens to store its features.

`cbind()` then pastes the two together positionally. The two orderings only agree if the
shapefile is already sorted alphabetically. Hunan's is not.

```{r}
tibble(Position = 1:5,
       `Shapefile order` = head(hunan$County, 5),
       `Alphabetical order` = head(sort(hunan$County), 5)) %>% kable()
```

Row 1 of the classification block belongs to Anhua, the first county alphabetically, but
gets labelled Anxiang, the first county in the shapefile. Every row is attached to the
wrong county. The output is a valid-looking table of the right shape, with the labels
shuffled.

This is the same failure mode as the `sfc` trap in Exercise 4 and the quadrant error in
Exercise 5. It does not raise an error, it does not look wrong, and every map built from
it would be confidently mislabelled.
:::

### 8.3 The Fix, and the Test That Confirms It

The fix is to name the rows in the order they were actually computed.

```{r}
ehsa <- ehsa_raw %>% mutate(location = sort(hunan$County))
```

If that diagnosis is right, the relationship between label and Gi\* should invert cleanly.

```{r}
bind_rows(
  ehsa_raw %>% mutate(version = "As shipped"),
  ehsa     %>% mutate(version = "Reassigned")
) %>%
  left_join(county_gi, by = c("location" = "County")) %>%
  filter(classification %in% c("sporadic hotspot", "sporadic coldspot")) %>%
  group_by(version, classification) %>%
  summarise(n = n(), `mean Gi*` = round(mean(mean_gi), 2), .groups = "drop") %>%
  kable()
```

As shipped, hot spots and cold spots are indistinguishable by Gi\* and sit on the wrong
sides of zero. After reassignment the hot spots sit well above zero and the cold spots
well below. The nine counties that are hot in all seventeen years all classify as hot
spots.

```{r}
ehsa %>%
  left_join(county_gi, by = c("location" = "County")) %>%
  group_by(classification) %>%
  summarise(n = n(),
            `mean Gi*` = round(mean(mean_gi), 2),
            `mean GDPPC` = round(mean(mean_GDPPC))) %>%
  arrange(-`mean Gi*`) %>% kable()
```

::: {.callout-note}
## Why this table is the proof, not the earlier one
The corrected classes line up monotonically with **two independent quantities at once**:
mean Gi\*, which comes from my own separate per-year computation, and mean GDPPC, which
is the raw input.

Hot spots are the richest counties and have the highest Gi\*. Cold spots are the poorest
and have the lowest. Oscillating hot spots sit in between on both. No shuffling of labels
would produce that alignment by chance, and the shipped version produced none of it.
:::

## 9. Mapping the Corrected Result

```{r}
#| fig-height: 5.5
hunan_ehsa <- hunan %>% left_join(ehsa, by = c("County" = "location"))

tm_shape(hunan_ehsa) +
  tm_polygons("classification",
              fill.scale = tm_scale_categorical(values = "brewer.set2"),
              fill.legend = tm_legend(title = "EHSA class")) +
  tm_borders(col_alpha = 0.4) +
  tm_title("Emerging hot spot analysis, Hunan GDPPC 2005-2021")
```

```{r}
#| fig-height: 5.5
hunan_sig <- hunan_ehsa %>%
  mutate(shown = ifelse(p_value < 0.05, classification, NA))

tm_shape(hunan_sig) +
  tm_polygons("shown",
              fill.scale = tm_scale_categorical(values = "brewer.set2"),
              fill.legend = tm_legend(title = "Significant only"),
              col_alpha = 0.4) +
  tm_borders(col_alpha = 0.4) +
  tm_title("Significant classifications only (p < 0.05)")
```

Half the province reaches significance. The rest is not evidence of stability, only
absence of evidence of change.

## 10. What the Corrected Analysis Says About Hunan

### 10.1 The Core is Fixed, Not Emerging

```{r}
ehsa %>%
  left_join(county_gi, by = c("location" = "County")) %>%
  filter(years_hot == 17) %>%
  select(County = location, `mean Gi*` = mean_gi, `mean GDPPC` = mean_GDPPC,
         classification) %>%
  mutate(`mean Gi*` = round(`mean Gi*`, 2), `mean GDPPC` = round(`mean GDPPC`)) %>%
  arrange(-`mean Gi*`) %>% kable()
```

These nine counties are the Changsha-Zhuzhou-Xiangtan corridor that Exercise 5 found in a
single year of data. Seventeen years of data show it was already there in 2005 and never
weakened. Exercise 5's snapshot was not a snapshot of a moment, it was a snapshot of a
structure.

### 10.2 The Movement is in a Middle Ring

```{r}
ehsa %>%
  left_join(county_gi, by = c("location" = "County")) %>%
  filter(classification == "oscilating hotspot", p_value < 0.05) %>%
  select(County = location, tau, `mean Gi*` = mean_gi, `mean GDPPC` = mean_GDPPC) %>%
  mutate(tau = round(tau, 2), `mean Gi*` = round(`mean Gi*`, 2),
         `mean GDPPC` = round(`mean GDPPC`)) %>%
  arrange(-tau) %>% kable()
```

::: {.callout-note}
## The interesting counties are the ones in the middle
The oscillating hot spots are not the rich core and not the poor west. Their mean GDPPC
sits between the two groups, and their Kendall tau values are strongly positive, meaning
their standing relative to their neighbours has been climbing.

Several of them, Yueyang, Linxiang, Huarong and Anxiang, sit north of Changsha toward
Dongting Lake. The reading that fits is that the Changsha core is not growing in
intensity, since it was already saturated in 2005, but its influence is reaching further
out, and the counties on that frontier are the ones whose relative position is changing.

That is a genuinely different statement from anything Exercise 5 could produce, and it is
the thing EHSA exists to find.
:::

### 10.3 The Cold Periphery Widened

```{r}
gi_stars %>%
  filter(Year %in% c(2005, 2021)) %>%
  group_by(Year) %>%
  summarise(`Hot (Gi* > 2)` = sum(gi_star > 2),
            `Cold (Gi* < -2)` = sum(gi_star < -2)) %>% kable()
```

The hot count barely moved. The cold count grew. Combined with the flat coefficient of
variation from section 4.1, the picture is a province where the top did not pull away in
relative terms, but the bottom became more spatially concentrated. Poverty in Hunan became
more *clustered* over the period even though it did not become relatively deeper.

## 11. What I Take Away

**Three exercises, three silent failures, and none of them threw an error.** The `sfc`
coordinate trap in Exercise 4, the quadrant axis error in Exercise 5, and now an ordering
mismatch inside a published CRAN package. Every one produced output of the correct shape
and plausible appearance. The only thing that caught any of them was checking a result
against the values it claimed to describe.

**The check that works is cheap and always the same.** Join the output back to the input
and ask whether it is consistent. Counties called hot spots should have high Gi\*. That
took one line and four seconds, and it is the difference between this exercise and a
confidently wrong map.

**Trusting a package is not the same as trusting a result.** `sfdep` is on CRAN, is
written by a competent author, and implements all seventeen ArcGIS EHSA classes correctly.
The classification logic is fine. One `cbind` at the end attaches it to the wrong rows. A
package being good is not evidence that a particular call to it returned what you think.

**The bug is invisible on data that happens to be sorted.** If Hunan's shapefile had
stored its counties alphabetically, this code would be correct and I would have found
nothing. Whether an analysis is right would then depend on the storage order of a file
nobody looks at. That is worth remembering when a workflow "works on my machine".

**On Hunan itself:** the Changsha corridor is not an emerging hot spot, it is a permanent
one, hot in all seventeen years and already established before the data begins. The change
is happening in a ring of middle-income counties to its north whose relative standing is
rising, and in a western periphery that has become more tightly clustered in its
disadvantage. Growth of 6.7 times province-wide moved the levels and left the structure
almost exactly as it found it.
