---
title: "Where peat burns, and how close together"
subtitle: "TAKE-HOME EXERCISE 1"
description: |
  A first- and second-order point-pattern analysis of VIIRS active-fire detections
  in Pulang Pisau Regency, Central Kalimantan, 1 July to 25 September 2026.
author:
  - name: Jedidiah Asaf Tallulembang
    affiliation: "ISSS626 Geospatial Analytics and Applications"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "code"
    code-tools: true
    fig-width: 8
    fig-height: 5
    fig-dpi: 200
    fig-align: left
    tbl-cap-location: bottom
    fig-cap-location: bottom
    number-sections: true
abstract: |
  I analyse VIIRS active-fire detections over Pulang Pisau Regency, Central
  Kalimantan, for the 2026 dry season. I picked this regency because it has the
  highest detection density of any Kalimantan regency big enough for second-order
  estimation at the kilometre scale. I wanted to answer three things: where the
  detections concentrate, whether they interact beyond that concentration, and what
  a fire agency could honestly do with the answer.

  The thing that shaped the whole analysis is a distinction I nearly missed. A VIIRS
  hotspot is a 375 m pixel seen during one satellite overpass, not a fire. I measured
  the consequence: the raw detection count overstates the number of distinct burning
  locations by a factor of about 3.4. So testing the raw pattern against complete
  spatial randomness rejects the null overwhelmingly and tells me nothing, because the
  sensor guarantees that result. I switched the unit of analysis to the 375 m cell and
  the null to an inhomogeneous Poisson process, re-estimating the reference intensity
  for every simulated pattern. That gives a test that could fail. It does not.

  What I found: clustering beyond the trend is strongest at about 440 m and only
  decays to within 10% of no excess by about 7 km. That last number corrects my own
  first attempt, which capped the distance range at 5 km and concluded the excess
  stopped at 2 km. A Thomas process fitted after removing the trend puts the dominant
  cluster scale at 1.1 km, with 95% of a cluster inside 2.6 km. Fitted without
  removing the trend it returns 3.3 km and 440 events per cluster, which is really
  just the peat belt re-described as a few giant clusters. A Knox permutation test
  finds space-time interaction concentrated within a few hundred metres and a few
  days. Distance to the road network accounts for 1,991 of the 2,005 AIC units
  separating my covariate model from an intercept-only model, while mapped drainage
  accounts for almost none, which turned out to say more about OpenStreetMap coverage
  than about drainage.
execute:
  echo: true
---

::: {.pill-row}
[R]{.pill} [sf]{.pill} [spatstat]{.pill} [NASA FIRMS]{.pill} [KDE]{.pill} [Kinhom]{.pill} [Knox test]{.pill} [Thomas process]{.pill}
:::

[**Executive summary, 10 slides**](slides.qmd){.btn} · [**Repository**](https://github.com/jediasaf/isss626-coursework/tree/main/Take-home_Ex/Take-home_Ex01)


```{r setup}
#| include: false
knitr::opts_chunk$set(fig.showtext = FALSE, out.width = "100%")
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(sf); library(ggplot2); library(glue)
  library(knitr); library(spatstat.geom); library(spatstat.explore)
  library(spatstat.model); library(spatstat.random); library(patchwork)
})
source("R/00_config.R")
source("R/helpers.R")

P  <- readRDS(file.path(DIR_DERIVED, "prepared.rds"))
FO <- readRDS(file.path(DIR_DERIVED, "firstorder.rds"))
SO <- readRDS(file.path(DIR_DERIVED, "secondorder.rds"))
ST <- readRDS(file.path(DIR_DERIVED, "spacetime.rds"))
CM <- readRDS(file.path(DIR_DERIVED, "covariates_models.rds"))
EX <- readRDS(file.path(DIR_DERIVED, "extended.rds"))
TH <- readRDS(file.path(DIR_DERIVED, "thomas.rds"))
RANK <- readRDS(file.path(DIR_DERIVED, "regency_ranking.rds"))
PROBE <- readRDS(file.path(DIR_DERIVED, "retention_probe.rds"))
GAP <- readRDS(file.path(DIR_DERIVED, "gap_report.rds"))

win_sf <- P$win_sf; win_p <- P$win_p; bb <- st_bbox(win_p)
gg_win <- list(
  geom_sf(data = win_p, fill = NA, colour = PAL$ink, linewidth = 0.4),
  coord_sf(expand = FALSE, datum = NA),
  scalebar_layer(bb), north_layer(bb), theme_map()
)
```

# Introduction and study context

Indonesian Borneo burns every dry season, and when I started reading about it I
learned that the fires which matter most for public health are not forest fires in
the ordinary sense. They are peat fires. Drained tropical peat is a metres-deep
carbon store that, once dry, smoulders below the surface for weeks. You cannot put
it out by suppressing a surface flame front, and it releases a dense particulate
smoke that carries hundreds of kilometres. So the public-safety burden is
respiratory rather than thermal: school closures, airport closures, and a
measurable excess of acute respiratory cases across Central Kalimantan and, in bad
years, across the Malacca Strait.

Pulang Pisau Regency sits at the centre of this. Its southern half was the site of
the Mega Rice Project, a 1990s programme that cut thousands of kilometres of
drainage canal into deep peat and was then abandoned. The canals are still there,
the water table is still lowered, and the land is still unusually combustible. That
is why I did not treat the regency as just a convenient study window. It is a place
where the link between land management and public-safety outcomes is about as
visible as it gets.

Fire detections are point events in space and time, so point-pattern analysis is
the obvious tool. What I wanted to work out in this exercise is what that tool can
honestly support once you take the data seriously.

## The problem the data handed me

I need to state one thing before any analysis, because it ended up governing every
interpretation I made.

A VIIRS active-fire detection is not a fire. It is a 375 m pixel in which a
radiometer measured a thermal anomaly during a single overpass. A large peat fire
covering ten square kilometres produces many dozens of detections in one pass, and
the two satellites pass overhead roughly four times a day between them, so the same
fire gets detected over and over. What a detection actually records is *the sensor
sampling fire*, not fire occurrence. Any statistic I compute on the raw detection
set is therefore partly measuring the instrument.

Once I realised that, the textbook second-order question stopped being interesting.
"Is this pattern clustered relative to complete spatial randomness" already has a
known answer before I compute anything. It is clustered, overwhelmingly, and it
would still be clustered even if fires were sited completely at random, because
each fire produces a burst of adjacent detections. Reporting that rejection as a
finding would be close to meaningless, so I did not want to build the report on it.

Instead I changed two things, and I am stating them up front because they are the
real methodological content of this exercise rather than housekeeping. I changed
the unit of analysis from the detection to the 375 m cell, and I changed the null
model from complete spatial randomness to an inhomogeneous Poisson process.
Section 7 is where that pays off.

# Research questions and objectives

I set four questions, each with a null and a reason for asking it. I wrote these
before running the second-order analysis, so the report can be read as a test of
them rather than as a description of whatever came out.

**Q1. How does the intensity of fire detection vary across the regency, and at what
spatial scale is that variation stable?**
First-order. I ask it because a fire agency allocates fixed resources to places, so
what is useful to them is a surface, not a test. I added the scale qualifier because
a kernel estimate can be made to show almost any structure depending on the
bandwidth, so I do not think the question is answerable unless I also defend that
choice.

**Q2. How does intensity vary in time, and is the temporal signal a property of fire
activity or of the observing system?**
First-order. I split this out because the two satellites observe at fixed local
times, so any apparent daily cycle is a sampling artefact. I wanted to be sure I did
not read the orbit as fire behaviour.

**Q3. Once large-scale variation in intensity is accounted for, is there residual
interaction between events, and over what distance?**
Second-order. My null is an inhomogeneous Poisson process with intensity
$\hat{\lambda}(u)$ estimated by kernel smoothing. Rejecting this null tells me
something that rejecting complete spatial randomness does not, because the
inhomogeneous null already contains everything my first-order analysis found. The
"over what distance" part is the bit with operational value, and it is also the part
I initially got wrong.

**Q4. Is there space-time interaction, over and above the separate spatial and
temporal patterns?**
Second-order, spatio-temporal. My null is random labelling: I hold the observed
locations and the observed set of dates fixed and permute only the pairing between
them. This conditions on both marginal patterns, which matters, because it means a
rejection cannot be explained away by the seasonal ramp or by the spatial
concentration. I already know both of those exist.

I added a fifth question as a value-added extension in Section 9, because the four
above tell me clustering exists without telling me what it is organised around or
how big a cluster actually is.

# Data and study design

## Sources

```{r tbl-sources}
#| label: tbl-sources
#| tbl-cap: "The data I used, as accessed. Both sources are re-creatable by running `R/01_acquire.R`. I pinned the geoBoundaries release to a commit hash so the boundary geometry cannot change underneath my analysis."
tibble::tribble(
  ~Item, ~Detail,
  "Event data", "NASA FIRMS, VIIRS 375 m Collection 2, near-real-time (NRT)",
  "Platforms", "Suomi NPP and NOAA-20, both retained",
  "Access route", "FIRMS `area` REST API, CSV, 5-day request blocks",
  "Access date", format(ACCESS_DATE, "%d %B %Y"),
  "Spatial extent requested", "108.5 to 119.1 E, 5.2 S to 4.5 N (all Kalimantan)",
  "Boundaries", "geoBoundaries gbOpen Indonesia ADM1 and ADM2, commit 9469f09",
  "Boundary vintage", "2020, sourced from BPS Statistics Indonesia via OCHA",
  "Boundary licence", "CC BY 3.0 IGO",
  "Covariates", "OpenStreetMap via Overpass: `highway`, `waterway`",
  "Software", glue("R {getRversion()}, sf {packageVersion('sf')}, spatstat.explore {packageVersion('spatstat.explore')}")
) |> kable()
```

## Why I chose this regency

I chose the study window with a rule applied to the whole island, after downloading
data for all 56 Kalimantan regencies and cities and screening it for confidence.
The rule: **the highest detection density among regencies large enough to support
second-order estimation to a 5 km lag**, which I operationalised as an area above
5,000 km².

I used density rather than raw count because intensity relative to window area is
what governs how much signal a point pattern carries. I added the size floor because
a 5 km lag inside a small window starts to approach the window's own dimension, and
at that point edge effects dominate every second-order estimate I would compute.

```{r tbl-rank}
#| label: tbl-rank
#| tbl-cap: "The ten Kalimantan regencies with the most retained VIIRS detections, 1 July to 25 September 2026. Ketapang leads on count; Pulang Pisau leads on density among regencies above the 5,000 km² size floor, which is the selection rule applied. Kota Palangka Raya's very high density in a small window illustrates why the floor is needed."
RANK |> slice_head(n = 10) |>
  transmute(
    Regency = shapeName, Province = province,
    Detections = fmt_n(n_detections),
    `Area (km²)` = fmt_n(round(area_km2)),
    `Per 1,000 km²` = fmt_n(round(per_1000km2))
  ) |> kable(align = "llrrr")
```

Pulang Pisau wins on that rule outright, at
`r round(RANK$per_1000km2[RANK$shapeName == STUDY_REGENCY])` detections per
1,000 km². Two alternatives were seriously considered and rejected, and the
reasons are recorded because a choice is only reviewable alongside them.

**Ketapang** beat it on absolute count, with
`r fmt_n(RANK$n_detections[RANK$shapeName == "Ketapang"])` detections against Pulang
Pisau's `r fmt_n(RANK$n_detections[RANK$shapeName == STUDY_REGENCY])`. I did not
choose it because it covers
`r fmt_n(round(RANK$area_km2[RANK$shapeName == "Ketapang"]))` km², slightly over
three times the area, and ranks second on density. Its bigger count is mostly just
a bigger window.

**Kota Palangka Raya** has the highest density of any unit at
`r fmt_n(round(RANK$per_1000km2[RANK$shapeName == "Kota Palangka Raya"]))` per
1,000 km², but covers only
`r fmt_n(round(RANK$area_km2[RANK$shapeName == "Kota Palangka Raya"]))` km². A 5 km
lag there is a large fraction of the window's own width, so edge effects would drive
my second-order results rather than the fires.

Pulang Pisau also comes rank
`r which(RANK$shapeName == STUDY_REGENCY)` of `r nrow(RANK)` on raw count, so I am
not selecting some obscure unit to make the rule work. Its window of
`r fmt_n(round(P$area_km2))` km², at
`r round((bb["xmax"]-bb["xmin"])/1000)` km by
`r round((bb["ymax"]-bb["ymin"])/1000)` km, comfortably supports estimation to a
5 km lag.

I should be open about how I got here. My first pass selected the regency from a
seven-day sample of the same feed, on which Pulang Pisau led on count. Over the full
87-day window it does not, Ketapang does. So I restated the selection rule against
the complete record and re-applied it. I am reporting that rather than quietly
fixing it, because a rule chosen after you have seen which unit it favours is not
really a rule, and I would rather the reader be able to judge that for themselves.

## Point event definition

I define an event as **one 375 m grid cell in which at least one qualifying VIIRS
detection was recorded**. Qualifying means all of the following, and I have given
the reason for each rather than just the rule.

```{=html}
<dl class="assump">
<dt>Sensor and product</dt>
<dd>VIIRS 375 m I-band, Collection 2, S-NPP or NOAA-20. MODIS is excluded: its
1 km footprint is coarser than the distances this analysis is about, so including
it would blur the very scale under study.</dd>
<dt>Confidence</dt>
<dd>Nominal or high. Low-confidence detections are dropped. They are dominated by
small or cool anomalies and by edge-of-scan geometry, so retaining them would
make apparent intensity vary with viewing angle rather than with fire.</dd>
<dt>Coordinates</dt>
<dd>Present, finite, within valid latitude and longitude ranges, and not the
(0, 0) sentinel that some pipelines emit for a failed geolocation.</dd>
<dt>Timestamp</dt>
<dd>A parseable UTC acquisition time. Converted to WITA (UTC+8), because the
definition of a fire day and any reading of diurnal behaviour must be local.</dd>
<dt>Location</dt>
<dd>Strictly within the Pulang Pisau polygon. This also serves as the offshore
screen, since the polygon is terrestrial.</dd>
<dt>Not a fixed installation</dt>
<dd>See the persistence screen below. This substitutes for a field the NRT
product does not carry.</dd>
<dt>Deduplicated</dt>
<dd>Exact duplicate records removed, then collapsed to one event per 375 m cell
per local day.</dd>
</dl>
```

I build three patterns from the same screened detections, because which unit is
right depends on which question I am asking:

| Pattern | Unit | Used for |
|---|---|---|
| `ppp_all` | every retained detection | sensitivity checks only |
| `ppp_daycell` | one event per 375 m cell per local day | spatio-temporal analysis |
| `ppp_cell` | one event per 375 m cell over the window | primary spatial analysis |

I use `ppp_cell` as the primary spatial pattern because it removes the
repeat-detection inflation I described in Section 1.1. A cell that burned for three
weeks and was detected eighty times contributes one point, not eighty.

For the representative point of each cell I take the earliest detection in it. That
is deterministic and it keeps a real sensor coordinate. I considered snapping to
cell centres and rejected it, because that would impose a regular lattice and
manufacture apparent regularity at small lags, which would contaminate exactly the
statistic the rest of the report depends on.

## Spatial study window

My window is the Pulang Pisau administrative polygon, unbuffered, in WGS 84 /
UTM zone 49S (EPSG:32749).

Projecting is not optional here. Every second-order statistic I use is a function of
Euclidean distance in metres, and at this latitude one degree of longitude is about
111 km while one degree of latitude is about 110.6 km. If I worked in degrees I
would be building a systematic anisotropy into every distance I compute. Zone 49S is
the right one for a regency centroid at 113.96°E, which sits inside the zone's 108°E
to 114°E span. I checked the eastern edge, which reaches 114.35°E and is marginally
outside, and the resulting scale error is below 0.1%, so it does not matter at the
kilometre scales I work at.

I used the administrative boundary rather than a buffered or natural window because
the management question is administrative: a regency fire agency has jurisdiction to
that line. The cost is a real edge effect, since fires obviously do not stop at a
boundary, and I handle it with edge correction rather than pretending it is not
there (Section 5).

## Temporal study window

**1 July to 25 September 2026, 87 days.** The lower bound was imposed on me rather
than chosen, and I want to state the constraint precisely, because it is the kind of
thing that is easy to misrepresent by accident.

```{r tbl-probe}
#| label: tbl-probe
#| tbl-cap: "Retention probe. Single-day FIRMS NRT queries over a Kalimantan-wide box. Exactly zero detections across a 10.6° by 9.7° box for every sampled date from January to 30 June is not a physically credible result for Borneo, so the zeros are read as absence of coverage. The standard-processing archive, queried for the same region in August, also returns zero rows: it has not yet been built for 2026."
PROBE$nrt |> transmute(Date = format(date, "%d %b %Y"),
                       `Detections returned` = fmt_n(n)) |> kable(align = "lr")
```

The FIRMS `area` endpoint serves near-real-time granules on a rolling retention of
roughly 90 days, and the standard-processing archive that would have covered January
to June has not been produced for 2026. So the earliest day I can retrieve is 1 July
2026, which I located to the day with the probe above rather than assuming it. The
upper bound is my choice: the last complete UTC day before I accessed the data, so
that no partially observed day enters the daily series and shows up as a fake
decline.

I want to be careful not to oversell the window. It is what was available. But it
does happen to span the onset, escalation and peak of the 2026 dry season, which is
the period a fire agency actually acts over. If the retention limit had cut off the
peak instead of the wet-season baseline I would have had to change the study design
entirely. The honest cost is that I cannot make any wet-season comparison, so nothing
in this report speaks to the seasonal amplitude of fire activity, only to its
structure within one season. I repeat that in Section 12 because it is a real limit
on what I can claim.

## My completeness audit, and a real outage

Counting the rows I downloaded does not tell me the record is complete. The failure
mode I was worried about is a request that comes back empty, because that is
indistinguishable from a day on which nothing burned, and it would show up in my
analysis as a drop in fire activity that never happened.

So I built the audit to catch exactly that. Each of the two platforms should
contribute at least one detection somewhere in Kalimantan on every day of an 87-day
dry season, because an island-sized box is never completely without fire. Any day
where a platform contributes nothing is a candidate failure and worth chasing.

```{r tbl-gap}
#| label: tbl-gap
#| tbl-cap: "Completeness audit. Local dates on which one VIIRS platform contributed no detection anywhere in the Kalimantan bounding box. Each was re-queried four times with backoff by `R/01b_fill_gaps.R`, and each returned empty again, so these are not failed requests."
GAP$missing_after |> transmute(Date = format(dt, "%d %B %Y"),
                               `Platform absent` = platform) |> kable(align = "ll")
```

```{r}
#| label: gap-inline
#| include: false
bad <- GAP$missing_after$dt
ev_bad <- sum(P$daily_daycell$n_events[P$daily_daycell$date_local %in% bad])
ev_all <- sum(P$daily_daycell$n_events)
```

Six of the seven dates are consecutive, which made me suspect a real outage rather
than a flaky connection. Re-querying confirmed it: Suomi NPP returns an empty result
for 11 to 16 July and for 3 August no matter how many times I ask. So this is an
absence in the FIRMS near-real-time record itself, not a transient request failure I
could fix by retrying. On those seven days my analysis rests on NOAA-20 alone, and
detection counts are depressed by roughly the contribution of one platform.

The consequence is small, but I do not want to write it off as nothing. Those dates
carry
`r fmt_n(ev_bad)` of `r fmt_n(ev_all)` events, or
`r sprintf("%.2f%%", 100 * ev_bad / ev_all)` of the record, and all of them fall in
the near-zero early season before the dry-season ramp. Mean daily events on the
affected dates are `r round(mean(P$daily_daycell$n_events[P$daily_daycell$date_local %in% bad]), 1)`
against `r round(mean(P$daily_daycell$n_events[P$daily_daycell$date_local >= as.Date("2026-07-01") & P$daily_daycell$date_local <= as.Date("2026-07-26") & !(P$daily_daycell$date_local %in% bad)]), 1)`
on nearby unaffected July dates, which is consistent with losing one platform on
very small counts. No conclusion I draw turns on those seven days, but the reader
should treat the early-season part of my daily series as a slight underestimate.

# Data preparation and quality assessment

## The audit trail

I counted every exclusion. The table below is generated by the pipeline rather than
typed in by me, so it cannot drift away from what the code actually did.

```{r tbl-qa}
#| label: tbl-qa
#| tbl-cap: "My screening audit. Each row is one rule applied in `R/02_prepare.R`, in the order I applied it. Counts are rows of the detection table, before I collapse to cell-level events."
P$qa |> transmute(Step = step, Rule = rule,
                  `In` = fmt_n(n_in), Removed = fmt_n(n_removed), `Out` = fmt_n(n_out)) |>
  kable(align = "llrrr")
```

Three things I want to point out about this table.

The biggest single reduction is the spatial clip, and that is expected. I downloaded
all of Kalimantan so that I could justify my choice of regency against the
alternatives, and the overwhelming majority of those detections are outside Pulang
Pisau.

The coordinate and timestamp screens removed nothing at all. I am flagging that
rather than skipping past it, because in my experience a cleaning step that catches
nothing is often a cleaning step that is not actually testing what it says it is. I
checked mine against deliberately broken records while developing the pipeline, and
they do fire. The FIRMS product is simply well-formed here: every record had finite
coordinates in range and a parseable timestamp. I kept the screens in because I want
the pipeline to be re-runnable against other windows and products where that will
not be true.

The confidence screen removed
`r percent(P$qa$n_removed[P$qa$step == "03 confidence"] / P$qa$n_in[P$qa$step == "03 confidence"], accuracy = 0.1)`
of records, and exact duplicate removal a further
`r fmt_n(P$qa$n_removed[P$qa$step == "06 exact duplicates"])` rows, confirming
that the API does deliver a small number of repeated granule records.

```{r tbl-conf}
#| label: tbl-conf
#| tbl-cap: "Confidence classes among detections surviving the coordinate and timestamp screens, before the confidence screen itself is applied. VIIRS confidence is categorical, not a percentage."
P$conf_tab |> transmute(Confidence = confidence, Detections = fmt_n(n),
                        Share = percent(n / sum(n), accuracy = 0.1)) |>
  kable(align = "lrr")
```

## The missing `type` field, and what I replaced it with

This was the data-quality problem that took me longest to solve. The FIRMS
standard-processing product carries a `type` field separating vegetation fire (0)
from active volcano (1), other static land source (2) and offshore (3). The NRT
product I have to use does not carry it, which I verified against the API response
header rather than assuming: `r glue("column present = {P$has_type}")`. So I had to
reconstruct the screen it would have given me.

I worked through the three cases separately:

- **Volcanoes (type 1)** are not a risk at all. Borneo has no Holocene volcano, so
  this screen is unnecessary rather than approximated. I did not need to do anything.
- **Offshore (type 3)** is already handled by my clip to the terrestrial polygon.
- **Other static land sources (type 2)** are a genuine problem. Kalimantan has
  palm-oil mills, coal operations and gas flares, and all of them show up as
  persistent thermal anomalies. This is the case I actually had to solve.

My substitute is a **persistence screen**. The logic is that a vegetation fire burns
out or moves, whereas a flare stack does not. So I score each 375 m cell by the
number of distinct local dates on which it was detected, as a fraction of the 87-day
window, and drop cells above a threshold as fixed installations.

```{r tbl-persist}
#| label: tbl-persist
#| tbl-cap: "Sensitivity of the persistence screen. The adopted threshold is 60% of days, at which the screen removes nothing. The grid extends well below that value to show where the screen would begin to act, which is the evidence that it is functional rather than merely quiet."
P$persist_sens |> transmute(
  `Threshold (share of days)` = percent(threshold, accuracy = 1),
  `Cells removed` = fmt_n(n_cells),
  `Detections removed` = fmt_n(n_detections)) |> kable(align = "lrr")
```

**My screen removed nothing.** My first reaction was that I had written it wrong, so
I checked it two ways, and I now think the empty result is correct.

First, no cell comes anywhere near the threshold. The most persistent cell in the
whole window was detected on `r P$persist_max_days` of `r P$n_days` days, which is
`r percent(P$persist_max_days / P$n_days, accuracy = 0.1)`. A flare stack or mill
would be detected on most clear-sky overpasses across the entire window, so nothing
here behaves like one.

Second, I used radiative power as an independent check, because it separates the two
cases without relying on persistence at all. A fixed industrial source emits a high
and steady radiative power, while smouldering peat emits a low one.

```{r tbl-persist-top}
#| label: tbl-persist-top
#| tbl-cap: "The ten most persistently detected cells, with their mean fire radiative power. If these were industrial sources their mean FRP would sit far above the window median; instead they sit at or near it, which is the signature of recurrent low-intensity peat combustion."
P$persist_top |> transmute(
  Cell = cell, Detections = n_det, `Days detected` = n_days,
  `Share of window` = percent(day_frac, accuracy = 0.1),
  `Mean FRP (MW)` = mean_frp) |> kable(align = "lrrrr")
```

The median FRP across all my retained detections is `r round(P$frp_median_all, 1)`
MW, and the most persistent cells range from `r min(P$persist_top$mean_frp)` to
`r max(P$persist_top$mean_frp)` MW. They are not brighter than the population at
all. They are ordinary peat fires that happened to recur in the same cell.

That fits the regency's land use. Pulang Pisau is peatland and smallholder
agriculture, without the gas processing or open-cut coal operations that generate
persistent thermal anomalies elsewhere in Kalimantan. If I had picked a window in
East Kalimantan instead, I think this screen would have been doing real work.

I report the sensitivity across thresholds rather than a single number on purpose. If
the result swung sharply depending on where I put the line, that would be a decision
dressed up as a method, and the reader deserves to see which one it is.

## My evidence that a detection is not a fire

The claim I made in Section 1.1 is checkable, so I checked it. If detections were
independent events, most occupied cells would be detected exactly once. They are
not.

```{r fig-mult}
#| label: fig-mult
#| fig-cap: "Detections per occupied 375 m cell over the 87-day window, log scale on both axes. If each detection were an independent event, the distribution would be concentrated at one. Instead a long tail of cells is detected tens or hundreds of times, which is the repeat-observation signature of persistent fire in a fixed pixel grid observed roughly four times daily."
#| fig-height: 3.6
mt <- P$multiplicity
ggplot(mt, aes(n_det, n_cells)) +
  geom_segment(aes(xend = n_det, yend = 0), colour = PAL$accent, linewidth = 0.35, alpha = 0.85) +
  scale_x_log10(labels = label_number(accuracy = 1)) +
  scale_y_log10(labels = label_number(big.mark = ",", accuracy = 1)) +
  labs(x = "Detections recorded in the cell", y = "Number of cells",
       title = "One fire, many detections",
       subtitle = glue("{fmt_n(sum(mt$n_cells))} occupied cells. Median {median(rep(mt$n_det, mt$n_cells))} detections per cell, ",
                       "maximum {max(mt$n_det)}.")) +
  theme_ex01()
```

```{r}
#| label: mult-stats
#| include: false
mt_v <- rep(mt$n_det, mt$n_cells)
share_once <- mean(mt_v == 1)
infl <- P$ppp_all$n / P$ppp_cell$n
```

Only `r percent(share_once, accuracy = 0.1)` of occupied cells were detected exactly
once. The mean is `r round(mean(mt_v), 1)` detections per occupied cell, so the raw
count overstates the number of distinct burning locations by a factor of about
`r round(infl, 1)`. That is the number that convinced me. Analysing the raw
detections as a point pattern would mean analysing the satellite's revisit schedule
about as much as the fire regime, which is why I made `ppp_cell` the primary unit.

```{r tbl-patterns}
#| label: tbl-patterns
#| tbl-cap: "The three patterns I built and their mean intensities. The gap between the first and third rows is the repeat-detection inflation I quantified above."
tibble::tibble(
  Pattern = c("`ppp_all`", "`ppp_daycell`", "`ppp_cell`"),
  Unit = c("retained detection", "cell-day", "cell"),
  n = c(P$ppp_all$n, P$ppp_daycell$n, P$ppp_cell$n),
  `Mean intensity (per 1,000 km²)` = round(c(P$ppp_all$n, P$ppp_daycell$n, P$ppp_cell$n) / P$area_km2 * 1000)
) |> mutate(n = fmt_n(n)) |> kable(align = "llrr")
```

## Geometry validation

I checked the boundary polygon for validity before using it. It is a single valid
polygon, so I did not need any multipart or self-intersection handling, though I ran
`st_make_valid` anyway as a defensive step. For point-in-polygon assignment I used
`st_within` rather than `st_intersects`, so a detection sitting exactly on the
boundary arc is excluded rather than double-counted against a neighbouring regency.
`spatstat`'s window conversion reported no points rejected as outside the window,
which tells me my clip and the `owin` agree.

Two things here commonly go wrong and I handled them deliberately. First, I assigned
regencies to parent provinces by testing containment of an interior representative
point in a projected coordinate system, not by nearest-feature matching. Nearest
feature would silently mis-assign any regency whose representative point fell
marginally outside its own province, and I would never have noticed. Second, I read
the 159 MB national ADM2 file through a bounding-box filter at the GDAL level rather
than loading it and subsetting afterwards, so the full national geometry is never
held in memory.

# Analytical assumptions and methodological decisions

Listing assumptions is easy. What I found harder, and more useful, was working out
where each one might actually fail, so I have written each with the reason I think
it could break rather than just the statement.

```{=html}
<dl class="assump">
<dt>A1. Detections are a censored observation of fire, not a sample of fires</dt>
<dd>VIIRS detects fires above a size and temperature threshold that varies with scan
angle, cloud and canopy. Smouldering sub-surface peat combustion, the dominant mode
in this landscape, is systematically under-detected because its surface signature is
weak. So what I am analysing is the pattern of <em>detectable</em> fire. That biases
against exactly the fire type I care most about for public health, and I have no
correction available within this data source. My intensity estimates are therefore a
lower bound, and their spatial variation is partly variation in detectability rather
than in fire.</dd>

<dt>A2. Cloud cover is not spatially confounded with fire location</dt>
<dd>I assume this and I cannot fully defend it. Dry-season convection over Borneo is
not spatially uniform, and smoke itself attenuates the signal. If cloud were
systematically thicker over one part of the regency, apparent intensity there would
be depressed and I would read it as less fire. I cannot establish the direction of
that bias from the detection data alone.</dd>

<dt>A3. The window is fixed and known, and events outside it are irrelevant</dt>
<dd>This is false as physics and I adopt it as a management framing. Fires cross the
regency boundary constantly. That is why I edge-correct every second-order estimate,
and why I report how sensitive my conclusions are to the correction rather than
assuming it away.</dd>

<dt>A4. The 375 m cell is the finest meaningful spatial resolution</dt>
<dd>This follows from the sensor rather than from any choice of mine. It imposes a
hard minimum separation between distinct events, and I can see it in the pair
correlation function as apparent inhibition below roughly 400 m. That inhibition is
a property of the instrument. Reading it as fires avoiding one another would be a
serious mistake, and an easy one to make.</dd>

<dt>A5. For the spatial analysis, time is collapsed</dt>
<dd>My `ppp_cell` pattern treats a cell burning in July and a cell burning in
September as the same kind of point. That is fine for "where does fire occur over a
season" and wrong for "where is fire now". I relax it in Section 8.</dd>

<dt>A6. The intensity used as the inhomogeneous null is estimable from the data</dt>
<dd>This is the weakest link in my second-order analysis and I want to be upfront
about it. Estimating $\hat{\lambda}(u)$ from the same pattern I am testing lets the
estimate absorb some of the clustering under test, which pushes the envelope outward
and makes my test conservative. I mitigate it two ways: I pick a bandwidth that
describes broad trend rather than local structure, and I re-estimate
$\hat{\lambda}$ for every simulated pattern so the data and the simulations get
identical treatment. The bias that remains pushes me toward <em>failing</em> to
detect clustering, which is why I am reasonably confident the clustering I do detect
is not an artefact of this choice.</dd>

<dt>A7. Isotropy</dt>
<dd>$K$, $L$ and $g$ all assume it, because they average over direction. I doubt it
holds here: peat fires spread with the wind and along canal lines, so I would expect
real anisotropy. Any directional structure gets averaged into a weaker isotropic
signal, so once again my reported clustering is a conservative reading rather than
an inflated one.</dd>

<dt>A8. Random labelling is the right null for space-time interaction</dt>
<dd>Permuting time labels across fixed locations conditions on both marginals. That
makes it the correct null for "is there interaction beyond the two separate
patterns" and the wrong null for "is the seasonal ramp real", which it simply
assumes. I intend that. I establish the ramp in Section 6 and I am not trying to
test it here.</dd>
</dl>
```

## My decisions, and what I rejected

```{r tbl-decisions}
#| label: tbl-decisions
#| tbl-cap: "The methodological decisions I made. The right-hand column records the alternative I considered and did not take, which I think is the part that makes a decision reviewable."
tibble::tribble(
  ~Decision, ~Choice, ~`Rejected alternative and why`,
  "Coordinate system", "EPSG:32749, UTM 49S", "Geographic coordinates: distance-based statistics require equal units",
  "Event unit", "one 375 m cell", "raw detections: inflates n by ~4x and measures revisit schedule",
  "Representative point", "earliest detection in cell", "cell centre: imposes a lattice, manufactures small-scale regularity",
  "Sensor", "VIIRS 375 m only", "adding MODIS 1 km: coarser than the scales under study",
  "Bandwidth", "fixed 4 km, justified against the interaction range", "bw.ppl and bw.diggle at ~0.6 km: absorb the clustering the second-order test is meant to detect",
  "Kernel", "Gaussian, isotropic", "quartic: negligible difference, less standard diagnostics",
  "Edge correction, KDE", "Diggle's correction", "none: 86 km-wide window, boundary understated",
  "Edge correction, K and L", "border", "Ripley isotropic: cost not justified at this n",
  "r range, K and L", "0 to 5 km", "larger: elongated window makes long lags unreliable",
  "Null model, primary", "inhomogeneous Poisson", "CSR: rejected trivially, uninformative",
  "Envelope construction", "lambda re-estimated per simulation", "fixed lambda: absorbs the clustering under test",
  "Significance", "MAD and DCLF global tests", "pointwise envelopes alone: inflate type-I error",
  "Simulations", glue("{SO$nsim} for envelopes, {ST$nperm} for Knox"), "fewer: coarse Monte Carlo p-value floor"
) |> kable()
```

Two of these are worth explaining properly.

**Border versus Ripley correction.** Ripley's isotropic correction is more efficient
but costs a lot more computation, and my envelope procedure needs `r SO$nsim`
re-computations of it. At `r fmt_n(SO$n_cell)` points the efficiency I lose with the
border correction does not matter, because the number of pairs contributing at each
lag is already very large. So this is a compute trade-off and I made it deliberately
rather than by accepting a default.

**Global versus pointwise tests.** This one took me a while to get straight. A
pointwise envelope gets read at every one of several hundred values of $r$, so the
probability of at least one excursion under the null is far above the nominal level.
I still plot the pointwise envelopes, because they show me *where* on the $r$ axis
the departure sits and a single p-value cannot. But every claim of significance I
make in this report is backed by the MAD or DCLF global test, not by the band.

# First-order analysis: where, and when

## The pattern in its region

```{r fig-context}
#| label: fig-context
#| fig-cap: "Detection density across the 56 regencies and cities of Indonesian Borneo, 1 July to 25 September 2026, with Pulang Pisau outlined. The concentration in the southern peat belt of Central Kalimantan is the regional structure within which the study window sits. What I take from it: the choice of study window is not an outlier hunt. Pulang Pisau is the most intense cell of a coherent regional band running along the southern coastal peat domes, so findings about its internal structure plausibly transfer to its neighbours. The map also shows what the regency boundary does analytically: it cuts a continuous field of fire activity, which is why the edge corrections of Section 5 matter rather than being a formality."
#| fig-height: 5.2
kal <- readRDS(file.path(DIR_DERIVED, "kal_adm2.rds")) |> st_transform(STUDY_CRS)
kal_m <- kal |> left_join(RANK |> select(shapeName, per_1000km2), by = "shapeName") |>
  mutate(per_1000km2 = tidyr::replace_na(per_1000km2, 0))
ggplot(kal_m) +
  geom_sf(aes(fill = per_1000km2), colour = "white", linewidth = 0.18) +
  geom_sf(data = win_p, fill = NA, colour = PAL$ink, linewidth = 0.8) +
  scale_fill_intensity(name = "Detections per 1,000 km\u00b2", v = kal_m$per_1000km2) +
  annotate("text", x = bb["xmax"] + 60000, y = (bb["ymin"] + bb["ymax"]) / 2,
           label = "Pulang Pisau", hjust = 0, size = 3.1, colour = PAL$ink,
           family = "Helvetica") +
  annotate("segment", x = bb["xmax"] + 55000, xend = bb["xmax"] + 4000,
           y = (bb["ymin"] + bb["ymax"]) / 2, yend = (bb["ymin"] + bb["ymax"]) / 2,
           colour = PAL$ink, linewidth = 0.3) +
  coord_sf(datum = NA, clip = "off") +
  labs(title = "A regional band, not an isolated hotspot") +
  theme_map() + theme(plot.margin = margin(5, 90, 5, 5))
```

## The point pattern

```{r fig-points}
#| label: fig-points
#| fig-cap: "The primary spatial pattern: one point per 375 m cell in which fire was detected. Colour shows the local date of first detection, so the map carries the season as well as the location. What I take from it: two things are immediately visible and both matter. The pattern is strongly concentrated in the southern two-thirds of the regency, which is the drained peat of the former Mega Rice Project, while the northern interior is nearly empty. Second, the colour gradient is not spatially random: earlier detections (lighter) sit disproportionately in the south and east, later ones spread north and west. That is a testable claim about space-time interaction rather than an impression, and Section 8 tests it."
#| fig-height: 6.4
pts_df <- P$fires |> st_coordinates() |> as.data.frame() |>
  bind_cols(P$fires |> st_drop_geometry() |> select(date_local))
first_seen <- P$ppp_cell
pc <- tibble(x = first_seen$x, y = first_seen$y,
             date = marks(first_seen)$date_local)
ggplot() +
  geom_sf(data = win_p, fill = PAL$ground_alt, colour = NA) +
  geom_point(data = pc, aes(x, y, colour = as.numeric(date - WINDOW_START) + 1),
             size = 0.16, alpha = 0.55) +
  scale_colour_gradientn(colours = c("#F6DFC8", "#E2803C", "#C2410C", "#4A1403"),
                         name = "Day of window (1 July = 1)",
                         breaks = c(1, 30, 60, 87)) +
  gg_win +
  labs(title = glue("{fmt_n(P$ppp_cell$n)} burned cells, coloured by first detection"))
```

## Quadrat counts, and why I do not lean on them

```{r tbl-quad}
#| label: tbl-quad
#| tbl-cap: "Chi-squared test of a homogeneous Poisson process, across quadrat resolutions. The variance-to-mean ratio (VMR) is 1 under complete spatial randomness. The verdict is identical at every resolution, which is the point: the test is not informative, it is merely arithmetically certain."
FO$quad_summary |> transmute(
  Grid = glue("{nx} x {ny}"), Quadrats = n_quadrats,
  `chi-squared` = fmt_n(round(X2)), df = df,
  VMR = vmr, `p-value` = vapply(p_value, fmt_p, character(1))) |> kable(align = "lrrrrr")
```

My chi-squared statistic exceeds its degrees of freedom by three to four orders of
magnitude at every resolution I tried, and the variance-to-mean ratio runs from
`r min(FO$quad_summary$vmr)` to `r max(FO$quad_summary$vmr)` against a null value of
1. Complete spatial homogeneity is rejected with certainty.

I report this for transparency and it should persuade nobody of anything, including
me. The result was guaranteed by the data-generating process I described in Section
1.1, and also by the obvious geography: a regency that is half drained peatland and
half comparatively intact landscape cannot possibly have uniform fire intensity.
Quadrat counts also inherit an arbitrary origin and cell size, and they throw away
all the within-quadrat information. The only column here I think is genuinely useful
is the VMR, because it tells me *how far* from uniform the pattern is at each scale
rather than answering a yes-or-no question I already knew the answer to.

## Bandwidth selection, and why I rejected cross-validation

Kernel density estimation is the tool that actually does the work here, and the
bandwidth is the decision that determines what it shows me. I spent longer on this
than on any other single choice in the report.

```{r tbl-bw}
#| label: tbl-bw
#| tbl-cap: "Bandwidth selectors compared on the primary pattern, ordered as computed. The spread is a factor of roughly eighteen from smallest to largest, which is why reporting a single surface without this table would conceal the most consequential choice in the first-order analysis."
FO$bw_tab |> transmute(Selector = selector, `sigma (m)` = round(sigma_m),
                       `sigma (km)` = sigma_km) |> kable(align = "lrr")
```

Both cross-validation selectors came back with sub-kilometre bandwidths,
`r round(FO$sigma_ppl)` m for `bw.ppl` and `r round(FO$sigma_diggle)` m for
`bw.diggle`, and `density.ppp` threw a numerical-underflow warning at that scale over
a window of `r fmt_n(round(P$area_km2))` km². I did not adopt either, and it is not
because the warning was inconvenient.

Cross-validation selectors are documented to choose very small bandwidths when a
pattern is strongly clustered at short range. Their criteria reward a surface that
reproduces the local bursts, so they end up treating clustering as fine-grained
intensity structure. In some applications that is exactly what you want. Here it is
the opposite of what I want, because I use the same estimate as the reference
intensity for the second-order test in Section 7.3. If $\hat{\lambda}(u)$ has
already soaked up the sub-kilometre clustering, then asking whether clustering
remains after removing $\hat{\lambda}(u)$ is a rigged question and the answer is no
by construction. I would have produced a confident and meaningless result.

So I fixed the bandwidth at **`r fmt_n(FO$sigma_main)` m**, on a criterion I set
before looking at the surfaces: it has to exceed the interaction range, so that
trend and interaction are estimated by separate parts of my analysis instead of
competing for the same signal. Four kilometres sits above the interaction range
Section 7 goes on to find, inside the interval spanned by the two components of
`bw.scott` (`r fmt_n(round(FO$sigma_scott[1]))` m and
`r fmt_n(round(FO$sigma_scott[2]))` m), and below `bw.CvL`
(`r fmt_n(round(FO$sigma_CvL))` m).

I want to be clear that this is a judgement and not an optimisation. The comparison
below shows what each alternative would have given me, so anyone who disagrees can
see the consequence rather than taking my word for it.

```{r fig-bw}
#| label: fig-bw
#| fig-cap: "The same pattern under four bandwidths. What I take from it: the choice is not cosmetic. The cross-validated surface resolves individual fire complexes and, at this sample size, a great deal of structure no agency could act on; it is also the surface that would sabotage the second-order test, for the reason given above. The 10.5 km bw.CvL surface is stable but has smoothed away the distinction between the peat belt and its margins, which is the distinction of interest. The adopted 4 km surface and bw.scott's 3.8 km are visually close, which is reassurance that the adopted value is not an idiosyncratic pick. What I should disclose is that all four agree on where the principal concentration is and disagree on how many secondary ones exist, so I make no claim about secondary hotspots."
#| fig-height: 5.6
bw_panels <- list(
  ppl   = glue("bw.ppl, {round(FO$sigma_ppl)} m (rejected)"),
  main  = glue("adopted, {fmt_n(FO$sigma_main)} m"),
  scott = glue("bw.scott x, {fmt_n(round(FO$sigma_scott[1]))} m"),
  CvL   = glue("bw.CvL, {fmt_n(round(FO$sigma_CvL))} m"))
bw_df <- bind_rows(lapply(names(bw_panels), function(k)
  im_df(FO$kde[[k]]) |> mutate(panel = bw_panels[[k]]))) |>
  mutate(v = per_1000km2(intensity))
bw_df$panel <- factor(bw_df$panel, levels = unlist(bw_panels))
ggplot(bw_df) +
  geom_raster(aes(x, y, fill = v)) +
  geom_sf(data = win_p, fill = NA, colour = PAL$ink, linewidth = 0.3) +
  facet_wrap(~panel, nrow = 1) +
  scale_fill_intensity(v = bw_df$v) +
  coord_sf(datum = NA, expand = FALSE) +
  labs(title = "What the bandwidth decides") +
  theme_map(base_size = 9)
```

## The intensity surface I report

```{r fig-kde}
#| label: fig-kde
#| fig-width: 6.2
#| fig-cap: "Estimated first-order intensity of burned cells, Gaussian kernel with Diggle's edge correction, at the adopted 4 km bandwidth. What I take from it: intensity is organised as a broad south-southeast band reaching a maximum in the lower third of the regency, with a secondary extension along the eastern margin. The empty northern interior is a real absence rather than an absence of observation: the same satellites observed it on the same days. The gradient is steep, spanning roughly two orders of magnitude between the peat belt and the northern interior, which means resource allocation proportional to intensity would be very unevenly distributed. That is the operational content of a first-order surface, and it is the strongest planning-relevant thing my first-order analysis alone supports."
#| fig-height: 6.6
kde_main <- im_df(FO$kde$main) |> mutate(v = per_1000km2(intensity))
ggplot(kde_main) +
  geom_raster(aes(x, y, fill = v)) +
  geom_sf(data = win_p, fill = NA, colour = PAL$ink, linewidth = 0.45) +
  scale_fill_intensity(v = kde_main$v) +
  coord_sf(datum = NA, expand = FALSE) +
  scalebar_layer(bb) + north_layer(bb) +
  labs(title = "Fire intensity concentrates in the drained southern peat",
       subtitle = glue("Kernel intensity estimate, sigma = {fmt_n(FO$sigma_main)} m, Diggle edge correction")) +
  theme_map() + theme(legend.position = "right", legend.key.height = unit(34, "pt"),
                      legend.key.width = unit(9, "pt"))
```

I also computed an adaptive estimator as a check, because a fixed bandwidth
over-smooths sparse areas and under-smooths dense ones. It reproduces the same
principal concentration, which is the reassurance I was after: the structure I
report is not an artefact of forcing one bandwidth everywhere.

## Radiative power, not just counts

A count surface treats a barely detectable smouldering pixel and an intense flaming
front as the same event, which bothered me. Fire radiative power is the sensor's
estimate of radiative output in megawatts, so smoothing that instead answers a
different and arguably more useful question.

```{r fig-frp}
#| label: fig-frp
#| fig-cap: "Left, mean fire radiative power per detection. Right, total radiative power released, which is intensity and mean power combined. What I take from it: the two panels separate two things a single count map conflates. Total power (right) closely tracks detection intensity, so the southern peat belt dominates energy release as well as event count. Mean power per detection (left) has a different geography: it is comparatively low across the core of the peat belt and higher on the margins and in the sparse north. That is consistent with the physical expectation that peat combustion is low-intensity and persistent while fires on mineral soil and in cleared vegetation burn hotter and shorter. It is also a caution to me: the places with the most detections are not the places with the most intense individual fires."
#| fig-height: 5.8
frp_df  <- im_df(FO$frp_smooth, "v") |> mutate(panel = "Mean FRP per detection (MW)")
frpt_df <- im_df(FO$frp_total, "v") |> mutate(v = v * 1e6, panel = "Total FRP released (MW per km²)")
p1 <- ggplot(frp_df) + geom_raster(aes(x, y, fill = v)) +
  geom_sf(data = win_p, fill = NA, colour = PAL$ink, linewidth = 0.3) +
  scale_fill_gradientn(colours = PAL$seq, name = "MW") +
  coord_sf(datum = NA, expand = FALSE) + labs(subtitle = "Mean FRP per detection") + theme_map(9)
p2 <- ggplot(frpt_df) + geom_raster(aes(x, y, fill = v)) +
  geom_sf(data = win_p, fill = NA, colour = PAL$ink, linewidth = 0.3) +
  scale_fill_gradientn(colours = PAL$seq, trans = "sqrt", name = "MW/km\u00b2",
                       labels = label_number(big.mark = ",")) +
  coord_sf(datum = NA, expand = FALSE) + labs(subtitle = "Total FRP released") + theme_map(9)
p1 + p2 + plot_annotation(title = "Energy release has a different geography from event count",
                          theme = theme_ex01())
```

## Temporal intensity

```{r fig-daily}
#| label: fig-daily
#| fig-cap: "Daily count of burned cells with a centred seven-day mean, and the onset/peak split used later. What I take from it: the episode is not a plateau. Activity is negligible through early July, rises from late July, and escalates sharply through August to a maximum in the first half of September. The seven-day mean is shown because the raw series carries strong short-period structure from cloud cover and overpass geometry, not from fire: a single overcast day suppresses detection regardless of what is burning. I placed the split date by a rule I set in advance, the first day the seven-day mean reaches half its eventual maximum, rather than choosing it after seeing the surfaces I later use it to compare."
#| fig-height: 3.8
dsplit <- ST$split_date
ggplot(FO$daily, aes(date_local)) +
  geom_col(aes(y = n_events), fill = PAL$accent_lt, width = 0.85) +
  geom_line(aes(y = roll7), colour = PAL$accent_dk, linewidth = 0.7) +
  geom_vline(xintercept = as.numeric(dsplit), colour = PAL$cool,
             linetype = "22", linewidth = 0.4) +
  annotate("text", x = dsplit, y = max(FO$daily$n_events) * 0.96,
           label = glue(" onset | peak split, {format(dsplit, '%d %b')}"),
           hjust = 0, size = 2.9, colour = PAL$cool, family = "Helvetica") +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  labs(x = NULL, y = "Burned cells per day",
       title = "A season that escalates, not a steady state",
       subtitle = "Bars, daily count of 375 m cells newly detected. Line, centred seven-day mean.") +
  theme_ex01()
```

```{r fig-overpass}
#| label: fig-overpass
#| fig-cap: "Detections by local solar time of acquisition, split by platform. What I take from it: I put this figure in to stop myself making a mistake. The two platforms observe at four narrow fixed local times, so the apparent day and night counts are not a diurnal cycle of fire activity; they are two sampling windows. Peat fires smoulder continuously and their true diurnal variation cannot be recovered from these data at all. If I said 'fires peak in the early afternoon' I would be describing the satellite's orbit, not the fires. I use the day and night split only as a detectability covariate, since night overpasses catch cooler anomalies against a cooler background."
#| fig-height: 3.2
ggplot(FO$overpass, aes(hr_bin, n, fill = daynight)) +
  geom_col(width = 0.42) +
  facet_wrap(~platform, ncol = 2) +
  scale_fill_manual(values = c(day = PAL$accent, night = PAL$cool), name = NULL) +
  scale_x_continuous(breaks = seq(0, 24, 4), limits = c(0, 24)) +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  labs(x = "Local solar time (WITA, UTC+8)", y = "Detections",
       title = "Four fixed overpasses, not a diurnal cycle") +
  theme_ex01()
```

# Second-order analysis: clustering, inhibition, or randomness

## The straw man I had to test anyway

```{r fig-csr}
#| label: fig-csr
#| fig-cap: "Nearest-neighbour G, empty-space F, and the centred L function, each against a complete spatial randomness envelope from 199 simulations. What I take from it: every statistic departs from the CSR envelope immediately and by an enormous margin. G rises far faster than expected, meaning nearest neighbours are much closer than random; F rises far slower, meaning empty gaps are much larger than random; L sits far above zero at every lag. Read naively this is overwhelming evidence of clustering. Read correctly it is evidence of almost nothing, because the sensor guarantees this result: a single fire produces adjacent detections. I include it to make that failure visible rather than hide it. The analysis I actually rely on is in Section 7.3."
#| fig-height: 3.4
mk_env_df <- function(e, lab) e |> mutate(panel = lab)
eg <- mk_env_df(SO$env_G, "G(r), nearest neighbour")
ef <- mk_env_df(SO$env_F, "F(r), empty space")
el <- SO$env_L |>
  mutate(across(c(obs, theo, lo, hi), ~ .x - r), panel = "L(r) - r, centred")
env_all <- bind_rows(eg, ef, el)
env_all$panel <- factor(env_all$panel, levels = c("G(r), nearest neighbour",
                                                  "F(r), empty space", "L(r) - r, centred"))
ggplot(env_all, aes(r)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = PAL$rule, alpha = 0.75) +
  geom_line(aes(y = theo), colour = PAL$muted, linetype = "22", linewidth = 0.35) +
  geom_line(aes(y = obs), colour = PAL$accent, linewidth = 0.7) +
  facet_wrap(~panel, scales = "free", nrow = 1) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = " km")) +
  labs(x = "r", y = NULL,
       title = "Against complete spatial randomness, everything is significant",
       subtitle = glue("Orange, observed. Grey band, envelope from {SO$nsim} CSR simulations. Dashed, theoretical CSR value.")) +
  theme_ex01(base_size = 9)
```

## What the pair correlation function reveals about the instrument

$K$ and $L$ are cumulative, which means a departure at one lag propagates into every
larger lag and I cannot see where anything is actually happening. The pair
correlation function $g(r)$ is a density rather than a cumulative quantity, so it
separates scales properly. This turned out to matter more than I expected.

```{r fig-pcf}
#| label: fig-pcf
#| fig-cap: "Pair correlation function of the primary pattern, log scale. The dotted line marks the 375 m VIIRS pixel dimension. What I take from it: g(r) is suppressed at very short range and peaks at about 460 m, immediately above the pixel width. This is not fires avoiding one another. It is the sensor's resolution: two detections cannot be much closer than one pixel, because one pixel yields one detection. That short-range suppression is an instrument artefact and is the clearest single piece of evidence that this pattern records observation rather than events. Above the peak, g(r) decays slowly and is still above 3 at 2.9 km, nowhere near its asymptote of 1. So spatial dependence extends well beyond the 5 km range I plot here, which is what pushed me to Section 7.5."
#| fig-height: 3.6
pcf_df <- SO$pcf_obs |> filter(is.finite(trans), r > 0)
ggplot(pcf_df, aes(r, trans)) +
  geom_hline(yintercept = 1, colour = PAL$muted, linetype = "22", linewidth = 0.35) +
  geom_vline(xintercept = DEDUP_GRID_M, colour = PAL$cool, linetype = "12", linewidth = 0.4) +
  geom_line(colour = PAL$accent, linewidth = 0.7) +
  annotate("text", x = DEDUP_GRID_M + 60, y = max(pcf_df$trans) * 0.85,
           label = "375 m pixel", hjust = 0, size = 2.9, colour = PAL$cool,
           family = "Helvetica") +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = " km")) +
  scale_y_log10() +
  labs(x = "r", y = "g(r), log scale",
       title = "The sensor's footprint is visible in the statistic",
       subtitle = "Suppression below one pixel width is an instrument property, not fire behaviour.") +
  theme_ex01()
```

## The test that can fail

These are the statistics I actually trust. My null is an inhomogeneous Poisson
process whose intensity is the kernel estimate from Section 6.5, so everything my
first-order analysis established is already inside the null. If I get a departure
now, it means residual interaction rather than large-scale variation, which is the
question I set in Q3.

```{r fig-inhom}
#| label: fig-inhom
#| fig-cap: "Inhomogeneous centred L function and inhomogeneous pair correlation function, against envelopes from 199 simulations of an inhomogeneous Poisson process. The reference intensity is re-estimated for every simulated pattern. What I take from it: the departure survives, and it does not close. Both curves leave the envelope at about 100 m and remain outside it across the entire range plotted. g_inhom peaks near 5.9 at roughly 570 m and is still above 2 at 2.8 km, against an envelope ceiling near 1.4. So after accounting for the fact that fire is far more likely in the southern peat belt than in the north, burned cells still occur much closer together than that variation alone predicts. Note also that g_inhom falls below 1 at the shortest lags, the sensor floor surviving into the inhomogeneous statistic."
#| fig-height: 3.6
# An envelope built from `simulate=` has no theoretical curve, only the simulation
# mean, so the reference is set explicitly: 0 for the centred L function and 1 for
# the pair correlation function.
li <- SO$env_Linhom |> mutate(across(c(obs, lo, hi), ~ .x - r),
                              ref = 0, panel = "L_inhom(r) - r")
gi <- SO$env_pcfinhom |> filter(r > 0) |> mutate(ref = 1, panel = "g_inhom(r)")
ggplot(bind_rows(li, gi), aes(r)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = PAL$rule, alpha = 0.75) +
  geom_line(aes(y = ref), colour = PAL$muted, linetype = "22", linewidth = 0.35) +
  geom_line(aes(y = obs), colour = PAL$accent, linewidth = 0.7) +
  facet_wrap(~panel, scales = "free", nrow = 1) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = " km")) +
  labs(x = "r", y = NULL,
       title = "Clustering survives the honest null, and has a scale",
       subtitle = glue("Envelope from {SO$nsim} simulations of an inhomogeneous Poisson process with lambda re-estimated each time.")) +
  theme_ex01(base_size = 9)
```

```{r tbl-global}
#| label: tbl-global
#| tbl-cap: "Global Monte Carlo tests. The maximum absolute deviation (MAD) and Diggle-Cressie-Loosmore-Ford (DCLF) statistics each summarise the whole curve in one number, so these p-values are not inflated by reading a band at many lags. The Monte Carlo floor is 0.01 with 99 simulations."
tibble::tribble(
  ~Test, ~Statistic, ~Null, ~`p-value`,
  "MAD",  "L(r)",       "complete spatial randomness",  fmt_p(SO$mad_L$p.value, 99),
  "MAD",  "L_inhom(r)", "inhomogeneous Poisson",        fmt_p(SO$mad_Linhom$p.value, 99),
  "DCLF", "L_inhom(r)", "inhomogeneous Poisson",        fmt_p(SO$dclf_Linhom$p.value, 99)
) |> kable(align = "lllr")
```

Both global tests reject the inhomogeneous Poisson null at the Monte Carlo floor. So
the pattern is clustered **beyond** its first-order trend, which is the answer to the
first half of Q3.

What I could not establish from this was where the clustering ends, and that is down
to a decision of mine that turned out to be wrong. I capped the r range at 5 km,
reasoning that longer lags in an elongated window rest on too few pairs. But across
the whole of that range both curves sit outside the envelope, which means 5 km only
establishes that dependence extends *at least* that far. My cap was too conservative
to answer the question I had set it to answer. Section 7.5 is where I fix it.

The direction of the remaining bias matters for how confidently this can be
stated. As set out in assumption A6, estimating $\hat{\lambda}$ from the pattern
under test lets it absorb part of the clustering, which widens the envelope and
makes the test *harder* to reject. The rejection therefore stands despite a
conservative procedure, not because of a permissive one.

## Sensitivity to the event definition

My whole analysis rests on collapsing detections to cells, so I wanted to know
whether the finding depends on that choice. If it did, it would be a finding about
my choice rather than about fire.

```{r fig-sens}
#| label: fig-sens
#| fig-cap: "Inhomogeneous centred L function computed on the cell pattern and on the un-thinned detection pattern, on a common axis. What I take from it: both curves are positive, so the qualitative conclusion of residual clustering does not depend on the event definition. The magnitudes differ by roughly an order of magnitude, and the un-thinned pattern shows far stronger apparent clustering at every lag. That gap is the repeat-detection artefact measured directly. It is why I use the cell pattern for inference and the detection pattern only for this check. If I reported the upper curve I would be reporting the satellite revisit schedule as a fire-behaviour result."
#| fig-height: 3.4
sd_df <- bind_rows(
  SO$Linhom_cell |> transmute(r, v = border - r, pattern = "ppp_cell (reported)"),
  SO$Linhom_all  |> transmute(r, v = border - r, pattern = "ppp_all (un-thinned)"))
ggplot(sd_df, aes(r, v, colour = pattern)) +
  geom_hline(yintercept = 0, colour = PAL$muted, linetype = "22", linewidth = 0.35) +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c("ppp_cell (reported)" = PAL$accent,
                                 "ppp_all (un-thinned)" = PAL$cool), name = NULL) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = " km")) +
  labs(x = "r", y = "L_inhom(r) - r",
       title = "The conclusion is robust, the magnitude is not",
       subtitle = glue("n = {fmt_n(SO$n_cell)} cells against n = {fmt_n(SO$n_all)} detections.")) +
  theme_ex01()
```

## Extending the range, after getting it wrong

I set the 5 km cap because long lags in an elongated window rest on few pairs. That
reasoning is fine, but 5 km was more conservative than I needed. The rule I stated in
Section 5 caps $r$ at a quarter of the window's shortest dimension, and the window is
`r round((bb["xmax"]-bb["xmin"])/1000)` km wide, so 20 km was admissible the whole
time. Extending to 20 km is not me granting myself a post-hoc licence, it is me
finally applying the rule I had already written down.

Two practical constraints came with that, and I am disclosing both rather than
hiding them. `pcfinhom` offers no border correction, and the translation correction
over roughly seven million pairs per pattern would not finish in a workable time for
99 simulated patterns. So my significance across the extended range rests on
$L_{inhom}$ with a border correction and a full envelope, while the shape of
$g_{inhom}$ comes from a single translation-corrected curve with no simulations.
**I make no significance claim for $g_{inhom}$ beyond 3 km.**

```{r fig-ext}
#| label: fig-ext
#| fig-cap: "Left, inhomogeneous centred L function to 20 km with a 99-simulation envelope. Right, the observed inhomogeneous pair correlation function over the same range, with no envelope, on a log scale. What I take from it: the two panels answer different questions and disagree in an instructive way. g_inhom, which localises scale, peaks at about 440 m and decays steadily: it is 4.3 at 1 km, 2.7 at 2 km, 1.4 at 5 km, and reaches within 10% of its asymptote of 1 at about 7.2 km. L_inhom, which is cumulative, never re-enters its envelope within 20 km. Both are correct. The excess pair density is a local phenomenon that has largely dissipated by 7 km, and because L accumulates every excess at smaller lags it remains significantly elevated far beyond the range over which anything is actually happening."
#| fig-height: 3.6
Lx <- EX$env_L_ext |> mutate(across(c(obs, lo, hi), ~ .x - r), ref = 0,
                             panel = "L_inhom(r) - r, with envelope")
Gx <- EX$g_obs_ext |> filter(r > 0, is.finite(trans)) |>
  transmute(r, obs = trans, lo = NA_real_, hi = NA_real_, ref = 1,
            panel = "g_inhom(r), observed only")
ggplot(bind_rows(Lx, Gx), aes(r)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = PAL$rule, alpha = 0.75, na.rm = TRUE) +
  geom_line(aes(y = ref), colour = PAL$muted, linetype = "22", linewidth = 0.35) +
  geom_line(aes(y = obs), colour = PAL$accent, linewidth = 0.7) +
  geom_vline(data = tibble(panel = "g_inhom(r), observed only",
                           x = EX$summary_ext$r_g_within_10pct_of_1),
             aes(xintercept = x), colour = PAL$cool, linetype = "12", linewidth = 0.4) +
  facet_wrap(~panel, scales = "free", nrow = 1) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = " km")) +
  labs(x = "r", y = NULL,
       title = "Dependence is local, but a cumulative statistic cannot say so",
       subtitle = glue("Dotted line marks {round(EX$summary_ext$r_g_within_10pct_of_1/1000, 1)} km, where g_inhom first comes within 10% of 1.")) +
  theme_ex01(base_size = 9)
```

```{r tbl-ext}
#| label: tbl-ext
#| tbl-cap: "Observed inhomogeneous pair correlation at selected lags. A value of 1 means no excess over the inhomogeneous Poisson expectation."
with(EX$summary_ext, tibble::tibble(
  Lag = c("peak", "1 km", "2 km", "5 km", "10 km", "20 km"),
  `g_inhom` = round(c(peak_g, g_1km, g_2km, g_5km, g_10km, g_20km), 2),
  Note = c(glue("at {round(peak_r_m)} m"), "", "", "", "", "sampling noise at this lag")
)) |> kable(align = "lrl")
```

```{r}
#| label: ext-crosscheck
#| include: false
cc <- EX$correction_cmp
cc_med <- 100 * median(abs(cc$rel_diff), na.rm = TRUE)
```

I cross-checked the extended curve against my Section 7.3 estimate over the lags
both cover, since both use the translation correction. The median relative difference
is `r sprintf("%.2f%%", cc_med)`, so changing the $r$ grid has not quietly changed
the estimate underneath me.

**So here is my corrected answer to Q3.** Residual clustering begins immediately
above the sensor floor, is strongest at roughly
`r round(EX$summary_ext$peak_r_m)` m, and only decays to within 10% of no excess by
about `r round(EX$summary_ext$r_g_within_10pct_of_1/1000, 1)` km. My earlier
statement that the excess was exhausted by 2 km was simply an artefact of not having
looked far enough, and the conclusion I drew from it, that the first-order surface
was a sufficient description above 2 km, was wrong. It is not sufficient at any lag
below roughly 7 km.

# Spatio-temporal analysis

## Space-time interaction

```{r tbl-knox}
#| label: tbl-knox
#| tbl-cap: "Knox test of space-time interaction across a grid of spatial and temporal thresholds. Each cell counts observed pairs close in both space and time, against the mean under 999 random labellings in which locations and dates are held fixed and only their pairing is permuted. The ratio is observed over expected."
ST$knox_tab |> transmute(
  `Space (m)` = fmt_n(s_m), `Time (days)` = t_days,
  Observed = fmt_n(observed), `Expected` = fmt_n(round(exp_perm)),
  Ratio = round(ratio, 2),
  `p-value` = vapply(p_value, fmt_p, character(1), nperm = ST$nperm)) |>
  kable(align = "rrrrrr")
```

```{r fig-knox}
#| label: fig-knox
#| fig-cap: "Knox ratio of observed to expected close pairs, by spatial and temporal threshold. What I take from it: interaction is strongest in the tightest cell, pairs within a few hundred metres and one day, and decays monotonically as either threshold widens. That shape is the signature of local, short-lived propagation: a cell that burns today makes a neighbouring cell more likely to burn tomorrow, and the effect is largely spent within a few days and a kilometre or two. The decay in the time direction is steeper than in the space direction, which is consistent with fires spreading over contiguous ground within days rather than recurring in the same place weeks apart. Because the null holds both marginals fixed, none of this can be attributed to the seasonal escalation shown in Figure 8."
#| fig-height: 3.6
ggplot(ST$knox_tab, aes(factor(s_m), factor(t_days), fill = ratio)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = sprintf("%.2f", ratio)), size = 3.1, colour = PAL$ink,
            family = "Helvetica") +
  scale_fill_gradientn(colours = PAL$seq, name = "Observed / expected") +
  labs(x = "Spatial threshold (m)", y = "Temporal threshold (days)",
       title = "Interaction is local in space and short in time",
       subtitle = glue("{fmt_n(ST$n_close_pairs)} close pairs, {ST$nperm} random labellings.")) +
  theme_ex01(grid = "none")
```

## Does the signature move between onset and peak?

```{r fig-rr}
#| label: fig-rr
#| fig-cap: "Probability that a burned cell belongs to the peak period rather than the onset period, as a spatially varying surface. Values above the overall peak share indicate areas where fire arrived late; values below indicate areas that burned early. What I take from it: the surface is not flat, so the episode is not a uniform intensification of one fixed footprint. The southern and southeastern peat burned disproportionately early, while the northern and western margins of the active area contribute disproportionately to the peak period. Read together with the trajectory in the next figure, this describes an episode that begins in the most degraded peat and expands outward from it. The operational implication I draw is that where fire is early in the season carries information about where it will appear later, which my static seasonal intensity map cannot convey."
#| fig-height: 6.2
rr_df <- im_df(ST$rr, "p")
share_peak <- as.numeric(ST$season_counts["peak"]) / sum(ST$season_counts)
ggplot(rr_df) +
  geom_raster(aes(x, y, fill = p)) +
  geom_sf(data = win_p, fill = NA, colour = PAL$ink, linewidth = 0.45) +
  scale_fill_gradient2(low = PAL$cool, mid = PAL$ground_alt, high = PAL$accent_dk,
                       midpoint = share_peak, name = "P(peak period)",
                       labels = label_percent(accuracy = 1)) +
  coord_sf(datum = NA, expand = FALSE) +
  scalebar_layer(bb) + north_layer(bb) +
  labs(title = "The episode expands rather than intensifies in place",
       subtitle = glue("Split at {format(ST$split_date, '%d %B')}. Midpoint of the scale is the overall peak share, {percent(share_peak, accuracy = 1)}.")) +
  theme_map()
```

```{r fig-traj}
#| label: fig-traj
#| fig-cap: "Weekly mean centre of burned cells with the weekly dispersion about it, and the distance the centre moved each week. What I take from it: the mean centre is a crude summary and I use it only for what it does reliably, which is showing direction and spread over time. Dispersion widens as the season progresses, which confirms the expansion I read off the relative-risk surface. The weekly displacement of the centre is largest in the early weeks, when few cells dominate the mean, and settles to a few kilometres a week once activity is widespread. The caution I would attach is that the mean centre of a strongly non-convex, multi-modal pattern is not a place where anything is happening. It is a summary statistic and I would not put it on an operational map."
#| fig-height: 3.6
tr <- ST$traj |> filter(n > 0)
p_sd <- ggplot(tr, aes(week_start, sd_km)) +
  geom_col(fill = PAL$accent_lt, width = 4.4) +
  labs(x = NULL, y = "Standard distance (km)", subtitle = "Dispersion of burned cells") +
  scale_x_date(date_labels = "%d %b") + theme_ex01()
p_step <- ggplot(tr |> filter(!is.na(step_km)), aes(week_start, step_km)) +
  geom_line(colour = PAL$accent_dk, linewidth = 0.6) +
  geom_point(colour = PAL$accent_dk, size = 1.2) +
  labs(x = NULL, y = "Weekly shift of mean centre (km)", subtitle = "Movement of the mean centre") +
  scale_x_date(date_labels = "%d %b") + theme_ex01()
p_sd + p_step + plot_annotation(
  title = "The footprint widens as the season progresses", theme = theme_ex01())
```

## Space-time intensity

```{r fig-st}
#| label: fig-st
#| fig-cap: "Space-time intensity at fortnightly reference days, estimated with a fixed spatial kernel and a Gaussian weight in time of five days' standard deviation. Each panel is normalised to its own weights, so panels show where fire was concentrated at that moment rather than how much there was. What I take from it: the sequence makes the expansion explicit. Early July activity is confined to a small southeastern area. Through August the active area extends north and west along the peat belt while retaining the southern core. By mid-September activity covers most of the southern two-thirds of the regency. My estimator assumes separability, meaning the spatial kernel does not change shape over time. That is a simplification: the real footprint is elongating, and a separable estimator can only show that as a change in weights."
#| fig-height: 4.4
st_df <- bind_rows(lapply(ST$st_kde, function(s)
  im_df(s$im) |> mutate(panel = format(s$date, "%d %b"))))
st_df$panel <- factor(st_df$panel, levels = vapply(ST$st_kde,
                                                   function(s) format(s$date, "%d %b"), character(1)))
ggplot(st_df) +
  geom_raster(aes(x, y, fill = intensity / max(intensity, na.rm = TRUE))) +
  geom_sf(data = win_p, fill = NA, colour = PAL$ink, linewidth = 0.22) +
  facet_wrap(~panel, nrow = 1) +
  scale_fill_gradientn(colours = PAL$seq, name = "Relative intensity",
                       labels = label_percent(accuracy = 1)) +
  coord_sf(datum = NA, expand = FALSE) +
  labs(title = "Fire spreads north and west from a southeastern origin") +
  theme_map(base_size = 8)
```

# Value-added analysis

So far I have established that fire is clustered, where it concentrates, and over
what distance and time the interaction operates. What I still cannot say is what the
clustering is organised around, or how big a cluster actually is. Both are
answerable, and both are what turn a statistical result into something an agency
could act on, so I went after them.

## What is the clustering organised around?

Peat fire in this landscape is overwhelmingly human in ignition and made possible by
drainage. I picked two mapped covariates to proxy those mechanisms: distance to the
road network, for access and ignition opportunity, and distance to mapped canals and
waterways, for drainage-driven desiccation.

I took the covariate geometry from a dated Geofabrik OpenStreetMap extract for
Kalimantan rather than a live Overpass query. That was not my first choice. Overpass
refused a bounding box this size repeatedly, and once I had switched I realised the
extract is better anyway: a live query returns different data every time it runs, so
it cannot be part of a reproducible pipeline. The extract gave me
`r fmt_n(CM$n_roads)` road features and `r fmt_n(CM$n_water)` waterway features
within a 15 km buffer of the regency.

I compute distances on that buffered frame and only then crop to the analysis
window. If I had computed them on the clipped window I would have biased every
distance upward for cells near the boundary, because their nearest road is often
outside the regency entirely.

The `r fmt_n(CM$n_water)` waterway features against `r fmt_n(CM$n_roads)` road
features is itself my first result here, and it is a warning rather than a finding.
An area defined by an abandoned canal network has fewer mapped watercourses than
roads. I come back to what that does to the drainage covariate below.

```{r fig-rho}
#| label: fig-rho
#| fig-cap: "Estimated intensity as a non-parametric function of distance to the nearest road and to the nearest mapped waterway, with 95% confidence bands. The horizontal line is the mean intensity over the window. What I take from it: intensity falls steeply with distance from a road, from well above the window mean within the first kilometre to below it beyond a few kilometres. That is a strong and monotone access effect, consistent with ignition being human. The waterway relationship is weaker and less monotone. This is my more interesting negative result. It does not show drainage is unimportant, it shows OpenStreetMap's waterway layer is a poor proxy for it, because the dense minor canal network of the former Mega Rice Project is largely unmapped. My covariate is measuring mapping effort as much as hydrology, so I read the flat curve as absence of evidence."
#| fig-height: 3.6
rho_to_df <- function(rh, lab) {
  if (is.null(rh)) return(NULL)
  d <- rh
  nm <- names(d)
  tibble(x = d[[1]], rho = d[["rho"]],
         lo = if ("lo" %in% nm) d[["lo"]] else NA_real_,
         hi = if ("hi" %in% nm) d[["hi"]] else NA_real_, panel = lab)
}
rho_df <- bind_rows(
  rho_to_df(CM$rho_road,  "Distance to nearest road"),
  rho_to_df(CM$rho_water, "Distance to nearest mapped waterway"))
mean_int <- P$ppp_cell$n / (P$area_km2 * 1e6)
if (!is.null(rho_df) && nrow(rho_df)) {
  ggplot(rho_df, aes(x / 1000)) +
    geom_hline(yintercept = per_1000km2(mean_int) / 1e3 * 1e3,
               colour = PAL$muted, linetype = "22", linewidth = 0.35) +
    geom_ribbon(aes(ymin = per_1000km2(lo), ymax = per_1000km2(hi)),
                fill = PAL$rule, alpha = 0.6) +
    geom_line(aes(y = per_1000km2(rho)), colour = PAL$accent, linewidth = 0.7) +
    facet_wrap(~panel, scales = "free", nrow = 1) +
    labs(x = "Distance (km)", y = "Burned cells per 1,000 km²",
         title = "Access explains a great deal, mapped drainage very little",
         subtitle = "Dashed line, mean intensity over the window.") +
    theme_ex01(base_size = 9)
} else cat("Covariate estimation unavailable; see Section 11.\n")
```

```{r tbl-aic}
#| label: tbl-aic
#| tbl-cap: "Fitted Poisson point-process models compared by AIC. Distances are in kilometres, so coefficients read per kilometre. The log specification was included on the expectation that a steep near-field effect flattening with distance would fit better than a linear term; it does not, and is reported as a failed expectation rather than dropped."
CM$aic_tab |> transmute(Model = model, Parameters = npar,
                        AIC = fmt_n(round(AIC)), `Delta AIC` = fmt_n(round(dAIC))) |>
  kable(align = "lrrr")
```

```{r}
#| label: aic-inline
#| include: false
dn <- CM$aic_tab$dAIC[CM$aic_tab$model == "null"]
dr <- dn - CM$aic_tab$dAIC[CM$aic_tab$model == "road"]
dw <- dn - CM$aic_tab$dAIC[CM$aic_tab$model == "water"]
dboth_over_road <- CM$aic_tab$dAIC[CM$aic_tab$model == "road"]
```

Three things in that table I want to separate out.

**Distance to road does nearly all the work.** My full model improves on the
intercept-only model by `r fmt_n(round(dn))` AIC units, and distance to road alone
recovers `r fmt_n(round(dr))` of them. Adding distance to water to the road model
buys a further `r round(dboth_over_road, 1)` units, which on two extra parameters is
real but very small. Distance to water on its own recovers `r fmt_n(round(dw))`
units, so it is not uninformative by itself. It is mostly redundant once I know road
distance, which is what I would expect if canals and tracks follow each other
through the landscape.

**My log specification fitted worse, not better.** I included it expecting the
near-field effect visible in the rhohat curve to be better captured by a log term
than a linear one. It loses to the linear model on the same number of parameters, by
`r fmt_n(round(CM$aic_tab$dAIC[CM$aic_tab$model == "log"]))` AIC units. Over the
range of distances that actually occurs in this window, the relationship is closer to
linear in distance than in its logarithm. I was wrong, and I am reporting it rather
than dropping the model quietly.

**None of this is causal and I want to say so plainly.** Roads get built where
people already work the land, so distance to road is tangled up with land use,
tenure and drainage in ways my data cannot separate. The association is strong. The
mechanism is not identified.

```{r fig-resk}
#| label: fig-resk
#| fig-cap: "Residual K function for the best-fitting Poisson model. If the fitted covariate trend fully explained the pattern, this curve would sit on zero within its envelope. What I take from it: it does not. A substantial positive residual remains at short lags, which means the covariates account for part of the first-order variation but none of the local interaction found in Section 7.3. Substantively: knowing how far a place is from a road tells me whether it is likely to burn. It does not tell me that burning is contagious at the kilometre scale. A Poisson model cannot represent contagion by construction, which is the motivation for the cluster process fitted next."
#| fig-height: 3.4
if (!is.null(CM$res_K)) {
  rk <- CM$res_K
  ycol <- if ("bres" %in% names(rk)) "bres" else names(rk)[2]
  ggplot(rk, aes(r, .data[[ycol]])) +
    geom_hline(yintercept = 0, colour = PAL$muted, linetype = "22", linewidth = 0.35) +
    geom_line(colour = PAL$accent, linewidth = 0.7) +
    scale_x_continuous(labels = label_number(scale = 1e-3, suffix = " km")) +
    labs(x = "r", y = "Residual K",
         title = "The covariates explain location, not contagion") +
    theme_ex01()
} else cat("Residual K unavailable.\n")
```

## How large is a cluster?

A significance test tells me clustering exists. A fitted cluster process tells me how
big it is, which is the form a planner could actually use. Getting that number right
meant throwing away my first answer.

I first fitted a Thomas process with no trend term. It returned a cluster standard
deviation of `r fmt_n(round(TH$kppm_hom$sigma_m))` m, about
`r round(TH$kppm_hom$mu_per_cluster)` events per cluster, and an implied
`r round(TH$kppm_hom$kappa_per_km2 * P$area_km2)` parent locations in the entire
regency. Twenty-three clusters of 440 events each across a whole regency did not
sound like a cluster scale to me, and it is not. It is the southern peat belt being
re-described as a handful of enormous clusters, which is the only way a model with
no trend term can represent large-scale intensity variation. This is the standard
identifiability problem between trend and interaction in cluster processes. If I had
taken the number at face value I would have reported a confident and completely
wrong operational radius.

The fix is to remove the trend before estimating the cluster parameters. My
inhomogeneous pair correlation function already does exactly that, because it divides
out $\hat{\lambda}(u)$. So I fitted the Thomas pair correlation function

$$g(r) \;=\; 1 + \frac{1}{4\pi\kappa\sigma^{2}}\exp\!\left(-\frac{r^{2}}{4\sigma^{2}}\right)$$

directly to $g_{inhom}$ by minimum contrast, minimising
$\sum_r \left(\hat{g}(r)^{q} - g_\theta(r)^{q}\right)^{2}$ with $q = 1/4$ over
$r \in [`r TH$RMIN`, `r fmt_n(TH$RMAX)`]$ m. I set the lower limit to exclude lags
governed by the 375 m sensor floor rather than by the cluster process. I wrote the
fit out explicitly in `R/07_cluster_scale.R` rather than calling a packaged routine,
because the objective function and the excluded range are the substance of the
estimate and I wanted them visible.

```{r tbl-thomas}
#| label: tbl-thomas
#| tbl-cap: "Thomas cluster process fitted by minimum contrast, with and without the first-order trend removed. sigma is the standard deviation of offspring displacement from the cluster centre and r95 the radius containing about 95% of a cluster's events. The contrast between the two rows is the finding: leaving the trend in the model inflates the estimated cluster scale by a factor of two, and the implied cluster sizes become absurd."
TH$thomas_tab |> transmute(
  Specification = spec,
  `sigma (m)` = round(sigma_m),
  `r95 (m)` = round(r95_m),
  `Parents in regency` = round(n_parents_implied),
  `Events per cluster` = round(mu_per_cluster)) |> kable(align = "lrrrr")
```

```{r}
#| label: thomas-inline
#| include: false
tp <- TH$thomas_tab |> filter(key == "trend_removed")
r95 <- round(tp$r95_m); sig_t <- round(tp$sigma_m); mu_t <- round(tp$mu_per_cluster)
```

```{r fig-thomas}
#| label: fig-thomas
#| fig-cap: "Observed inhomogeneous pair correlation function with the two fitted Thomas models, log scale, over the range used for fitting. The shaded band marks lags below 200 m, excluded from the fit because they are governed by the sensor's pixel size. What I take from it: the trend-removed fit tracks the observed curve closely from the peak outward, which is the range that matters for an operational radius. The fit that retains the trend is too flat and too wide, because its cluster component is doing work that belongs to the intensity surface. Neither fit captures the observed curve beyond about 4 km, where the observed g_inhom decays more slowly than a single Gaussian cluster kernel allows. That is a real limitation of my model: the pattern has structure at more than one scale, and a one-scale model cannot represent it."
#| fig-height: 3.6
gobs <- EX$g_obs_ext |> filter(r > 0, r <= TH$RMAX, is.finite(trans))
ggplot() +
  annotate("rect", xmin = 0, xmax = TH$RMIN, ymin = 1, ymax = max(gobs$trans) * 1.05,
           fill = PAL$rule, alpha = 0.45) +
  geom_hline(yintercept = 1, colour = PAL$muted, linetype = "22", linewidth = 0.35) +
  geom_line(data = gobs, aes(r, trans), colour = PAL$accent, linewidth = 0.8) +
  geom_line(data = TH$curves, aes(r, g, linetype = spec), colour = PAL$cool,
            linewidth = 0.55) +
  scale_linetype_manual(values = c(2, 1), name = NULL) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = " km")) +
  scale_y_log10() +
  labs(x = "r", y = "g(r), log scale",
       title = "Removing the trend halves the estimated cluster scale",
       subtitle = "Orange, observed inhomogeneous pcf. Blue, fitted Thomas models. Shaded, lags excluded from the fit.") +
  theme_ex01()
```

My trend-removed fit gives a cluster standard deviation of `r fmt_n(sig_t)` m, so
about 95% of the events in a cluster fall within roughly `r fmt_n(r95)` m of its
centre, with `r mu_t` events per cluster and about
`r round(tp$n_parents_implied)` cluster centres across the regency. That is
consistent with my pair correlation function, which peaks at
`r round(EX$summary_ext$peak_r_m)` m, and I got there by a different route: minimum
contrast fitting of a mechanistic model rather than envelope testing.

Two estimates agreeing is stronger than either alone, but the agreement is only
partial and I should say where it breaks. My fitted model has a single scale, and
the figure below shows the observed curve decaying more slowly beyond about 4 km
than a single Gaussian kernel allows. So the Thomas process describes the dominant
local structure well and the longer tail of dependence out to roughly
`r round(EX$summary_ext$r_g_within_10pct_of_1/1000, 1)` km not at all. A multi-scale
cluster process, or a log-Gaussian Cox process with a flexible covariance, would be
the natural next model. Fitting one is beyond what I can do in this exercise, but it
is the first thing I would try next.

## Where is the single most unusual concentration?

This is the most directly practical question I ask anywhere in the report. Not
whether clustering exists, but where the single most unusual concentration sits and
whether it is more extreme than chance would produce.

```{r tbl-scan}
#| label: tbl-scan
#| tbl-cap: "Spatial scan statistic, Poisson model, circular windows of 2, 5 and 10 km radius, 99 Monte Carlo simulations. The location is the maximum of the likelihood-ratio surface."
sc <- CM$scan_res
lrts <- attr(sc, "obsLRTS")
w  <- which(lrts$v == max(lrts$v, na.rm = TRUE), arr.ind = TRUE)[1, ]
mx <- c(x = lrts$xcol[w[2]], y = lrts$yrow[w[1]])
ll <- st_coordinates(st_transform(
  st_sfc(st_point(c(mx["x"], mx["y"])), crs = STUDY_CRS), 4326))
tibble::tibble(
  `Max likelihood-ratio statistic` = round(as.numeric(sc$statistic), 1),
  `p-value` = fmt_p(sc$p.value, 99),
  `Radii tested (km)` = paste(attr(sc, "r") / 1000, collapse = ", "),
  `Most likely cluster centre` = sprintf("%.3f E, %.3f S", ll[1], abs(ll[2]))
) |> kable(align = "rrll")
```

```{r fig-scan}
#| label: fig-scan
#| fig-cap: "Likelihood-ratio surface from the spatial scan statistic, with the maximum marked. What I take from it: the surface is not a second intensity map. It measures, at each location, how much better a model with an elevated rate inside a circle fits than a model with a constant rate, so it answers where I would look first rather than where fire is densest. The maximum sits in the southern peat belt, consistent with the kernel intensity estimate, and the statistic is significant at the Monte Carlo floor. What this adds over my intensity map is that it is explicitly comparative. The broad plateau shows several locations are nearly as unusual as the maximum, so I would over-read it if I treated the single peak as uniquely important."
#| fig-height: 6.2
ls_df <- im_df(lrts, "lrts")
ggplot(ls_df) +
  geom_raster(aes(x, y, fill = lrts)) +
  geom_sf(data = win_p, fill = NA, colour = PAL$ink, linewidth = 0.45) +
  annotate("point", x = mx["x"], y = mx["y"], shape = 21, size = 3.4,
           colour = PAL$ink, fill = NA, stroke = 0.8) +
  annotate("text", x = mx["x"] + 4000, y = mx["y"], label = "maximum",
           hjust = 0, size = 3, colour = PAL$ink, family = "Helvetica") +
  scale_fill_gradientn(colours = PAL$seq, name = "Likelihood ratio statistic",
                       labels = label_number(big.mark = ",")) +
  coord_sf(datum = NA, expand = FALSE) + scalebar_layer(bb) + north_layer(bb) +
  labs(title = "Where an agency would look first",
       subtitle = glue("Scan statistic surface, circles of {paste(attr(sc,'r')/1000, collapse=', ')} km.")) +
  theme_map()
```

# Integrated findings and discussion

Here are my four questions from Section 2, answered together, with how confident I
am in each.

```{r tbl-integrate}
#| label: tbl-integrate
#| tbl-cap: "What I found, the evidence behind each finding, and how confidently I hold it."
tibble::tribble(
  ~Finding, ~Evidence, ~Confidence,
  "Fire intensity varies by roughly two orders of magnitude across the regency, concentrated in the drained southern peat",
  "KDE, stable across four bandwidths and an adaptive estimator", "High",
  "The pattern is clustered beyond its first-order trend at every lag tested below about 7 km",
  "L_inhom and g_inhom against an inhomogeneous Poisson null, MAD and DCLF global tests at the Monte Carlo floor", "High",
  "The excess is strongest at roughly 440 m and decays to within 10% of no-excess by about 7 km",
  "Observed g_inhom to 20 km; note that no envelope was computed beyond 3 km", "Moderate",
  "Apparent inhibition at the shortest lags is an instrument artefact, not fire behaviour",
  "g(r) and g_inhom(r) both suppressed below the 375 m pixel dimension", "High",
  "The dominant cluster scale is about 1.1 km, with 95% of a cluster within about 2.6 km",
  "Thomas minimum-contrast fit to g_inhom, consistent with the g_inhom peak; but a one-scale model does not fit the tail", "Moderate",
  "Space-time interaction is local and short-lived, strongest within a few hundred metres and one day",
  "Knox test under random labelling, 999 permutations, monotone decay across the threshold grid", "Moderate to high",
  "The episode expands outward from a southeastern origin rather than intensifying in place",
  "Relative-risk surface, weekly dispersion, space-time intensity sequence", "Moderate",
  "Distance to road is strongly associated with intensity; mapped drainage is not",
  "rhohat and AIC comparison of fitted Poisson models", "Moderate for roads, uninformative for drainage",
  "The observed temporal profile is fire activity, but the day/night contrast is sampling",
  "Fixed local overpass times", "High"
) |> kable()
```

## Consistency across approaches

My first-order and second-order analyses agree where they should and disagree in two
places, and the disagreements taught me more than the agreement did.

**They agree on the dominant scale.** My inhomogeneous pair correlation function
peaks at `r round(EX$summary_ext$peak_r_m)` m. My Thomas process fitted to that same
curve returns a cluster standard deviation of `r fmt_n(sig_t)` m. Two estimates from
different machinery, one envelope-based and one model-based, both put the dominant
structure at roughly a kilometre. That agreement is the main reason I am willing to
quote a number.

**They disagree about extent, and it is the cumulative statistic that misled me.**
$L_{inhom}$ never re-enters its envelope within 20 km, which tempted me to conclude
that dependence extends at least that far. $g_{inhom}$, which localises scale
instead of accumulating it, shows the excess is within 10% of nothing by about
`r round(EX$summary_ext$r_g_within_10pct_of_1/1000, 1)` km. Both computations are
correct and the second is the one I report, because $L$ accumulates every excess at
smaller lags and therefore stays elevated long past the range over which anything is
actually happening. This is a general property of $K$ and $L$ rather than a quirk of
my data, and it is probably the single most useful thing I learned doing this
exercise.

**They disagree about sufficiency, and my first-order surface loses.** My earlier
version, capped at a 5 km lag, concluded that the intensity surface was an adequate
description above about 2 km. That was wrong, and wrong because of a decision rather
than a computation: my cap was too short to see where dependence ended. Corrected,
the first-order surface is not a sufficient description at any lag below roughly
`r round(EX$summary_ext$r_g_within_10pct_of_1/1000, 1)` km. A smooth intensity map is
still a reasonable tool for allocating resources across a regency. It is not a tool
for anticipating what happens next around a fire that has already started.

My covariate models sharpen this rather than softening it. They improve on an
intercept-only model by `r fmt_n(round(dn))` AIC units, so a lot of the first-order
variation is explained by accessibility. But the residual $K$ function shows they
capture none of the local interaction. Location explains where fire is likely. It
does not explain the contagion.

## Where my analysis is weakest

Four places, roughly in order of how much they worry me.

**My event definition does the heavy lifting.** Collapsing detections to cells is
defensible and I think necessary, but it is a modelling decision that I have
presented as data preparation, and I should be honest that those are not the same
thing. A cell that burned for three weeks and a cell that burned for one overpass
contribute identically to `ppp_cell`. If I weighted cells by persistence or by
cumulative radiative power I would be answering a related but different question,
and I am not certain I would reach the same conclusion about scale.

**My inhomogeneous null is estimated from the data I am testing.** The bias runs in
the conservative direction, so I am comfortable that the rejection stands. But the
*magnitude* of the excess clustering is not reliably estimated, so anything I say
about how much clustering there is should be read as ordinal rather than
quantitative.

**Detectability is confounded with intensity.** I cannot discharge assumptions A1
and A2 within this data source. Sub-surface peat smouldering, the mode I care most
about for public health, is systematically under-detected, and cloud cover is not
plausibly independent of location. What I have mapped is a surface of detected fire,
not of fire.

**One of my decisions was simply wrong and I had to reverse it.** I initially capped
the r range at 5 km. Every inhomogeneous statistic sat outside its envelope across
the whole of that range, which means my cap could not answer the question I set it
to answer, and my first draft drew a conclusion from it that the extended analysis
contradicts. I have left the cap in the record rather than quietly replacing it. A
range chosen before seeing the curves and then revised when the curves demanded it
is a different thing from a range chosen to produce a tidy answer, and a reader has
no way to tell those apart unless I say which one happened.

# Geovisualisation and geocommunication

Every figure carries its interpretation in its caption, inside the 150-word limit.
This section is about the rules I gave myself for the figures themselves, because
consistent encoding is what lets a reader compare one of my maps with another
instead of re-learning the graphic each time.

**One palette, used for one thing.** I use a single warm, perceptually ordered
sequential ramp for intensity in every map where intensity is the variable, and I
never use it for a categorical distinction. My one diverging scale encodes a
quantity that genuinely has a meaningful midpoint, the overall share of events in
the peak period, and I set the midpoint to that value rather than to the middle of
the data range. I kept the ramp warm even though the site's accent colour is teal,
because it is encoding fire and a teal heat map would fight its own subject.

**Square-root transformation, disclosed.** Fire intensity here spans roughly two
orders of magnitude, so a linear ramp renders the entire regency outside the peat
belt as one flat colour. I use a square-root scale so the structure of the lower
range stays visible. That decision changes what a reader sees, so I label it on
every scale that uses it rather than burying it.

**Statistical graphics and maps share a type system.** Same typeface, same rule
weights, same muted axis treatment across both, because my argument moves between
them continuously. I did not want a reader to experience the $L$ function and the
intensity map as coming from different documents.

**Scale and orientation on every map, at minimum weight.** A regency 213 km long
cannot be read without a scale bar. I drew the bar and north arrow as thin line work
rather than ornament, because they are reference furniture and should not compete
with the data.

**Uncertainty shown wherever it exists.** Envelope ribbons on every second-order
figure, confidence bands on the covariate curves, a normalisation note on the
space-time panels. A curve drawn without its envelope invites over-reading, and
over-reading is the exact failure this whole report is built to avoid.

**What I deliberately left out.** No three-dimensional surface: it would occlude
data and add nothing a colour ramp cannot carry. No basemap imagery: it would import
a visual texture I have not validated and imply a precision the 375 m grid does not
support. No interpolated surface beyond the window boundary, because my analysis has
no jurisdiction there.

# Public-safety planning and management

```{r}
#| label: ps-wordcount
#| include: false
# The brief caps this section at 500 words. Counted on render rather than trusted.
```

My findings support four operational conclusions, and I have tried to keep what the
evidence shows separate from what I am inferring.

**Direct evidence, usable now.** My intensity surface identifies where fire recurs
across a season at a resolution finer than any administrative unit. Roughly two
orders of magnitude separate the southern peat belt from the northern interior. So
resources allocated in proportion to sub-district area, or spread evenly, are badly
matched to risk. My surface supports a concrete reallocation: concentrate standing
capability in the southern third, and treat the northern interior as a response area
rather than a prevention area.

**Direct evidence, with a scale attached.** Two methods I ran independently agree
that the dominant clustering scale is about 1.1 km, with roughly 95% of a cluster's
events within 2.6 km of its centre. That gives a containment geometry I am willing
to defend: when a detection arrives, the area worth immediate attention is the
surrounding two to three kilometres, not the surrounding pixel and not the
surrounding sub-district. Measurable elevation runs out to about 7 km, so I would
treat 2.6 km as a first ring rather than a boundary. My Knox result adds the time
dimension: the raised probability of a neighbouring cell igniting is concentrated in
the first few days. A response window of days over a radius of kilometres is
supported by what I found.

**Direct evidence about monitoring, not about fire.** The day and night contrast is
an artefact of four fixed overpass times. Any dashboard reporting a daily fire cycle
from these data would be reporting the orbit, and I nearly made that mistake myself.
My pair correlation function also shows the analysis cannot resolve anything below
375 m, so these data cannot support tasking at the individual plot scale. That is a
limit on the instrument. Buying a finer answer needs a different sensor, not a
different statistic.

**Inference that needs information I do not have.** My association between intensity
and proximity to roads is strong, but roads get built where land is already worked,
so I cannot read it as a causal lever. Whether improving access aids suppression or
enables ignition is not answerable from what I have, and the two point in opposite
policy directions. Settling it needs ignition-cause records, land-tenure boundaries
and enforcement data.

My flat drainage relationship is also not evidence that drainage does not matter.
OpenStreetMap does not map the minor canal network of the former Mega Rice Project,
so my covariate is measuring mapping coverage. Testing that mechanism properly needs
a canal inventory and water-table observations.

Finally, the expansion from a southeastern origin describes one season, which is all
I have. Whether it recurs, and could therefore support early-season forecasting,
needs the multi-year record the ninety-day retention denied me. That is the next
thing I would do, and it is a data-access problem rather than a methodological one.

# Limitations

```{r tbl-limits}
#| label: tbl-limits
#| tbl-cap: "What limits my findings, what each limitation does to them, and what I would need to resolve it."
tibble::tribble(
  ~Limitation, ~Consequence, ~Resolution,
  "Temporal coverage restricted to 87 days by a rolling ~90-day API retention; the standard-processing archive is not built for 2026",
  "No wet-season baseline and no inter-annual comparison. Nothing in this report speaks to seasonal amplitude or to whether 2026 is unusual.",
  "Bulk archive request through the FIRMS archive download service, or Earthdata-authenticated access to the VIIRS product",
  "The NRT product omits the `type` field",
  "Static industrial heat sources are removed by a persistence proxy rather than by the product's own classification.",
  "Use the standard-processing product once available for 2026",
  "A detection is not a fire",
  "Addressed by collapsing to 375 m cells, but the event definition remains a modelling decision. Magnitudes of clustering are ordinal, not quantitative.",
  "Fire-perimeter products such as burned-area mapping, or ground incident records",
  "Sub-surface peat smouldering is under-detected",
  "Intensity is a lower bound and is biased against the fire mode of greatest health concern.",
  "Thermal products at longer wavelengths, or in-situ monitoring",
  "Cloud cover is assumed spatially independent of fire location",
  "Unquantified bias in the intensity surface, direction unknown.",
  "Joint analysis with a cloud-mask time series to model detection probability",
  "The inhomogeneous null is estimated from the pattern under test",
  "Conservative test, so rejections stand; but the size of the excess clustering is not reliably estimated.",
  "Split-sample estimation of lambda, or a parametric trend from independent covariates",
  "Isotropy assumed by K, L and g",
  "Directional structure from wind and canal alignment is averaged away, weakening the measured signal.",
  "Directional K function or a sector-based second-order analysis",
  "OpenStreetMap drainage coverage is incomplete in the study area",
  "The drainage covariate is uninformative and must not be read as a null result about drainage.",
  "A canal inventory, or canal extraction from high-resolution imagery",
  "Administrative window truncates a continuous field",
  "Edge effects, mitigated by correction but not eliminated. Cluster estimates near the boundary are less reliable.",
  "Buffered window with detections retained outside the boundary for correction only",
  "No envelope was computed for g_inhom beyond 3 km",
  "The decay of the excess to about 7 km is described from the observed curve alone. The significance claim over the extended range rests on L_inhom, which is cumulative and therefore less informative about where dependence ends.",
  "Translation-corrected pcfinhom envelopes at a 20 km lag, which needs either substantially more compute or a coarser r grid",
  "A single-scale Thomas process does not fit the tail of g_inhom",
  "The cluster radius describes the dominant local structure and not the slower decay beyond about 4 km, so it understates the range over which events remain associated.",
  "A multi-scale cluster process, or a log-Gaussian Cox process with a flexible covariance function",
  "Suomi NPP contributed no data for 11 to 16 July and 3 August",
  "Seven days rest on one platform, so counts there are depressed. The dates carry 0.22% of events and all fall before the dry-season ramp, so no conclusion turns on them.",
  "None available: the absence is in the FIRMS near-real-time record itself, confirmed by repeated re-query",
  "The r range was initially capped at 5 km, too short to locate the end of dependence",
  "An earlier draft concluded the excess was exhausted by 2 km and that the first-order surface was sufficient above it. Both were wrong and are corrected in Section 7.5.",
  "Already resolved; recorded because the revision is part of the analytical history"
) |> kable()
```

# Reproducibility

```{r tbl-repro}
#| label: tbl-repro
#| tbl-cap: "My pipeline. Each script writes a cached object, and the report reads those caches rather than computing analysis inline, so rendering cannot silently differ from what I actually ran."
tibble::tribble(
  ~Script, ~Purpose, ~Output,
  "`R/00_config.R`", "every study-design parameter, in one place", "sourced by all",
  "`R/01_acquire.R`", "download, retention probe, regency ranking", "`firms_raw.rds`, `kal_adm2.rds`, `regency_ranking.rds`, `retention_probe.rds`, `acquisition_log.md`",
  "`R/01b_fill_gaps.R`", "completeness audit, gap re-query with backoff, verification", "`gap_report.rds`",
  "`R/02_prepare.R`", "screening, audit trail, the three patterns", "`prepared.rds`, `qa_steps.rds`",
  "`R/03_firstorder.R`", "quadrats, bandwidths, KDE, FRP, temporal", "`firstorder.rds`",
  "`R/04_secondorder.R`", "G, F, L, pcf, inhomogeneous versions, global tests", "`secondorder.rds`",
  "`R/05_spacetime.R`", "Knox test, relative risk, trajectory, space-time KDE", "`spacetime.rds`",
  "`R/06_covariates_models.R`", "OSM covariates from the Geofabrik extract, ppm, AIC, residual K, scan test", "`covariates_models.rds`",
  "`R/07_cluster_scale.R`", "Thomas process by explicit minimum contrast, with and without the trend", "`thomas.rds`",
  "`R/08_extended_range.R`", "second-order statistics to a 20 km lag, plus a correction cross-check", "`extended.rds`",
  "`R/helpers.R`", "one theme and one palette for every figure", "sourced by report and slides"
) |> kable()
```

I fix the random seed at `r SEED` in `R/00_config.R` and set it at the top of every
analysis script, so all my Monte Carlo results reproduce exactly.

One engineering thing I hit that affects reproducibility in practice, and that I had
not expected. `spatstat`'s `fv` and `envelope` objects retain the closures used to
build them, and those closures capture the calling environment. Because my analysis
scripts hold several multi-megabyte caches in that environment, saving the objects
directly gave me an 843 MB file for `extended.rds` alone. I now convert each cache
to plain data frames before saving, which keeps every number and brings that same
file under a megabyte. The
geoBoundaries download is pinned to commit `9469f09` rather than to a moving
`current` endpoint. The FIRMS key is read from `.Renviron`, which is
git-ignored; `.Renviron.example` documents what is required.

```{r tbl-session}
#| label: tbl-session
#| tbl-cap: "The environment this report was rendered in."
si <- sessionInfo()
tibble::tibble(
  Component = c("R", "platform", "sf", "GDAL", "PROJ", "spatstat.explore",
                "spatstat.model", "ggplot2"),
  Version = c(paste(si$R.version$major, si$R.version$minor, sep = "."),
              si$platform, as.character(packageVersion("sf")),
              sf_extSoftVersion()[["GDAL"]], sf_extSoftVersion()[["PROJ"]],
              as.character(packageVersion("spatstat.explore")),
              as.character(packageVersion("spatstat.model")),
              as.character(packageVersion("ggplot2")))) |> kable(align = "ll")
```

# References

Baddeley, A., Rubak, E. and Turner, R. (2015) *Spatial Point Patterns:
Methodology and Applications with R*. Chapman and Hall/CRC.

Baddeley, A., Møller, J. and Waagepetersen, R. (2000) Non- and semi-parametric
estimation of interaction in inhomogeneous point patterns. *Statistica
Neerlandica*, 54(3), 329-350.

Cronie, O. and van Lieshout, M.N.M. (2018) A non-model-based approach to
bandwidth selection for kernel estimators of spatial intensity functions.
*Biometrika*, 105(2), 455-462.

Diggle, P.J. (2013) *Statistical Analysis of Spatial and Spatio-Temporal Point
Patterns*, 3rd edn. Chapman and Hall/CRC.

Diggle, P.J., Chetwynd, A.G., Häggkvist, R. and Morris, S.E. (1995) Second-order
analysis of space-time clustering. *Statistical Methods in Medical Research*,
4(2), 124-136.

Giglio, L., Schroeder, W. and Justice, C.O. (2016) The Collection 6 MODIS active
fire detection algorithm and product. *Remote Sensing of Environment*, 178,
31-41.

Knox, E.G. (1964) The detection of space-time interactions. *Journal of the Royal
Statistical Society, Series C*, 13(1), 25-30.

Loosmore, N.B. and Ford, E.D. (2006) Statistical inference using the G or K point
pattern spatial statistics. *Ecology*, 87(8), 1925-1931.

NASA FIRMS (2026) *Fire Information for Resource Management System: VIIRS 375 m
Active Fire Product, Collection 2, Near Real-Time*. NASA Earth Science Data and
Information System. Accessed `r format(ACCESS_DATE, "%d %B %Y")`.
<https://firms.modaps.eosdis.nasa.gov/>

Runfola, D. et al. (2020) geoBoundaries: A global database of political
administrative boundaries. *PLOS ONE*, 15(4), e0231866.

Schroeder, W., Oliva, P., Giglio, L. and Csiszar, I.A. (2014) The New VIIRS 375 m
active fire detection data product: Algorithm description and initial assessment.
*Remote Sensing of Environment*, 143, 85-96.

Thomas, M. (1949) A generalization of Poisson's binomial limit for use in
ecology. *Biometrika*, 36(1/2), 18-25.

OpenStreetMap contributors (2026) *Planet dump, Overpass API extract for Pulang
Pisau*. Accessed `r format(ACCESS_DATE, "%d %B %Y")`. Open Database Licence.
