---
title: "Where peat burns, and how close together"
subtitle: "VIIRS active-fire detections in Pulang Pisau Regency, Central Kalimantan, 1 July to 25 September 2026. ISSS626 Take-home Exercise 1."
format:
  revealjs:
    theme: [dark, slides.scss]
    slide-number: c/t
    progress: true
    transition: none
    background-transition: none
    width: 1280
    height: 760
    margin: 0.07
    fig-align: left
    fig-dpi: 190
    embed-resources: true
    code-fold: true
    echo: false
    toc: false
execute:
  echo: false
  warning: false
  message: false

knitr:
  opts_chunk:
    dev: ragg_png
    dev.args:
      background: transparent
    out.width: "100%"
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(sf); library(ggplot2); library(glue)
  library(spatstat.geom); library(spatstat.explore); library(patchwork); library(knitr)
})
source("R/00_config.R"); source("R/helpers.R")

# The figure theme reads PAL at call time, so overriding it here converts every
# plot in the deck to the dark palette without touching a single plot definition.
PAL <- modifyList(PAL, list(
  ink = "#F4E7DF", ground = "#2B1512", ground_alt = "#3A1D18",
  rule = "#593029", muted = "#C3A79B",
  accent = "#F0864B", accent_dk = "#F3A26A", accent_lt = "#C96A3C",
  cool = "#8FBDC0",
  seq = c("#2B1512", "#5E2A17", "#93441C", "#C2571F", "#E5893F", "#F3B579", "#FBE3C4")))

P  <- readRDS(file.path(DIR_DERIVED, "prepared.rds"))
FO <- readRDS(file.path(DIR_DERIVED, "firstorder.rds"))
SO <- readRDS(file.path(DIR_DERIVED, "secondorder.rds"))
ST <- readRDS(file.path(DIR_DERIVED, "spacetime.rds"))
CM <- readRDS(file.path(DIR_DERIVED, "covariates_models.rds"))
EX <- readRDS(file.path(DIR_DERIVED, "extended.rds"))
TH <- readRDS(file.path(DIR_DERIVED, "thomas.rds"))
RANK <- readRDS(file.path(DIR_DERIVED, "regency_ranking.rds"))

win_p <- P$win_p; bb <- st_bbox(win_p)
infl <- P$ppp_all$n / P$ppp_cell$n
tp   <- TH$thomas_tab |> filter(key == "trend_removed")
sig_t <- round(tp$sigma_m); r95 <- round(tp$r95_m); mu_t <- round(tp$mu_per_cluster)
r_decay <- EX$summary_ext$r_g_within_10pct_of_1
peak_r  <- EX$summary_ext$peak_r_m
dnull <- CM$aic_tab$dAIC[CM$aic_tab$model == "null"]
droad <- dnull - CM$aic_tab$dAIC[CM$aic_tab$model == "road"]
knox_best <- ST$knox_tab |> arrange(desc(ratio)) |> slice(1)
```

## Contents {.smaller}

::: {.subhead}
One argument running through all of it: a satellite hotspot is not a fire, and
almost everything I did follows from taking that seriously.
:::

1. Why I chose this place, and what the data actually is
2. The problem the data handed me: a detection is not a fire
3. First order: where fire concentrates
4. First order: when, and what the satellite cannot tell me
5. Second order: why complete spatial randomness is a straw man
6. Second order: the test that can fail
7. Space and time together
8. What the clustering is organised around, and how big it is
9. What I think this supports for planning
10. What it does not support

## 1. A regency chosen by rule, not reputation

::: {.split}
::: {.col}
```{r sl-context}
#| fig-height: 5.4
#| fig-width: 5.4
kal <- readRDS(file.path(DIR_DERIVED, "kal_adm2.rds")) |> st_transform(STUDY_CRS) |>
  left_join(RANK |> select(shapeName, per_1000km2), by = "shapeName") |>
  mutate(per_1000km2 = replace_na(per_1000km2, 0))
ggplot(kal) +
  geom_sf(aes(fill = per_1000km2), colour = PAL$ground, linewidth = 0.15) +
  geom_sf(data = win_p, fill = NA, colour = PAL$ink, linewidth = 0.8) +
  scale_fill_intensity(name = "Detections per 1,000 km\u00b2", v = kal$per_1000km2) +
  coord_sf(datum = NA, expand = FALSE) + theme_map(base_size = 13)
```
:::
::: {.note}
**My rule, set first.** Highest detection density among regencies above
5,000 km², the size floor I need for second-order estimation to a 5 km lag.

**Pulang Pisau wins outright**, at
`r fmt_n(round(RANK$per_1000km2[RANK$shapeName == STUDY_REGENCY]))` per 1,000 km².

**What I rejected.** Ketapang leads on raw count
(`r fmt_n(RANK$n_detections[RANK$shapeName == "Ketapang"])`) but is 3x the area.
Kota Palangka Raya is denser but only
`r fmt_n(round(RANK$area_km2[RANK$shapeName == "Kota Palangka Raya"]))` km².

*My first pass picked it from a 7-day sample where it led on count. Over the full
window it does not, so I restated the rule against the whole record.*
:::
:::

## 2. Data, and a constraint worth naming

```{r}
#| output: asis
kf <- function(...) {
  items <- list(...)
  cat('<div class="kf">')
  for (it in items) cat(sprintf('<div><div class="n">%s</div><div class="l">%s</div></div>',
                                it[[1]], it[[2]]))
  cat('</div>')
}
kf(list(fmt_n(P$qa$n_in[1]), "rows downloaded, all Kalimantan"),
   list(fmt_n(P$ppp_all$n),  "detections inside the regency"),
   list(fmt_n(P$ppp_cell$n), "burned 375 m cells, the analysis unit"),
   list(P$n_days,            "days, 1 Jul to 25 Sep"))
```

VIIRS 375 m Collection 2, both platforms. My window is the administrative polygon
in UTM 49S, `r fmt_n(round(P$area_km2))` km².

**The temporal window was imposed on me, not chosen.** The FIRMS `area` API retains
near-real-time granules for roughly 90 days, and the standard-processing archive is
not built for 2026. My Kalimantan-wide queries return exactly zero rows for every
sampled date from January to 30 June, and non-zero from 1 July. Six months of zero
detections across a 10° box is not credible, so I read that as absence of coverage,
not absence of fire.

*The cost: no wet-season baseline, so nothing here speaks to seasonal amplitude. It
does span onset to peak, which is the period an agency acts on.*

## 3. A detection is not a fire

::: {.split}
::: {.col}
```{r sl-mult}
#| fig-height: 4.6
#| fig-width: 5.6
mt <- P$multiplicity
ggplot(mt, aes(n_det, n_cells)) +
  geom_segment(aes(xend = n_det, yend = 0), colour = PAL$accent, linewidth = 0.4) +
  scale_x_log10() + scale_y_log10(labels = label_number(big.mark = ",")) +
  labs(x = "Detections in the cell", y = "Cells") +
  theme_ex01(base_size = 13)
```
:::
::: {.note}
One large peat fire fills many 375 m pixels, and two satellites re-detect it about
four times a day.

**I measured it: the raw count overstates distinct burning locations by
`r round(infl, 1)`x.**

So the pattern I started with partly measures the satellite's revisit schedule,
which means testing against complete spatial randomness was guaranteed to reject
before I computed anything.

**So I changed two things.** The unit becomes the 375 m cell. The null becomes an
inhomogeneous Poisson process.

That is the spine of everything else I did.
:::
:::

## 4. Where it burns

::: {.split}
::: {.col}
```{r sl-kde}
#| fig-height: 5.9
#| fig-width: 4.6
kd <- im_df(FO$kde$main) |> mutate(v = per_1000km2(intensity))
ggplot(kd) +
  geom_raster(aes(x, y, fill = v)) +
  geom_sf(data = win_p, fill = NA, colour = PAL$ink, linewidth = 0.4) +
  scale_fill_intensity(v = kd$v) +
  coord_sf(datum = NA, expand = FALSE) + scalebar_layer(bb) + north_layer(bb) +
  theme_map(base_size = 11) +
  theme(legend.position = "right", legend.key.height = unit(52, "pt"),
        legend.key.width = unit(8, "pt"))
```
:::
::: {.note}
Gaussian kernel, Diggle edge correction, **bandwidth I fixed at 4 km**.

Cross-validated selectors gave me about 0.6 km and I **rejected** them: they absorb
the very clustering my second-order test exists to detect, which would have rigged
the test on Slide 7.

Intensity spans roughly **two orders of magnitude** between the southern peat belt
and the northern interior.

*So allocating in proportion to sub-district area is badly matched to risk. I think
that is the most directly usable thing I found.*
:::
:::

## 5. When it burns, and what the orbit decides

::: {.split}
::: {.col}
```{r sl-daily}
#| fig-height: 4.6
#| fig-width: 5.6
ggplot(FO$daily, aes(date_local)) +
  geom_col(aes(y = n_events), fill = PAL$accent_lt, width = 0.9) +
  geom_line(aes(y = roll7), colour = PAL$accent_dk, linewidth = 0.8) +
  geom_vline(xintercept = as.numeric(ST$split_date), colour = PAL$cool,
             linetype = "22", linewidth = 0.4) +
  scale_x_date(date_breaks = "3 weeks", date_labels = "%d %b") +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  labs(x = NULL, y = "Burned cells per day") +
  theme_ex01(base_size = 13)
```
:::
::: {.note}
Heavily back-loaded: week 1 recorded 7 events, the week of 16 September recorded
8,087.

**The day/night contrast is not a daily cycle, and I nearly read it as one.** Both
platforms observe at four narrow fixed local times, so if I claimed fire peaks at a
given hour I would be describing the orbit.

I use the split only as a detectability covariate, since night overpasses catch
cooler anomalies against a cooler background.
:::
:::

## 6. Against CSR, everything is significant

```{r sl-csr}
#| fig-height: 3.5
#| fig-width: 12.0
eg <- SO$env_G |> mutate(panel = "G(r)")
ef <- SO$env_F |> mutate(panel = "F(r)")
el <- SO$env_L |> mutate(across(c(obs, theo, lo, hi), ~ .x - r), panel = "L(r) - r")
ea <- bind_rows(eg, ef, el)
ea$panel <- factor(ea$panel, levels = c("G(r)", "F(r)", "L(r) - r"))
ggplot(ea, aes(r)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = PAL$rule) +
  geom_line(aes(y = theo), colour = PAL$muted, linetype = "22", linewidth = 0.35) +
  geom_line(aes(y = obs), colour = PAL$accent, linewidth = 0.8) +
  facet_wrap(~panel, scales = "free", nrow = 1) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = " km")) +
  labs(x = "r", y = NULL) + theme_ex01(base_size = 12)
```

Every statistic leaves the envelope immediately. **This should persuade nobody,
including me**: the sensor guarantees it. I show it so the failure is visible rather
than hidden.

My pair correlation function makes the point sharper. It is suppressed below 375 m,
the pixel dimension. That apparent short-range inhibition is the instrument, not
fires avoiding one another.

## 7. The test that can fail

::: {.split}
::: {.col}
```{r sl-inhom}
#| fig-height: 4.6
#| fig-width: 6.4
Lx <- EX$env_L_ext |> mutate(across(c(obs, lo, hi), ~ .x - r), ref = 0,
                             panel = "L_inhom(r) - r, envelope")
Gx <- EX$g_obs_ext |> filter(r > 0, is.finite(trans)) |>
  transmute(r, obs = trans, lo = NA_real_, hi = NA_real_, ref = 1,
            panel = "g_inhom(r), observed")
ggplot(bind_rows(Lx, Gx), aes(r)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = PAL$rule, na.rm = TRUE) +
  geom_line(aes(y = ref), colour = PAL$muted, linetype = "22", linewidth = 0.35) +
  geom_line(aes(y = obs), colour = PAL$accent, linewidth = 0.8) +
  facet_wrap(~panel, scales = "free", nrow = 1) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = " km")) +
  labs(x = "r", y = NULL) + theme_ex01(base_size = 11)
```
:::
::: {.note}
My null is an **inhomogeneous Poisson process** with intensity from the 4 km kernel
estimate, so everything my first-order analysis found is already inside the null. I
**re-estimate the reference intensity for every simulated pattern**.

**The departure survives.** Both global tests reject at the Monte Carlo floor.

**The two panels disagree, and that was the lesson for me.** L_inhom never re-enters
its envelope in 20 km. g_inhom, which localises scale instead of accumulating it,
peaks at `r round(peak_r)` m and is within 10% of no-excess by
`r round(r_decay/1000, 1)` km.

*L accumulates every excess at smaller lags, so it stays significant far past the
range where anything is happening. My first draft capped r at 5 km and drew the
wrong conclusion from exactly this.*
:::
:::

## 8. Space and time are not separable

::: {.split}
::: {.col}
```{r sl-knox}
#| fig-height: 4.4
#| fig-width: 5.6
ggplot(ST$knox_tab, aes(factor(s_m), factor(t_days), fill = ratio)) +
  geom_tile(colour = PAL$ground, linewidth = 1.4) +
  geom_text(aes(label = sprintf("%.2f", ratio)), size = 4.1, colour = PAL$ground,
            fontface = "bold") +
  scale_fill_gradientn(colours = PAL$seq, name = "Obs / exp") +
  labs(x = "Spatial threshold (m)", y = "Temporal threshold (days)") +
  theme_ex01(base_size = 13, grid = "none")
```
:::
::: {.note}
Knox test. My null is **random labelling**: I hold locations and dates both fixed
and permute only their pairing, across `r fmt_n(ST$nperm)` permutations of
`r fmt_n(ST$n_close_pairs)` close pairs.

Because both marginals stay fixed, none of this can be explained away by the
seasonal ramp or the spatial concentration.

**Strongest at the tightest cell**, decaying monotonically outward. I read that as
short-range propagation: a cell that burns today raises the chance a neighbour burns
tomorrow, and the effect is largely spent within days and a kilometre or two.

*The decay is steeper in time than in space, which fits fire spreading over
contiguous ground within days rather than recurring in the same place weeks apart.*
:::
:::

## 9. What it is organised around, and how big it is

::: {.split}
::: {.col}
```{r sl-rho}
#| fig-height: 4.4
#| fig-width: 5.6
rd <- CM$rho_road
ggplot(rd, aes(.data[[names(rd)[1]]] / 1000)) +
  geom_ribbon(aes(ymin = per_1000km2(lo), ymax = per_1000km2(hi)),
              fill = PAL$rule, alpha = 0.7) +
  geom_line(aes(y = per_1000km2(rho)), colour = PAL$accent, linewidth = 0.8) +
  labs(x = "Distance to nearest road (km)", y = "Burned cells per 1,000 km²") +
  theme_ex01(base_size = 13)
```
:::
::: {.note}
**Access dominates.** Of `r fmt_n(round(dnull))` AIC units separating my covariate
model from an intercept-only model, distance to road accounts for
`r fmt_n(round(droad))`.

**Mapped drainage adds almost nothing** (13 AIC units). That is not evidence
drainage does not matter: OSM does not map the Mega Rice Project canal network, so
my covariate is measuring mapping coverage.

**Cluster scale `r fmt_n(sig_t)` m**, 95% within **`r fmt_n(r95)` m**, from a Thomas
process I fitted to g_inhom by minimum contrast.

*My first fit left the trend in and returned
`r fmt_n(round(TH$kppm_hom$sigma_m))` m and 440 events per cluster, which is really
the peat belt re-described as a few giant clusters. Trend and interaction are not
separately identifiable unless I force them apart.*

Residual K shows my covariates explain *location*, not *contagion*.
:::
:::

## 10. What this supports, and what it does not

::: {.split .even}
::: {.col}
[What I would act on]{.colhead}

**Reallocate by intensity, not by area.** Two orders of magnitude separate the
southern peat belt from the northern interior.

**A containment geometry.** First ring `r round(r95/1000, 1)` km, elevation out to
about `r round(r_decay/1000, 1)` km, days not weeks. Two methods I ran independently
agree on the dominant scale.

**A monitoring limit.** I cannot resolve anything below 375 m. A finer answer needs
a different sensor, not a different statistic.
:::
::: {.note}
[What I would not claim]{.colhead}

**Roads as a policy lever.** Roads follow worked land. I cannot tell from this
whether access aids suppression or enables ignition, and the two point opposite
ways.

**Any claim about drainage.** My covariate measures OSM coverage.

**Any claim about time of day.** Four fixed overpasses.

**Anything about how unusual 2026 is.** I only analysed 87 days. The REST API
retains ~90, but FIRMS serves VIIRS back to 2012 through its archive download
service, which authenticates by email and so does not fit a scripted pipeline. That
limit is my tooling, not the data.
:::
:::

::: {.footer-note}
My technical report, code and audit trail are in the repository. Data: NASA FIRMS, geoBoundaries, OpenStreetMap via Geofabrik.
:::
