---
title: "Where peat burns, and how close together"
subtitle: "VIIRS active-fire detections in Pulang Pisau Regency, Central Kalimantan, 1 July to 25 September 2026. ISSS626 Take-home Exercise 1."
format:
  revealjs:
    theme: [dark, slides.scss]
    slide-number: c/t
    progress: true
    transition: none
    background-transition: none
    width: 1280
    height: 760
    margin: 0.07
    fig-align: left
    fig-dpi: 190
    embed-resources: true
    code-fold: true
    echo: false
    toc: false
execute:
  echo: false
  warning: false
  message: false

knitr:
  opts_chunk:
    dev: ragg_png
    dev.args:
      background: transparent
    out.width: "100%"
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(sf); library(ggplot2); library(glue)
  library(spatstat.geom); library(spatstat.explore); library(patchwork); library(knitr)
})
source("R/00_config.R"); source("R/helpers.R")

# The figure theme reads PAL at call time, so overriding it here converts every
# plot in the deck to the dark palette without touching a single plot definition.
PAL <- modifyList(PAL, list(
  ink = "#F4E7DF", ground = "#2B1512", ground_alt = "#3A1D18",
  rule = "#593029", muted = "#C3A79B",
  accent = "#F0864B", accent_dk = "#F3A26A", accent_lt = "#C96A3C",
  cool = "#8FBDC0",
  seq = c("#2B1512", "#5E2A17", "#93441C", "#C2571F", "#E5893F", "#F3B579", "#FBE3C4")))

P  <- readRDS(file.path(DIR_DERIVED, "prepared.rds"))
FO <- readRDS(file.path(DIR_DERIVED, "firstorder.rds"))
SO <- readRDS(file.path(DIR_DERIVED, "secondorder.rds"))
ST <- readRDS(file.path(DIR_DERIVED, "spacetime.rds"))
CM <- readRDS(file.path(DIR_DERIVED, "covariates_models.rds"))
EX <- readRDS(file.path(DIR_DERIVED, "extended.rds"))
TH <- readRDS(file.path(DIR_DERIVED, "thomas.rds"))
RANK <- readRDS(file.path(DIR_DERIVED, "regency_ranking.rds"))

win_p <- P$win_p; bb <- st_bbox(win_p)
infl <- P$ppp_all$n / P$ppp_cell$n
tp   <- TH$thomas_tab |> filter(key == "trend_removed")
sig_t <- round(tp$sigma_m); r95 <- round(tp$r95_m); mu_t <- round(tp$mu_per_cluster)
r_decay <- EX$summary_ext$r_g_within_10pct_of_1
peak_r  <- EX$summary_ext$peak_r_m
dnull <- CM$aic_tab$dAIC[CM$aic_tab$model == "null"]
droad <- dnull - CM$aic_tab$dAIC[CM$aic_tab$model == "road"]
knox_best <- ST$knox_tab |> arrange(desc(ratio)) |> slice(1)
```

## Contents {.smaller}

::: {.subhead}
Six tasks, one argument running through them.
:::

1. Why this place, and what the data actually is
2. The problem the data creates: a detection is not a fire
3. First order: where fire concentrates
4. First order: when, and what the satellite cannot tell us
5. Second order: why complete spatial randomness is a straw man
6. Second order: the test that can fail
7. Space and time together
8. What the clustering is organised around, and how big it is
9. What this supports for planning
10. What it does not support

## 1. A regency chosen by rule, not reputation

::: {.split}
::: {.col}
```{r sl-context}
#| fig-height: 5.4
#| fig-width: 5.4
kal <- readRDS(file.path(DIR_DERIVED, "kal_adm2.rds")) |> st_transform(STUDY_CRS) |>
  left_join(RANK |> select(shapeName, per_1000km2), by = "shapeName") |>
  mutate(per_1000km2 = replace_na(per_1000km2, 0))
ggplot(kal) +
  geom_sf(aes(fill = per_1000km2), colour = PAL$ground, linewidth = 0.15) +
  geom_sf(data = win_p, fill = NA, colour = PAL$ink, linewidth = 0.8) +
  scale_fill_intensity(name = "Detections per 1,000 km\u00b2", v = kal$per_1000km2) +
  coord_sf(datum = NA, expand = FALSE) + theme_map(base_size = 13)
```
:::
::: {.note}
**Rule, stated first.** Highest detection density among regencies above
5,000 km², the size floor needed for second-order estimation to a 5 km lag.

**Pulang Pisau wins outright**, at
`r fmt_n(round(RANK$per_1000km2[RANK$shapeName == STUDY_REGENCY]))` per 1,000 km².

**Rejected.** Ketapang leads on raw count
(`r fmt_n(RANK$n_detections[RANK$shapeName == "Ketapang"])`) but is 3x the area.
Kota Palangka Raya is denser but only
`r fmt_n(round(RANK$area_km2[RANK$shapeName == "Kota Palangka Raya"]))` km².

It is the drained peat of the former Mega Rice Project, so the public-safety
mechanism is respiratory, not thermal.
:::
:::

## 2. Data, and a constraint worth naming

```{r}
#| output: asis
kf <- function(...) {
  items <- list(...)
  cat('<div class="kf">')
  for (it in items) cat(sprintf('<div><div class="n">%s</div><div class="l">%s</div></div>',
                                it[[1]], it[[2]]))
  cat('</div>')
}
kf(list(fmt_n(P$qa$n_in[1]), "rows downloaded, all Kalimantan"),
   list(fmt_n(P$ppp_all$n),  "detections inside the regency"),
   list(fmt_n(P$ppp_cell$n), "burned 375 m cells, the analysis unit"),
   list(P$n_days,            "days, 1 Jul to 25 Sep"))
```

VIIRS 375 m Collection 2, both platforms. Window is the administrative polygon in
UTM 49S, `r fmt_n(round(P$area_km2))` km².

**The window is imposed, not chosen.** The FIRMS `area` API retains near-real-time
granules for roughly 90 days, and the standard-processing archive is not built for
2026. Kalimantan-wide queries return exactly zero rows for every sampled date from
January to 30 June, and non-zero from 1 July. Six months of zero detections across
a 10° box is not credible, so this is absence of coverage, not absence of fire.

*The cost: no wet-season baseline, so nothing here speaks to seasonal amplitude.
The window does span onset to peak, which is the period an agency acts on.*

## 3. A detection is not a fire

::: {.split}
::: {.col}
```{r sl-mult}
#| fig-height: 4.6
#| fig-width: 5.6
mt <- P$multiplicity
ggplot(mt, aes(n_det, n_cells)) +
  geom_segment(aes(xend = n_det, yend = 0), colour = PAL$accent, linewidth = 0.4) +
  scale_x_log10() + scale_y_log10(labels = label_number(big.mark = ",")) +
  labs(x = "Detections in the cell", y = "Cells") +
  theme_ex01(base_size = 13)
```
:::
::: {.note}
One large peat fire fills many 375 m pixels, and two satellites re-detect it about
four times a day.

**The raw count overstates distinct burning locations by `r round(infl, 1)`x.**

So the pattern partly measures the satellite's revisit schedule. Consequence:
testing against complete spatial randomness is guaranteed to reject before any
analysis is done.

**Two responses.** Change the unit to the 375 m cell. Change the null from CSR to
an inhomogeneous Poisson process.

This is the methodological spine of the whole exercise.
:::
:::

## 4. Where it burns

::: {.split}
::: {.col}
```{r sl-kde}
#| fig-height: 5.9
#| fig-width: 4.6
kd <- im_df(FO$kde$main) |> mutate(v = per_1000km2(intensity))
ggplot(kd) +
  geom_raster(aes(x, y, fill = v)) +
  geom_sf(data = win_p, fill = NA, colour = PAL$ink, linewidth = 0.4) +
  scale_fill_intensity(v = kd$v) +
  coord_sf(datum = NA, expand = FALSE) + scalebar_layer(bb) + north_layer(bb) +
  theme_map(base_size = 11) +
  theme(legend.position = "right", legend.key.height = unit(52, "pt"),
        legend.key.width = unit(8, "pt"))
```
:::
::: {.note}
Gaussian kernel, Diggle edge correction, **bandwidth fixed at 4 km**.

Cross-validated selectors returned about 0.6 km and were **rejected**: they absorb
the very clustering the second-order test exists to detect, which would rig the
test in Slide 7.

Intensity spans roughly **two orders of magnitude** between the southern peat belt
and the northern interior.

*So allocation proportional to sub-district area is badly matched to risk. That is
the single most directly usable result in the deck.*
:::
:::

## 5. When it burns, and what the orbit decides

::: {.split}
::: {.col}
```{r sl-daily}
#| fig-height: 4.6
#| fig-width: 5.6
ggplot(FO$daily, aes(date_local)) +
  geom_col(aes(y = n_events), fill = PAL$accent_lt, width = 0.9) +
  geom_line(aes(y = roll7), colour = PAL$accent_dk, linewidth = 0.8) +
  geom_vline(xintercept = as.numeric(ST$split_date), colour = PAL$cool,
             linetype = "22", linewidth = 0.4) +
  scale_x_date(date_breaks = "3 weeks", date_labels = "%d %b") +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  labs(x = NULL, y = "Burned cells per day") +
  theme_ex01(base_size = 13)
```
:::
::: {.note}
Heavily back-loaded: week 1 recorded 7 events, the week of 16 September recorded
8,087.

**The day/night contrast is not a diurnal cycle.** Both platforms observe at four
narrow fixed local times, so any claim that fire peaks at a given hour would be a
claim about the orbit.

Correct use of the split is as a detectability covariate, since night overpasses
detect cooler anomalies against a cooler background.
:::
:::

## 6. Against CSR, everything is significant

```{r sl-csr}
#| fig-height: 3.5
#| fig-width: 12.0
eg <- SO$env_G |> mutate(panel = "G(r)")
ef <- SO$env_F |> mutate(panel = "F(r)")
el <- SO$env_L |> mutate(across(c(obs, theo, lo, hi), ~ .x - r), panel = "L(r) - r")
ea <- bind_rows(eg, ef, el)
ea$panel <- factor(ea$panel, levels = c("G(r)", "F(r)", "L(r) - r"))
ggplot(ea, aes(r)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = PAL$rule) +
  geom_line(aes(y = theo), colour = PAL$muted, linetype = "22", linewidth = 0.35) +
  geom_line(aes(y = obs), colour = PAL$accent, linewidth = 0.8) +
  facet_wrap(~panel, scales = "free", nrow = 1) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = " km")) +
  labs(x = "r", y = NULL) + theme_ex01(base_size = 12)
```

Every statistic leaves the envelope immediately. **This persuades nobody**: the
sensor guarantees it. Reported so the failure is visible rather than concealed.

The pair correlation function makes the point sharper still. It is suppressed below
375 m, the pixel dimension. That apparent short-range inhibition is the instrument,
not fires avoiding one another.

## 7. The test that can fail

::: {.split}
::: {.col}
```{r sl-inhom}
#| fig-height: 4.6
#| fig-width: 6.4
Lx <- EX$env_L_ext |> mutate(across(c(obs, lo, hi), ~ .x - r), ref = 0,
                             panel = "L_inhom(r) - r, envelope")
Gx <- EX$g_obs_ext |> filter(r > 0, is.finite(trans)) |>
  transmute(r, obs = trans, lo = NA_real_, hi = NA_real_, ref = 1,
            panel = "g_inhom(r), observed")
ggplot(bind_rows(Lx, Gx), aes(r)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = PAL$rule, na.rm = TRUE) +
  geom_line(aes(y = ref), colour = PAL$muted, linetype = "22", linewidth = 0.35) +
  geom_line(aes(y = obs), colour = PAL$accent, linewidth = 0.8) +
  facet_wrap(~panel, scales = "free", nrow = 1) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = " km")) +
  labs(x = "r", y = NULL) + theme_ex01(base_size = 11)
```
:::
::: {.note}
Null is an **inhomogeneous Poisson process** with intensity from the 4 km kernel
estimate, so everything the first-order analysis found is already inside the null.
The reference intensity is **re-estimated for every simulated pattern**.

**The departure survives.** Both global tests reject at the Monte Carlo floor.

**The two panels disagree, and that is the lesson.** L_inhom never re-enters its
envelope in 20 km. g_inhom, which localises scale instead of accumulating it, peaks
at `r round(peak_r)` m and is within 10% of no-excess by
`r round(r_decay/1000, 1)` km.

*L accumulates every excess at smaller lags, so it stays significant far past the
range over which anything is happening.*
:::
:::

## 8. Space and time are not separable

::: {.split}
::: {.col}
```{r sl-knox}
#| fig-height: 4.4
#| fig-width: 5.6
ggplot(ST$knox_tab, aes(factor(s_m), factor(t_days), fill = ratio)) +
  geom_tile(colour = PAL$ground, linewidth = 1.4) +
  geom_text(aes(label = sprintf("%.2f", ratio)), size = 4.1, colour = PAL$ground,
            fontface = "bold") +
  scale_fill_gradientn(colours = PAL$seq, name = "Obs / exp") +
  labs(x = "Spatial threshold (m)", y = "Temporal threshold (days)") +
  theme_ex01(base_size = 13, grid = "none")
```
:::
::: {.note}
Knox test. Null is **random labelling**: locations and dates both held fixed, only
their pairing permuted, across `r fmt_n(ST$nperm)` permutations of
`r fmt_n(ST$n_close_pairs)` close pairs.

Because both marginals are fixed, none of this can be explained by the seasonal
ramp or the spatial concentration.

**Strongest at the tightest cell** and decaying monotonically outward: a cell that
burns today raises the chance a neighbour burns tomorrow, and the effect is largely
spent within days and a kilometre or two.
:::
:::

## 9. What it is organised around, and how big it is

::: {.split}
::: {.col}
```{r sl-rho}
#| fig-height: 4.4
#| fig-width: 5.6
rd <- CM$rho_road
ggplot(rd, aes(.data[[names(rd)[1]]] / 1000)) +
  geom_ribbon(aes(ymin = per_1000km2(lo), ymax = per_1000km2(hi)),
              fill = PAL$rule, alpha = 0.7) +
  geom_line(aes(y = per_1000km2(rho)), colour = PAL$accent, linewidth = 0.8) +
  labs(x = "Distance to nearest road (km)", y = "Burned cells per 1,000 km²") +
  theme_ex01(base_size = 13)
```
:::
::: {.note}
**Access dominates.** Of `r fmt_n(round(dnull))` AIC units separating the covariate
model from an intercept-only model, distance to road accounts for
`r fmt_n(round(droad))`.

**Mapped drainage adds almost nothing** (13 AIC units). Not evidence that drainage
does not matter: OSM does not map the Mega Rice Project canal network, so the
covariate measures mapping coverage.

**Cluster scale `r fmt_n(sig_t)` m**, 95% within **`r fmt_n(r95)` m**, from a
Thomas process fitted to g_inhom by minimum contrast.

*Fitted without the trend removed it returns
`r fmt_n(round(TH$kppm_hom$sigma_m))` m and 440 events per cluster, which is the
peat belt being re-described as a few giant clusters. Trend and interaction are not
separately identifiable unless you make them so.*

Residual K shows the covariates explain *location*, not *contagion*.
:::
:::

## 10. What this supports, and what it does not

::: {.split .even}
::: {.col}
[Supported now]{.colhead}

**Reallocate by intensity, not by area.** Two orders of magnitude separate the
southern peat belt from the northern interior.

**A containment geometry.** First ring `r round(r95/1000, 1)` km, elevation out to
about `r round(r_decay/1000, 1)` km, days not weeks. Two independent methods agree
on the dominant scale.

**A monitoring limit.** Nothing below 375 m is resolvable. Buying a finer answer
needs a different sensor, not a different statistic.
:::
::: {.note}
[Not supported]{.colhead}

**Roads as a policy lever.** Roads follow worked land. Whether access aids
suppression or enables ignition is not answerable here, and the two have opposite
implications.

**Any claim about drainage.** The covariate measures OSM coverage.

**Any diurnal claim.** Four fixed overpasses.

**Anything about how unusual 2026 is.** The 90-day retention denied the multi-year
record. Obtaining it is the highest-value next step, and it is a data-access task,
not a methodological one.
:::
:::

::: {.footer-note}
Technical report, code and audit trail: see the repository. Data: NASA FIRMS, geoBoundaries, OpenStreetMap via Geofabrik.
:::
